---
title: "Case study"
subtitle: "Mouse gut metaproteome: 20% yeast vs 20% casein protein"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

This page runs the `repdist` workflow on a published metaproteome: beta
diversity with MMD (`sample_mmd()`, the maximum mean discrepancy under a
Gaussian RBF kernel on the protein embeddings), the representation space,
MCL bins, and bin-level differential abundance.

```{r libs}
library(repdist)
library(ggplot2)
library(vegan)
```

# Study and comparison

Blakeley-Ruiz *et al.* [@blakeleyruiz2025diet] fed twelve mice (two cages of
three per sex) a series of defined diets, one week each, and measured faecal
metaproteomes by spectral counting (PRIDE PXD041586).

**The input sequences are already sequence-reduced.** The microbial protein
catalog comes from metagenome-assembled genomes annotated with PROKKA;
unbinned genes were added only if no binned gene matched them at 90% identity
(CD-HIT-2D), and the host and diet reference proteomes were clustered at 95%
identity. Every reported protein also has at least one protein-unique
peptide. Near-identical proteins are therefore rare in the table (see
[sequence identity of graph edges](#edge-identity)): the relatedness that MMD
and the bins add beyond protein identity is mostly remote homology. This is
consistent with the modest difference between MMD and protein-level
Bray-Curtis below.

This page compares two diets of the same protein amount: **20% yeast protein
(Yeast) against 20% casein (Casein)**, casein being the reference protein of
the series. Only mice sampled on both diets are used (11 pairs), and diet is
tested within mouse. The yeast diet came three weeks after the casein diet,
so diet is confounded with feeding period in this serial design.

```{r data}
d <- readRDS(file.path("curated", "diet_yeast.rds"))
x <- d$proteins # sequences with TM-Vec embeddings
counts <- d$counts # raw PSM counts, samples x proteins
samples <- d$samples
ann <- d$annotation
d$provenance$contrast
table(samples$diet, samples$sex)
```

The curated object was built by `data-raw/diet_yeast.R` from the paper's
supplementary Data Set 1 (PSM counts and sequences) and Data Set 6
(annotation): every microbial protein with at least one PSM in these 22
samples, its species group and GTDB lineage, the study's curated function,
KEGG orthology (KO) assignments and CAZy families, and a TM-Vec embedding
computed with `embed_proteins()`.

**Protein length.** Proteins longer than `MAX_AA` = `r d$provenance$max_aa`
residues were removed before embedding:
`r d$provenance$steps[["longer_than_max"]]` of
`r d$provenance$steps[["detected"]]` proteins, carrying
`r sprintf("%.1f%%", 100 * d$provenance$long_protein_psm_share)` of the PSMs.
This is an analysis choice, not a package limit: `embed_proteins()` embeds
sequences of any length, but the TM-Vec head was trained on chains of up to
1000 residues, so predicted TM-scores for much longer chains are
extrapolations.

```{r dims}
c(samples = nrow(counts), proteins = ncol(counts))
```

# Beta diversity

Sequencing depth differs by diet (yeast samples have about twice the PSMs of
casein samples), and depth changes which proteins are observed. Both
distances are therefore computed on counts rarefied to the smallest sample.

```{r rarefy}
tapply(rowSums(counts), samples$diet, range)
set.seed(1)
rarefied <- rrarefy(counts, min(rowSums(counts)))
D <- list(
    `Bray-Curtis (protein)` = vegdist(rarefied, "bray"),
    MMD = sample_mmd(rarefied, x)
)
attr(D$MMD, "sigma")
```

PERMANOVA [@anderson2001permanova] and PERMDISP [@anderson2006permdisp] use
999 permutations restricted to within mouse.

```{r permanova}
perm <- permute::how(blocks = samples$subject, nperm = 999)
tests <- t(vapply(D, function(Dk) {
    set.seed(1)
    a <- adonis2(Dk ~ diet, data = samples, permutations = perm)
    bd <- betadisper(Dk, samples$diet)
    set.seed(1)
    pd <- permutest(bd, permutations = perm)
    c(R2 = a$R2[1], F = a$F[1], p = a$`Pr(>F)`[1],
        dispersion_Casein = mean(bd$distances[samples$diet == "Casein"]),
        dispersion_Yeast = mean(bd$distances[samples$diet == "Yeast"]),
        PERMDISP_p = pd$tab[1, "Pr(>F)"])
}, numeric(6)))
knitr::kable(tests, digits = 3)
```

```{r pcoa, fig.width = 8, fig.height = 3.8, fig.alt = "PCoA under protein-level Bray-Curtis and MMD; lines join the two samples of each mouse"}
ord <- do.call(rbind, lapply(names(D), function(n) {
    e <- cmdscale(D[[n]], k = 2, eig = TRUE)
    rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
    data.frame(PCo1 = e$points[, 1], PCo2 = e$points[, 2],
        diet = samples$diet, subject = samples$subject,
        panel = sprintf("%s\nPCo1 %.0f%%, PCo2 %.0f%%", n, rel[1], rel[2]))
}))
ord$panel <- factor(ord$panel, unique(ord$panel))
ggplot(ord, aes(PCo1, PCo2)) +
    geom_line(aes(group = subject), colour = "grey75") +
    geom_point(aes(colour = diet), size = 2) +
    facet_wrap(~panel, scales = "free") +
    labs(colour = NULL) +
    theme_bw() +
    theme(legend.position = "bottom")
```

Each grey line joins the casein and yeast samples of one mouse.

Both distances separate the diets with similar effect sizes, and for both
PERMDISP also detects unequal spread, in opposite directions: casein
samples are more dispersed under Bray-Curtis, yeast samples under MMD.
PERMANOVA with equal group sizes is fairly robust to unequal dispersion, and
the paired samples are displaced along the first axis in both ordinations,
so the separation is mainly a shift in location, with a difference in spread
on top. Under Bray-Curtis the second axis separates the two mouse groups,
which came from different vendor rooms.

# Representation space

`plot_tsne()` maps all proteins by cosine distance (one minus the predicted
TM-score), coloured by the study's manually curated broad function; proteins
without one are grey. The map shows neighbourhoods in representation space;
it is not evidence of function, and t-SNE does not preserve distances between
groups.

```{r tsne-function, fig.width = 7.5, fig.height = 4.6, fig.alt = "t-SNE of protein embeddings coloured by broad function"}
set.seed(1)
plot_tsne(x, color = setNames(ann$broad_function, ann$protein))
```

The study assigned a KO to
`r sprintf("%.0f%%", 100 * mean(!is.na(ann$ko)))` of the proteins. Proteins
without a KO have fewer close relatives in the catalog: the quartiles of each
protein's nearest-neighbour predicted TM-score, by KO status:

```{r nearest-neighbour}
ko_status <- ifelse(is.na(ann$ko), "no KO", "KO assigned")
Z <- S4Vectors::mcols(x)$embedding
Z <- Z / sqrt(rowSums(Z^2))
sim <- tcrossprod(Z)
diag(sim) <- NA
nearest <- apply(sim, 1, max, na.rm = TRUE)
rm(sim)
tapply(nearest, ko_status, quantile, probs = c(0.25, 0.5, 0.75))
```

# Protein bins

The similarity graph and MCL bins use the package defaults: an edge needs a
predicted TM-score of at least 0.7 and a local alignment covering at least
half of both sequences; MCL inflation is 2.

```{r bins}
S <- protein_similarity(x, n_cores = BiocParallel::multicoreWorkers())
b <- bin_proteins(S)
b$graph
sizes <- lengths(b$clusters)
c(proteins = length(x), bins = length(sizes),
    singleton_bins = sum(sizes == 1),
    singleton_fraction = round(sum(sizes == 1) / length(x), 3))
table(cut(sizes, c(0, 1, 2, 5, 10, 50, Inf),
    labels = c("1", "2", "3-5", "6-10", "11-50", ">50")))
```

The `r b$graph[["proteins"]]` proteins form `r b$graph[["components"]]`
connected components, the largest with
`r b$graph[["largest_component"]]` proteins; MCL splits the components into
`r length(sizes)` bins, `r sum(sizes == 1)` of them singletons.

## Sequence identity of graph edges {#edge-identity}

A random sample of 5,000 graph edges, aligned as in `protein_similarity()`
(local, BLOSUM62), by percent identity:

```{r edge-identity}
set.seed(1)
ij <- which(upper.tri(S) & S > 0, arr.ind = TRUE)
ij <- ij[sample(nrow(ij), 5000), ]
seqs <- x
S4Vectors::mcols(seqs) <- NULL
aln <- pwalign::pairwiseAlignment(seqs[ij[, 1]], seqs[ij[, 2]],
    substitutionMatrix = "BLOSUM62", type = "local",
    gapOpening = 11, gapExtension = 1)
identity <- pwalign::pid(aln, type = "PID1")
round(prop.table(table(cut(identity, c(0, 30, 40, 50, 70, 90, 100.01),
    right = FALSE, labels = c("<30%", "30-40%", "40-50%", "50-70%",
        "70-90%", ">=90%")))), 3)
```

Most edges join proteins with similar predicted structure but low sequence
identity (median `r round(median(identity))`%): remote homologs, which
sequence identity alone would not group.

# Differential abundance

Raw (unrarefied) protein counts are summed to bins with `bin_glom()` and
tested with DESeq2 [@love2014deseq2]. The design `~ subject + diet` pairs the
two samples of each mouse; size factors use the `poscounts` estimator because
no bin is observed in every sample. Fold changes are Yeast relative to Casein,
shrunk with the `normal` prior for display; p-values are from the Wald test.

```{r deseq}
binned <- bin_glom(counts, b)
dds <- DESeq2::DESeqDataSetFromMatrix(round(t(binned)), samples,
    design = ~ subject + diet)
dds <- DESeq2::DESeq(dds, sfType = "poscounts", quiet = TRUE)
res <- DESeq2::results(dds, name = "diet_Yeast_vs_Casein", alpha = 0.05)
lfc <- DESeq2::lfcShrink(dds, coef = "diet_Yeast_vs_Casein", type = "normal",
    res = res, quiet = TRUE)
DESeq2::summary(res)
```

```{r volcano, fig.width = 6, fig.height = 4, fig.alt = "Volcano plot of bin-level differential abundance, Yeast vs Casein"}
volcano <- data.frame(log2FC = lfc$log2FoldChange, p = res$pvalue,
    significant = !is.na(res$padj) & res$padj < 0.05)
ggplot(volcano[!is.na(volcano$p), ], aes(log2FC, -log10(p),
    colour = significant)) +
    geom_point(size = 1, alpha = 0.6) +
    scale_colour_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60"),
        labels = c(`TRUE` = "padj < 0.05", `FALSE` = "not significant")) +
    labs(x = "log2 fold change, Yeast vs Casein (shrunk)",
        y = expression(-log[10](p)), colour = NULL) +
    theme_bw()
```

Each bin's annotation summarises its members. A label shared by every
annotated member is shown with its count (`3/4` means three of four members
carry it, the fourth is unannotated); bins whose annotated members disagree
are marked `mixed` with the number of distinct labels and the most frequent
one.

```{r summarise}
summarise_labels <- function(v) {
    n <- length(v)
    v <- v[!is.na(v)]
    if (!length(v)) return("-")
    tab <- sort(table(v), decreasing = TRUE)
    top <- sprintf("%s (%d/%d)", strtrim(names(tab)[1], 40), tab[[1]], n)
    if (length(tab) == 1L) top else
        sprintf("mixed, %d: %s", length(tab), top)
}
bin_table <- function(bins) {
    data.frame(
        bin = bins,
        proteins = sizes[bins],
        taxon = vapply(b$clusters[bins],
            function(m) summarise_labels(ann[m, "taxon"]), ""),
        `function` = vapply(b$clusters[bins],
            function(m) summarise_labels(ann[m, "consensus_annotation"]), ""),
        KO = vapply(b$clusters[bins],
            function(m) summarise_labels(ann[m, "ko"]), ""),
        log2FC = round(lfc[bins, "log2FoldChange"], 2),
        p = formatC(res[bins, "pvalue"], format = "e", digits = 1),
        padj = formatC(res[bins, "padj"], format = "e", digits = 1),
        check.names = FALSE, row.names = NULL
    )
}
sig <- rownames(res)[!is.na(res$padj) & res$padj < 0.05]
sig <- sig[order(res[sig, "padj"])]
up_yeast <- head(sig[lfc[sig, "log2FoldChange"] > 0], 10)
up_casein <- head(sig[lfc[sig, "log2FoldChange"] < 0], 10)
```

The ten most significant bins higher on the yeast diet:

```{r table-yeast}
knitr::kable(bin_table(up_yeast))
```

The most significant yeast-enriched bin, `r up_yeast[1]`, holds
`r sizes[[up_yeast[1]]]` proteins that the study annotated as GH92 family
glycosyl hydrolases (α-mannosidases), almost all from *B. thetaiotaomicron*.
The source study reported these enzymes increasing on the yeast diet and
linked them to the mannan of yeast cell-wall glycoproteins
[@blakeleyruiz2025diet]. None of the bin's members has a KO, so a KO-level
summary would not have grouped them.

The ten most significant bins higher on the casein diet:

```{r table-casein}
knitr::kable(bin_table(up_casein))
```

Several casein-enriched bins are large and mixed across taxa, for example
substrate-binding proteins of carbohydrate ABC transporters: they summarise
a structural family shared by many organisms rather than one organism's
protein.

# Inside a bin

`plot_bin_similarity()` shows all pairwise predicted TM-scores within one
bin. A black outline marks a pair that **is** an edge of the graph: it passed
both thresholds (predicted TM-score ≥ 0.7, coverage ≥ 0.5) and keeps its
similarity as the edge weight. A cell without an outline is not an edge
(weight 0 in `S`). MCL clusters along the outlined edges. Shown are the most
significant bin with 4 to 30 members in each direction, with a count of
their member pairs: edges, pairs below the similarity threshold, and pairs
above it that failed the coverage gate.

```{r pick}
pick <- function(bins) bins[sizes[bins] >= 4 & sizes[bins] <= 30][1]
shown <- c(Yeast = pick(sig[lfc[sig, "log2FoldChange"] > 0]),
    Casein = pick(sig[lfc[sig, "log2FoldChange"] < 0]))
pair_types <- function(bin) {
    m <- b$clusters[[bin]]
    R <- tcrossprod(Z[m, ])
    E <- S[m, m] > 0
    up <- upper.tri(E)
    c(pairs = sum(up), edges = sum(E[up]), below_0.7 = sum(R[up] < 0.7),
        failed_coverage = sum(R[up] >= 0.7 & !E[up]))
}
data.frame(bin = shown, t(vapply(shown, pair_types, numeric(4))))
```

```{r bin-yeast, fig.width = 7, fig.height = 6, fig.alt = "Pairwise predicted TM-scores in the selected yeast-enriched bin; edges outlined"}
plot_bin_similarity(b, shown[["Yeast"]], S, x)
```

```{r members-yeast}
m <- b$clusters[[shown[["Yeast"]]]]
knitr::kable(ann[m, c("taxon", "consensus_annotation", "ko", "cazy")],
    row.names = TRUE)
```

```{r bin-casein, fig.width = 7, fig.height = 6, fig.alt = "Pairwise predicted TM-scores in the selected casein-enriched bin; edges outlined"}
plot_bin_similarity(b, shown[["Casein"]], S, x)
```

```{r members-casein}
m <- b$clusters[[shown[["Casein"]]]]
knitr::kable(ann[m, c("taxon", "consensus_annotation", "ko", "cazy")],
    row.names = TRUE)
```

In the GH92 bin every pair of members is an edge. In
`r shown[["Casein"]]`, mostly unannotated proteins from several
*Acetatifactor* species groups and unbinned contigs, some members are joined only
through shared neighbours (pairs below 0.7) and some similar pairs are not
edges because their alignment covers less than half of one sequence. MCL
keeps both kinds of pair in one bin; the thresholds act on edges, not on
every pair.

# Notes

- Diet is confounded with feeding period (weeks 2 and 5) in this serial
  design, so the differences cannot be attributed to diet alone.
- Beta diversity uses rarefied counts; differential abundance uses raw
  counts with DESeq2 size factors.
- A bin is a similarity-network cluster. Agreement among its members'
  annotations is evidence about its function; bins can span species and,
  when large, related functions.
- KO assignments are the study's own (Mantis, else MicrobeAnnotator). A
  missing KO does not mean an unknown function, as the GH92 bin shows.

```{r session}
sessionInfo()
```

# References {-}
