---
title: "Real-data validation: the mouse dietary-protein metaproteome"
author: "Yixuan Yang"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
```

The [simulation benchmark](simulation.html) tests the method against a known
functional truth. This page runs the same workflow on a real metaproteome,
where there is no such truth, to check that it behaves sensibly at the size
of a full catalog:

1. **Beta diversity** with RBF-MMD and PERMANOVA.
2. **MCL bins** on the TM-Vec similarity graph at the package defaults.
3. **Differential abundance** on the bins with DESeq2.
4. **Functional reading** of the most responsive annotated bins.

```{r libs}
library(repdist)
library(ggplot2)
library(phyloseq)
```

# The full dataset

The study is Blakeley-Ruiz *et al.*, *ISME J* **19**(1) wraf048: mouse faecal
metaproteomes across six dietary protein sources, quantified by spectral
counting.

```{r data}
path <- file.path("data", "test_study")
MAX_AA <- 3000L

ps <- readRDS(file.path(path, "ps.rds"))
floor_depth <- min(sample_sums(ps))

set.seed(1)
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)
psd <- prune_samples(sample_data(psd)$phase != "baseline", psd)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)

n_after_rarefaction <- ntaxa(psd)
protein_length <- Biostrings::width(refseq(psd)[taxa_names(psd)])
too_long <- protein_length > MAX_AA
long_abundance <- sum(taxa_sums(psd)[too_long]) / sum(taxa_sums(psd))
psd <- prune_taxa(!too_long, psd)

keep <- taxa_names(psd)
psu <- prune_samples(sample_data(ps)$phase != "baseline", ps)
psu <- prune_taxa(taxa_names(psu) %in% keep, psu)

counts <- Matrix::Matrix(t(as(otu_table(psd), "matrix")), sparse = TRUE)
counts_raw <- t(as(otu_table(psu), "matrix"))[rownames(counts), keep, drop = FALSE]

md <- data.frame(sample_data(psd))[rownames(counts), ]
md$diet <- factor(md$diet)
md$dose <- factor(md$dose)
md$subject <- factor(md$subject)

# Annotations from Extended Data Table 6: join on Protein Identifier, fall
# back to the species-code column, and keep a match only where the residue
# count agrees.
t6 <- read.delim(file.path(path, "annotated",
                           "ExtendedDataTable6AnnotatedMicrobialMetaproteomes.txt"),
                 skip = 2, quote = "\"", check.names = FALSE,
                 colClasses = "character", comment.char = "", fileEncoding = "latin1")
len <- Biostrings::width(refseq(psd)[keep])
agrees <- function(r) !is.na(r) & suppressWarnings(as.integer(t6$AA[r])) == len
r_id <- match(keep, t6[["Protein Identifier"]])
r_code <- match(keep, t6[["Protein Identifier with Species Code"]])
row <- ifelse(agrees(r_id), r_id, ifelse(agrees(r_code), r_code, NA))
annotation <- t6[["Consensus annotation"]][row]
annotation[!nzchar(trimws(annotation)) | tolower(annotation) == "unknown"] <- NA
taxa <- data.frame(
  genus = as(tax_table(psd), "matrix")[keep, "Genus"],
  annotation = annotation,
  row.names = keep, stringsAsFactors = FALSE)

d <- list(counts = counts, counts_raw = counts_raw, metadata = md, taxa = taxa)

c(samples = nrow(counts), proteins = ncol(counts),
  removed_over_3000aa = sum(too_long),
  removed_abundance_pct = round(100 * long_abundance, 3))
table(md$diet)
```

Rarefaction provides the abundance filter: proteins absent after subsampling
are removed. It retained `r n_after_rarefaction` proteins; the 3,000aa ceiling
then removed `r sum(too_long)` more, representing only
`r sprintf("%.3f", 100 * long_abundance)`% of rarefied abundance. The count
matrix is sparse; `counts_raw` contains unrarefied counts for exactly the same
proteins and samples.

**The ceiling is a choice made here, not a limit `repdist` imposes.**
`embed_proteins()` embeds every sequence whole -- it never truncates and never
silently drops -- and no function in the package takes a maximum length. The
reason to cut at 3,000 residues is TM-Vec: its head was trained on structure
pairs through 1,000 residues, so a predicted TM-score for a chain many times
longer is extrapolation well outside what the model saw. At
`r sprintf("%.3f", 100 * long_abundance)`% of abundance the cut costs
essentially nothing here, which is what makes it worth taking. On a catalog
where long proteins carry real signal, raise it or drop it.

# Embedding

`embed_proteins()` embeds every sequence whole. The full TM-Vec embeddings are
cached outside the render because model inference is not part of the statistical
analysis.

```{r Z}
cached <- readRDS(file.path(path, "embeddings_tmvec1_large.rds"))
x <- refseq(psd)[keep]
stopifnot(identical(as.character(x), cached$seq[keep]))
S4Vectors::mcols(x)$embedding <- cached$Z[keep, ]
rm(cached)
x
```

`x` is the `AAStringSet` every `repdist` function takes: the sequences, with
one embedding row per protein in `mcols(x)$embedding`. TM-Vec's head was
trained to regress TM-score, so cosine similarity between two embeddings is a
predicted TM-score. `known_models()` lists the other registered
representations.

```{r models}
names(known_models())
```

# Step 1: beta diversity

Bray-Curtis is the protein-identity baseline. MMD asks how far apart different
proteins actually are. `sample_mmd()` builds a Euclidean RBF kernel from the
embeddings, so the kernel remains positive semi-definite; on unit-norm TM-Vec
embeddings that distance is a monotone transform of predicted TM-score.

```{r distances}
D <- list(
  `Bray-Curtis` = vegan::vegdist(as.matrix(d$counts), "bray"),
  `MMD` = sample_mmd(d$counts, x))
```

Both are abundance-weighted, so richness is checked before interpreting the
diet effect.

```{r qc-richness}
rich <- vegan::specnumber(as.matrix(d$counts))
round(vapply(D, function(x) cor(rich, rowMeans(as.matrix(x))), numeric(1)), 2)
```

## PERMANOVA

Amount of protein, sex and age remain as covariates and diet enters last.
Permutations are blocked within mouse.

```{r permanova}
perm <- permute::how(blocks = md$subject, nperm = 999)

res <- do.call(rbind, lapply(names(D), function(n) {
  set.seed(1)
  a <- vegan::adonis2(D[[n]] ~ dose + sex + age_week + diet, data = md,
                      by = "terms", permutations = perm)
  data.frame(distance = n, R2 = round(a["diet", "R2"], 3),
             F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"])
}))
knitr::kable(res, row.names = FALSE)
```

## Ordination

```{r pcoa, fig.height = 3.6, fig.width = 6.5, fig.alt = "PCoA under Bray-Curtis and RBF-MMD, coloured by dietary protein source"}
ord <- do.call(rbind, lapply(names(D), function(n) {
  e <- cmdscale(D[[n]], k = 2, eig = TRUE)
  rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
  data.frame(
    Axis1 = e$points[, 1], Axis2 = e$points[, 2],
    diet = md[rownames(e$points), "diet"],
    panel = sprintf("%s  (PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2]))
}))

ggplot(ord, aes(Axis1, Axis2, colour = diet)) +
  geom_point(size = 1.6, alpha = 0.85) +
  stat_ellipse(type = "t", linewidth = 0.3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "PCo1", y = "PCo2", colour = "diet") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Step 2: MCL bins

`protein_similarity()` keeps a pair as an edge when its predicted TM-score
is at least `min_sim` (0.7 by default) and its local alignment covers at
least half of both sequences, so two proteins sharing only one domain are not
joined. `bin_proteins()` runs Markov clustering on the resulting graph at
inflation 2; proteins with no edge stay singletons. The floor decides which
pairs connect, not what a bin holds: a bin can hold pairs below it, joined
through shared neighbours.

```{r ground}
S <- protein_similarity(x, n_cores = BiocParallel::multicoreWorkers())
```

```{r bins}
b <- bin_proteins(S, inflation = 2)
b$graph
```

The unrarefied counts sum to bin level with `bin_glom()`.

```{r bin-counts}
bin_cnt <- bin_glom(d$counts_raw, b)
c(proteins = ncol(d$counts_raw), bins = ncol(bin_cnt))
```

```{r bin-sizes}
sizes <- lengths(b$clusters)
table(cut(sizes, c(0, 1, 2, 5, 10, Inf),
          labels = c("1", "2", "3-5", "6-10", ">10")))
```

`r sum(sizes == 1)` proteins remain singleton bins. The other
`r sum(sizes > 1)` bins are communities of proteins connected above the
floor, reducing the number of tested units by a factor of
`r round(ncol(d$counts_raw) / ncol(bin_cnt), 1)`.

# Step 3: differential abundance with DESeq2

DESeq2 fits a negative-binomial model per bin on unrarefied counts. The paired
design blocks on subject, and diet is tested by likelihood ratio against the
model without diet.

```{r deseq}
dds <- DESeq2::DESeqDataSetFromMatrix(t(bin_cnt), md,
                                      design = ~ subject + diet)
dds <- DESeq2::DESeq(dds, test = "LRT", reduced = ~ subject,
                     sfType = "poscounts", quiet = TRUE)
res_lrt <- DESeq2::results(dds)

c(bins_tested = nrow(res_lrt),
  diet_significant = sum(res_lrt$padj <= 0.05, na.rm = TRUE))
```

Each significant bin is assigned its largest shrunken diet-vs-Casein fold
change for ranking and display.

```{r effects}
lev <- setdiff(levels(md$diet), "Casein")
ddsw <- DESeq2::nbinomWaldTest(dds, quiet = TRUE)
shrunk <- lapply(lev, function(l)
  DESeq2::lfcShrink(ddsw, coef = paste0("diet_", l, "_vs_Casein"),
                    type = "normal", quiet = TRUE))
names(shrunk) <- lev
fc <- vapply(shrunk, function(o) o$log2FoldChange, numeric(nrow(res_lrt)))
rownames(fc) <- rownames(res_lrt)

sig <- rownames(res_lrt)[!is.na(res_lrt$padj) & res_lrt$padj <= 0.05]
top <- data.frame(
  bin = sig,
  members = sizes[sig],
  padj = formatC(res_lrt[sig, "padj"], digits = 2, format = "e"),
  diet = lev[apply(abs(fc[sig, , drop = FALSE]), 1, which.max)],
  log2FC = round(fc[cbind(sig, lev[apply(abs(fc[sig, , drop = FALSE]),
                                        1, which.max)])], 2))
top <- top[order(-abs(top$log2FC)), ]
knitr::kable(head(top, 8), row.names = FALSE)
```

```{r volcano, fig.height = 3.4, fig.width = 9, fig.alt = "Diet-vs-Casein volcano plots for structural bins"}
volc <- do.call(rbind, lapply(lev, function(l) {
  o <- shrunk[[l]]
  data.frame(bin = rownames(o), diet = l, log2FoldChange = o$log2FoldChange,
             pvalue = o$pvalue, reject = !is.na(o$padj) & o$padj <= 0.05)
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                     guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) +
  theme_bw()
```

# Step 4: responsive bins, read functionally

Select the three most responsive multi-member bins with at least two annotated
proteins.

```{r pick}
n_ann <- vapply(b$clusters[top$bin],
                function(m) sum(!is.na(d$taxa[m, "annotation"])), integer(1))
picks <- head(top[top$members > 1 & n_ann >= 2, ], 3)
knitr::kable(picks, row.names = FALSE)
```

```{r members}
info <- do.call(rbind, lapply(seq_len(nrow(picks)), function(i) {
  m <- b$clusters[[picks$bin[i]]]
  data.frame(
    bin = picks$bin[i],
    genus = d$taxa[m, "genus"],
    annotation = substr(d$taxa[m, "annotation"], 1, 52),
    mean_Casein = round(colMeans(d$counts_raw[md$diet == "Casein", m,
                                               drop = FALSE]), 1),
    mean_top = round(colMeans(d$counts_raw[md$diet == picks$diet[i], m,
                                           drop = FALSE]), 1))
}))
knitr::kable(info, row.names = FALSE)
```

An MCL bin makes no all-pairs promise: members are joined through the
graph, so a pair inside a bin need not be an edge. The within-bin predicted
TM-scores of the leading annotated bin, with the graph's edges outlined:

```{r bin-similarity, fig.height = 4.5, fig.alt = "Pairwise predicted TM-scores inside the leading annotated bin, edges outlined"}
plot_bin_similarity(b, picks$bin[1], S, x)
```

A bin is a community in a similarity graph. Agreement among its members'
annotations is evidence about its function; membership alone is not.

# Practical notes

**Rarefy first** for whole-community distances. The differential-abundance
model uses the corresponding unrarefied counts and handles depth through size
factors.

**Memory.** `protein_similarity()` and `sample_mmd()` hold dense
protein-by-protein matrices, so memory grows with the square of the protein
count; bin the identified subset of a catalog, never a whole search database.
MCL's cost follows the largest connected component, which the `MCL` package
holds as one dense matrix.

**Bandwidth.** `sample_mmd()` returns its bandwidth as `attr(, "sigma")`;
pass it back when comparisons across studies need one fixed kernel.

```{r session}
sessionInfo()
```
