---
title: "Case study"
subtitle: "Mouse gut metaproteomes across a dietary-protein series"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

This page runs the `repdist` workflow on a published metaproteome: beta
diversity with MMD (`sample_mmd()`, the maximum mean discrepancy under a
Gaussian RBF kernel on the protein embeddings), the representation space,
MCL bins, and bin-level differential abundance.

```{r libs}
library(repdist)
library(ggplot2)
library(vegan)
```

# Study and comparison

Blakeley-Ruiz *et al.* [@blakeleyruiz2025diet] fed twelve mice, four cages of
three, a series of defined diets, one week each, and measured faecal
metaproteomes by spectral counting (PRIDE PXD041586). The males (C1, C3) and
females (C2, C4) came from different rooms at the Jackson Laboratory, so sex
also marks two starting microbiomes, which the paper calls mouse groups.

**The input sequences are already sequence-reduced.** The microbial protein
catalog comes from metagenome-assembled genomes annotated with PROKKA;
unbinned genes were added only if no binned gene matched them at 90% identity
(CD-HIT-2D), and the host and diet reference proteomes were clustered at 95%
identity. Every reported protein also has at least one protein-unique
peptide. Near-identical proteins are therefore rare in the table (see
[sequence identity of graph edges](#edge-identity)): the relatedness that MMD
and the bins add beyond protein identity is mostly remote homology.

**Beta diversity** compares the six dietary protein sources (soy, casein,
rice, yeast, pea and egg) across the series. **The bins and differential
abundance** compare two diets of the same protein amount, **20% yeast protein
(Yeast) against 20% casein (Casein)**, casein being the reference protein of
the series, in the 11 mice sampled on both. Diets were fed in a fixed order,
one week each, so protein source is partly confounded with feeding week.

```{r data}
d <- readRDS(file.path("curated", "diet_series.rds"))
proteins <- d$proteins # sequences with TM-Vec embeddings
counts <- d$counts # raw PSM counts, samples x proteins
samples <- d$samples
ann <- d$annotation
table(samples$diet)
```

The curated object was built by `data-raw/diet_series.R` from the paper's
supplementary Data Set 1 (PSM counts and sequences) and Data Set 6
(annotation): every microbial protein with at least one PSM in any sample,
its species group and GTDB lineage, the study's curated function,
KEGG orthology (KO) assignments and CAZy families, and a TM-Vec embedding
computed with `embed_proteins()`.

**Protein length.** Proteins longer than `MAX_AA` = `r d$provenance$max_aa`
residues were removed before embedding:
`r d$provenance$steps[["longer_than_max"]]` of
`r d$provenance$steps[["detected"]]` proteins, carrying
`r sprintf("%.1f%%", 100 * d$provenance$long_protein_psm_share)` of the PSMs.
This is an analysis choice, not a package limit: `embed_proteins()` embeds
sequences of any length, but the TM-Vec head was trained on chains of up to
1000 residues, so predicted TM-scores for much longer chains are
extrapolations.

```{r dims}
c(samples = nrow(counts), proteins = ncol(counts))
```

# Beta diversity across protein sources

The samples are named by diet token (percentage and protein source). The
`T0` baseline on standard chow is left out. The 40% soy and 40% casein weeks
count under their protein source, and so does the `CTL` re-feeding at the end
of the series: 20% soy in cages C1 and C2, 20% casein in C3 and C4.

```{r sources}
token <- as.character(samples$diet)
protein_source <- unname(c(SOY = "Soy", CAS = "Casein", RIC = "Rice",
    YST = "Yeast", PEA = "Pea", EGG = "Egg")[sub("^[0-9]+", "", token)])
ctl <- token == "CTL"
protein_source[ctl] <- ifelse(samples$cage[ctl] %in% c("C1", "C2"), "Soy",
    "Casein")
md <- data.frame(diet = protein_source,
    dose = ifelse(startsWith(token, "40"), "40", "20"),
    cage = samples$cage, subject = samples$subject,
    # the stopped chicken-bone week elapsed before CTL
    week = ifelse(ctl, 10L, samples$period),
    row.names = samples$sample)[token != "T0", ]
md$diet <- factor(md$diet)
md$dose <- factor(md$dose)
md <- droplevels(md)
table(md$diet, md$dose)
```

Sequencing depth differs between samples, and depth changes which proteins
are observed. Both distances are therefore computed on counts rarefied to the
smallest sample.

```{r rarefy}
range(rowSums(counts[rownames(md), ]))
set.seed(1)
rarefied <- rrarefy(counts[rownames(md), ], min(rowSums(counts[rownames(md), ])))
rarefied <- rarefied[, colSums(rarefied) > 0]
D <- list(
    `Bray-Curtis (protein)` = vegdist(rarefied, "bray"),
    MMD = sample_mmd(rarefied, proteins)
)
```

PERMANOVA [@anderson2001permanova] tests protein source after protein amount,
cage and week, and PERMDISP [@anderson2006permdisp] tests for unequal spread
between sources; both use 999 permutations restricted to within mouse.
Cage nests sex. Its R² is reported but not tested: within-mouse permutations
cannot test a between-mouse term.

```{r permanova}
perm <- permute::how(blocks = md$subject, nperm = 999)
tests <- t(vapply(D, function(Dk) {
    set.seed(1)
    a <- adonis2(Dk ~ dose + cage + week + diet, data = md, by = "terms",
        permutations = perm)
    set.seed(1)
    pd <- permutest(betadisper(Dk, md$diet), permutations = perm)
    c(`PERMANOVA R2` = a["diet", "R2"], `PERMANOVA p` = a["diet", "Pr(>F)"],
        `PERMDISP p` = pd$tab[1, "Pr(>F)"], `Cage R2` = a["cage", "R2"])
}, numeric(4)))
knitr::kable(tests, digits = 3)
```

Ellipses are 95% ellipses per protein source (multivariate t).

```{r pcoa, fig.width = 9, fig.height = 4.2, fig.alt = "PCoA under protein-level Bray-Curtis and MMD, coloured by protein source, shaped by cage, with 95% ellipses"}
ord <- do.call(rbind, lapply(names(D), function(n) {
    e <- cmdscale(D[[n]], k = 2, eig = TRUE)
    rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
    data.frame(PCo1 = e$points[, 1], PCo2 = e$points[, 2],
        diet = md$diet, cage = md$cage,
        panel = sprintf("%s\nPCo1 %.0f%%, PCo2 %.0f%%", n, rel[1], rel[2]))
}))
ord$panel <- factor(ord$panel, unique(ord$panel))
ggplot(ord, aes(PCo1, PCo2, colour = diet, shape = cage)) +
    geom_point(size = 1.8) +
    stat_ellipse(aes(group = diet), linewidth = 0.4) +
    facet_wrap(~panel, scales = "free") +
    labs(colour = NULL, shape = NULL) +
    theme_bw()
```

Both distances detect protein source. MMD attributes more of the variation
to it than protein-level Bray-Curtis does (R²
`r sprintf("%.3f", tests["MMD", "PERMANOVA R2"])` against
`r sprintf("%.3f", tests["Bray-Curtis (protein)", "PERMANOVA R2"])`), and
only Bray-Curtis also shows unequal spread between sources (PERMDISP p
`r sprintf("%.3f", tests["Bray-Curtis (protein)", "PERMDISP p"])` against
`r sprintf("%.3f", tests["MMD", "PERMDISP p"])`), so the MMD separation is
mainly a shift in location.

Cage explains less under MMD (R²
`r sprintf("%.3f", tests["MMD", "Cage R2"])` against
`r sprintf("%.3f", tests["Bray-Curtis (protein)", "Cage R2"])`), and almost
all of it is the split between the two mouse groups: the first Bray-Curtis
axis separates C1 and C3 from C2 and C4, while the first MMD axis separates
yeast and egg from the other sources.

To see what separates the mouse groups, the next chunk sets aside proteins
seen in only one group and finds each one's nearest protein in the other
group by TM-Vec similarity.

```{r group-private}
group <- ifelse(md$cage %in% c("C1", "C3"), "A", "B")
psm <- rowsum(counts[rownames(md), ], group)
mine <- ifelse(psm["B", ] == 0, "A", ifelse(psm["A", ] == 0, "B", "shared"))
group_private <- c(A = sum(psm["A", mine == "A"]) / sum(psm["A", ]),
    B = sum(psm["B", mine == "B"]) / sum(psm["B", ]))
# nearest protein seen in the other group, for a PSM-weighted sample of 300
# group-private proteins per group
Z <- S4Vectors::mcols(proteins)$embedding
Z <- Z / sqrt(rowSums(Z^2))
seqs <- proteins
S4Vectors::mcols(seqs) <- NULL
set.seed(1)
nn <- do.call(rbind, lapply(c("A", "B"), function(g) {
    own <- which(mine == g)
    other <- which(psm[setdiff(c("A", "B"), g), ] > 0)
    i <- sample(own, 300, prob = psm[g, own])
    cs <- Z[i, ] %*% t(Z[other, ])
    j <- other[max.col(cs, "first")]
    aln <- pwalign::pairwiseAlignment(seqs[i], seqs[j],
        substitutionMatrix = "BLOSUM62", type = "local",
        gapOpening = 11, gapExtension = 1)
    data.frame(group = g, tm = apply(cs, 1, max),
        identity = pwalign::pid(aln, type = "PID1"))
}))
round(prop.table(table(cut(nn$identity, c(0, 30, 50, 70, 90, 100.01),
    right = FALSE, labels = c("<30%", "30-50%", "50-70%", "70-90%",
        ">=90%")))), 3)
```

The two mouse groups, A (C1, C3) and B (C2, C4), carry different variants
of the same protein families. Proteins seen in one group only hold
`r sprintf("%.0f%%", 100 * group_private["A"])` of A's and
`r sprintf("%.0f%%", 100 * group_private["B"])` of B's PSMs, and their
nearest protein in the other group is a same-fold homolog (median predicted
TM-score `r sprintf("%.2f", median(nn$tm))`) at under 90% identity in
`r sprintf("%.0f%%", 100 * mean(nn$identity < 90))` of cases. Bray-Curtis
counts these as different proteins and MMD counts them as close.

# The Yeast–Casein contrast

From here on the page uses the 11 mice sampled on both 20% casein and 20%
yeast, and the proteins seen in those 22 samples.

```{r contrast}
yc <- samples$diet %in% c("20CAS", "20YST")
yc <- yc & samples$subject %in% names(which(table(samples$subject[yc]) == 2))
samples <- droplevels(samples[yc, ])
samples$diet <- factor(ifelse(samples$diet == "20YST", "Yeast", "Casein"),
    c("Casein", "Yeast"))
counts <- counts[yc, ]
counts <- counts[, colSums(counts) > 0]
proteins <- proteins[colnames(counts)]
ann <- ann[colnames(counts), ]
c(samples = nrow(counts), proteins = ncol(counts))
```

# Representation space

`plot_tsne()` maps the contrast's proteins by cosine distance (one minus the predicted
TM-score), coloured by the study's manually curated broad function; proteins
without one are grey. The map shows neighbourhoods in representation space;
it is not evidence of function, and t-SNE does not preserve distances between
groups.

```{r tsne-function, fig.width = 7.5, fig.height = 4.6, fig.alt = "t-SNE of protein embeddings coloured by broad function"}
set.seed(1)
plot_tsne(proteins, color = setNames(ann$broad_function, ann$protein))
```

The study assigned a KO to
`r sprintf("%.0f%%", 100 * mean(!is.na(ann$ko)))` of the proteins. Proteins
without a KO have fewer close relatives in the catalog: the quartiles of each
protein's nearest-neighbour predicted TM-score, by KO status:

```{r nearest-neighbour}
ko_status <- ifelse(is.na(ann$ko), "no KO", "KO assigned")
Z <- S4Vectors::mcols(proteins)$embedding
Z <- Z / sqrt(rowSums(Z^2))
sim <- tcrossprod(Z)
diag(sim) <- NA
nearest <- apply(sim, 1, max, na.rm = TRUE)
rm(sim)
tapply(nearest, ko_status, quantile, probs = c(0.25, 0.5, 0.75))
```

# Protein bins

The similarity graph and MCL bins use the package defaults: an edge needs a
predicted TM-score of at least 0.7 and a local alignment covering at least
half of both sequences; MCL inflation is 2.

```{r bins}
S <- protein_similarity(proteins, n_cores = BiocParallel::multicoreWorkers())
bins <- bin_proteins(S)
bins$graph
sizes <- lengths(bins$clusters)
c(proteins = length(proteins), bins = length(sizes),
    singleton_bins = sum(sizes == 1),
    singleton_fraction = round(sum(sizes == 1) / length(proteins), 3))
table(cut(sizes, c(0, 1, 2, 5, 10, 50, Inf),
    labels = c("1", "2", "3-5", "6-10", "11-50", ">50")))
```

The `r bins$graph[["proteins"]]` proteins form `r bins$graph[["components"]]`
connected components, the largest with
`r bins$graph[["largest_component"]]` proteins; MCL splits the components into
`r length(sizes)` bins, `r sum(sizes == 1)` of them singletons.

## Sequence identity of graph edges {#edge-identity}

A random sample of 5,000 graph edges, aligned as in `protein_similarity()`
(local, BLOSUM62), by percent identity:

```{r edge-identity}
set.seed(1)
ij <- which(upper.tri(S) & S > 0, arr.ind = TRUE)
ij <- ij[sample(nrow(ij), 5000), ]
seqs <- proteins
S4Vectors::mcols(seqs) <- NULL
aln <- pwalign::pairwiseAlignment(seqs[ij[, 1]], seqs[ij[, 2]],
    substitutionMatrix = "BLOSUM62", type = "local",
    gapOpening = 11, gapExtension = 1)
identity <- pwalign::pid(aln, type = "PID1")
round(prop.table(table(cut(identity, c(0, 30, 40, 50, 70, 90, 100.01),
    right = FALSE, labels = c("<30%", "30-40%", "40-50%", "50-70%",
        "70-90%", ">=90%")))), 3)
```

Most edges join proteins with similar predicted structure but low sequence
identity (median `r round(median(identity))`%): remote homologs, which
sequence identity alone would not group.

# Differential abundance

Raw (unrarefied) protein counts are summed to bins with `bin_glom()` and
tested with DESeq2 [@love2014deseq2]. The design `~ subject + diet` pairs the
two samples of each mouse; size factors use the `poscounts` estimator because
no bin is observed in every sample. Fold changes are Yeast relative to Casein,
shrunk with the `normal` prior for display; p-values are from the Wald test.

```{r deseq}
bin_counts <- bin_glom(counts, bins)
dds <- DESeq2::DESeqDataSetFromMatrix(round(t(bin_counts)), samples,
    design = ~ subject + diet)
dds <- DESeq2::DESeq(dds, sfType = "poscounts", quiet = TRUE)
res <- DESeq2::results(dds, name = "diet_Yeast_vs_Casein", alpha = 0.05)
lfc <- DESeq2::lfcShrink(dds, coef = "diet_Yeast_vs_Casein", type = "normal",
    res = res, quiet = TRUE)
DESeq2::summary(res)
```

DESeq2's MA plot shows each bin's shrunk log2 fold change against its mean
normalised count; bins with adjusted p < 0.05 are coloured.

```{r ma-plot, fig.width = 6, fig.height = 4, fig.alt = "MA plot of bin-level differential abundance, Yeast vs Casein"}
DESeq2::plotMA(lfc, alpha = 0.05,
    ylab = "log2 fold change, Yeast vs Casein (shrunk)")
```

Each bin's annotation summarises its members. A label shared by every
annotated member is shown with its count (`3/4` means three of four members
carry it, the fourth is unannotated); bins whose annotated members disagree
are marked `mixed` with the number of distinct labels and the most frequent
one.

```{r summarise}
summarise_labels <- function(v) {
    n <- length(v)
    v <- v[!is.na(v)]
    if (!length(v)) return("-")
    tab <- sort(table(v), decreasing = TRUE)
    top <- sprintf("%s (%d/%d)", strtrim(names(tab)[1], 40), tab[[1]], n)
    if (length(tab) == 1L) top else
        sprintf("mixed, %d: %s", length(tab), top)
}
bin_table <- function(ids) {
    data.frame(
        bin = ids,
        proteins = sizes[ids],
        taxon = vapply(bins$clusters[ids],
            function(m) summarise_labels(ann[m, "taxon"]), ""),
        `function` = vapply(bins$clusters[ids],
            function(m) summarise_labels(ann[m, "consensus_annotation"]), ""),
        KO = vapply(bins$clusters[ids],
            function(m) summarise_labels(ann[m, "ko"]), ""),
        log2FC = round(lfc[ids, "log2FoldChange"], 2),
        p = formatC(res[ids, "pvalue"], format = "e", digits = 1),
        padj = formatC(res[ids, "padj"], format = "e", digits = 1),
        check.names = FALSE, row.names = NULL
    )
}
sig <- rownames(res)[!is.na(res$padj) & res$padj < 0.05]
sig <- sig[order(res[sig, "padj"])]
up_yeast <- head(sig[lfc[sig, "log2FoldChange"] > 0], 10)
up_casein <- head(sig[lfc[sig, "log2FoldChange"] < 0], 10)
```

The ten most significant bins higher on the yeast diet:

```{r table-yeast}
knitr::kable(bin_table(up_yeast))
```

The most significant yeast-enriched bin, `r up_yeast[1]`, holds
`r sizes[[up_yeast[1]]]` proteins that the study annotated as GH92 family
glycosyl hydrolases (α-mannosidases), almost all from *B. thetaiotaomicron*.
The source study reported these enzymes increasing on the yeast diet and
linked them to the mannan of yeast cell-wall glycoproteins
[@blakeleyruiz2025diet]. None of the bin's members has a KO, so a KO-level
summary would not have grouped them.

The ten most significant bins higher on the casein diet:

```{r table-casein}
knitr::kable(bin_table(up_casein))
```

Several casein-enriched bins are large and mixed across taxa, for example
substrate-binding proteins of carbohydrate ABC transporters: they summarise
a structural family shared by many organisms rather than one organism's
protein.

# Inside a bin

`plot_bin_similarity()` shows all pairwise predicted TM-scores within one
bin. An uncrossed cell is an edge of the graph: the pair passed both
thresholds (predicted TM-score ≥ 0.7, coverage ≥ 0.5) and keeps its
similarity as the edge weight. A crossed cell is not an edge (weight 0 in
`S`). MCL clusters along the edges. Shown are the most
significant bin with 4 to 30 members in each direction, with a count of
their member pairs: edges, pairs below the similarity threshold, and pairs
above it that failed the coverage gate.

```{r pick}
pick <- function(ids) ids[sizes[ids] >= 4 & sizes[ids] <= 30][1]
shown <- c(Yeast = pick(sig[lfc[sig, "log2FoldChange"] > 0]),
    Casein = pick(sig[lfc[sig, "log2FoldChange"] < 0]))
pair_types <- function(bin) {
    m <- bins$clusters[[bin]]
    R <- tcrossprod(Z[m, ])
    E <- S[m, m] > 0
    up <- upper.tri(E)
    c(pairs = sum(up), edges = sum(E[up]), below_0.7 = sum(R[up] < 0.7),
        failed_coverage = sum(R[up] >= 0.7 & !E[up]))
}
data.frame(bin = shown, t(vapply(shown, pair_types, numeric(4))))
```

```{r bin-yeast, fig.width = 7, fig.height = 6, fig.alt = "Pairwise predicted TM-scores in the selected yeast-enriched bin; pairs that are not edges crossed"}
plot_bin_similarity(bins, shown[["Yeast"]], S, proteins)
```

```{r members-yeast}
m <- bins$clusters[[shown[["Yeast"]]]]
knitr::kable(ann[m, c("taxon", "consensus_annotation", "ko", "cazy")],
    row.names = TRUE)
```

```{r bin-casein, fig.width = 7, fig.height = 6, fig.alt = "Pairwise predicted TM-scores in the selected casein-enriched bin; pairs that are not edges crossed"}
plot_bin_similarity(bins, shown[["Casein"]], S, proteins)
```

```{r members-casein}
m <- bins$clusters[[shown[["Casein"]]]]
knitr::kable(ann[m, c("taxon", "consensus_annotation", "ko", "cazy")],
    row.names = TRUE)
```

In the GH92 bin every pair of members is an edge. In
`r shown[["Casein"]]`, mostly unannotated proteins from several
*Acetatifactor* species groups and unbinned contigs, some members are joined only
through shared neighbours (pairs below 0.7) and some similar pairs are not
edges because their alignment covers less than half of one sequence. MCL
keeps both kinds of pair in one bin; the thresholds act on edges, not on
every pair.

# Notes

- Diet is confounded with feeding period in this serial design (Casein and
  Yeast were periods 2 and 5), so the differences cannot be attributed to
  diet alone.
- Beta diversity uses rarefied counts; differential abundance uses raw
  counts with DESeq2 size factors.
- A bin is a similarity-network cluster. Agreement among its members'
  annotations is evidence about its function; bins can span species and,
  when large, related functions.
- KO assignments are the study's own (Mantis, else MicrobeAnnotator). A
  missing KO does not mean an unknown function, as the GH92 bin shows.

```{r session}
sessionInfo()
```

# References {-}
