---
title: "Data curation: Extended Data Table 1 to a phyloseq object"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE, fig.width = 7)
```

```{r libs, cache = FALSE, message = FALSE}
library(phyloseq)
library(vegan)
library(permute)
HERE <- file.path("data", "test_study")
```

Source: Blakeley-Ruiz *et al.*, *ISME J* **19**(1), wraf048, "Dietary protein source
alters gut microbiota composition and function".

**Extended Data Table 1** is the abundance source. Its sample columns are
peptide-spectral-match **counts** — unrarefied spectral counts summed per taxon is what
the paper calls "proteinaceous biomass" (see below) — and it carries amino-acid
sequences inline. **Extended Data Table 6** publishes microbe-restricted %NSAF, a
different abundance normalization not used here; it is loaded only for its curated GTDB
lineage per protein, joined onto Table 1's proteins by id.

This report builds the phyloseq object and checks it against the published PERMANOVA.
If that check fails nothing downstream is worth running, so it lives here rather than
in the analysis. Output: `data/test_study/ps.rds`.

# Reading the table

Two title lines, a header on line 3, five annotation columns, then one column per
sample. Blank cells are zeros. Column 1 codes each protein `Diet`, `DietLlactis`,
`Host` or `Microbiome`; only microbial proteins are kept, matching the published
microbiota analysis.

```{r read}
raw <- read.delim(file.path(HERE, "ExtendedDataTable1CompleteMetaproteome.txt"),
                  skip = 2, quote = "\"", check.names = FALSE,
                  colClasses = "character", comment.char = "")
names(raw)[1:5] <- c("code", "protein_id", "aa_length", "description", "sequence")
raw <- raw[raw$code == "Microbiome", ]

counts <- vapply(raw[, -(1:5)], function(x) {
  blank <- !nzchar(trimws(x))                  # blank cell = protein not seen
  v <- suppressWarnings(as.numeric(x))
  bad <- is.na(v) & !blank                      # a parse failure that ISN'T blank is
  if (any(bad))                                 # unexpected data, not a documented zero
    stop("Non-numeric, non-blank count cell(s): ",
         paste(unique(x[bad]), collapse = ", "), call. = FALSE)
  v[blank] <- 0
  v
}, numeric(nrow(raw)))
stopifnot(all(counts >= 0), all(counts == round(counts)))  # PSM counts only
storage.mode(counts) <- "integer"
dimnames(counts) <- list(raw$protein_id, names(raw)[-(1:5)])

c(proteins = nrow(counts), samples = ncol(counts), total_psm = sum(counts))
```

# Design, from the sample names

Names are `{cage}_{mouse}_{diet}`, e.g. `C2_1_20RIC`. The diet token gives both the
protein source and its percentage, and — through the published feeding order — the
period, which is what `age_week` needs. `T0` is the pre-intervention baseline on
standard chow; `CTL` is the repeat feeding at study end, 20% soy for C1/C2 and 20%
casein for C3/C4.

The ninth diet, 20% chicken bone, caused weight loss, was stopped after three days and
yielded no samples, so no token exists for it.

```{r design}
diets <- data.frame(
  token  = c("T0", "20SOY", "20CAS", "20RIC", "40SOY", "20YST", "40CAS", "20PEA",
             "20EGG", "CTL"),
  period = 0:9,
  diet   = c(NA, "Soy", "Casein", "Rice", "Soy", "Yeast", "Casein", "Pea", "Egg", NA),
  dose   = c(NA, "20", "20", "20", "40", "20", "40", "20", "20", "20"),
  phase  = c("baseline", rep("treatment", 8), "control"))

f <- strsplit(colnames(counts), "_")
cage  <- vapply(f, `[`, "", 1)
token <- vapply(f, `[`, "", 3)
stopifnot(token %in% diets$token)

md <- data.frame(diets[match(token, diets$token), c("period", "phase", "diet", "dose")],
                 sample_id = colnames(counts), cage = cage,
                 subject = paste(cage, vapply(f, `[`, "", 2), sep = "_"),
                 # sex is not in the name and is not free to vary; it is also the
                 # paper's "mouse group", the sexes having come from different rooms
                 sex = ifelse(cage %in% c("C1", "C3"), "Male", "Female"),
                 row.names = colnames(counts))
# control re-feeds take their source from the cage
md$diet[md$phase == "control"] <- ifelse(cage[md$phase == "control"] %in% c("C1", "C2"),
                                         "Soy", "Casein")
# adonis2 on a continuous term is invariant to a constant shift, so weeks-since-T0
# stands in for age. The bone week still elapsed, so CTL sits two weeks after period 8.
md$age_week <- ifelse(md$period == 9, 10L, md$period)

table(md$period, md$diet, useNA = "ifany")
```

`C3_1` contributes 3 samples and `C4_2` 8; the other ten mice contribute 10 each.

# Bins

Protein ids encode the MAG bin (`AB1-1_Cage2_bin.36_00001` -> `AB1-1_Cage2_bin.36`), so
the bin is recoverable from the id alone. But a bin is per-assembly, not per-species: the
same organism gets a different bin number in every cage it was assembled in, so `Bin`
over-splits whatever it is used as a stand-in for species.

Extended Data Table 6 carries a real GTDB lineage per protein, curated rather than parsed
from an id, and does not have that problem -- the same species gets the same lineage
string regardless of which cage's assembly it came from. It has no counts of its own
(§ intro), so it is loaded here only for its five annotation columns, joined onto Table 1's
proteins by protein id.

```{r bins}
bin <- sub("_[0-9]+$", "", rownames(counts))

t6 <- read.delim(file.path(HERE, "annotated", "ExtendedDataTable6AnnotatedMicrobialMetaproteomes.txt"),
                 skip = 2, quote = "\"", check.names = FALSE, colClasses = "character",
                 comment.char = "", fileEncoding = "latin1")
lineage <- setNames(t6[["Taxa Lineage"]], t6[["Protein Identifier"]])[rownames(counts)]

ranks <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rk <- t(vapply(strsplit(lineage, ";"), function(x) {
  x <- sub("^[a-z]__", "", x)
  x[!nzchar(x)] <- NA
  length(x) <- length(ranks)
  x
}, character(length(ranks))))
dimnames(rk) <- list(rownames(counts), ranks)

tax <- cbind(Bin = bin, rk)
c(bins = length(unique(bin)), genera = length(unique(na.omit(tax[, "Genus"]))),
 unmatched = sum(is.na(lineage)), proteins = nrow(tax))
```

`r sum(is.na(lineage))` of `r nrow(tax)` proteins have no Table 6 entry and stay
unclassified at every rank; `Bin` is unaffected and still covers all of them.

## No phylogenetic tree {#no-tree}

`phy_tree` is the one phyloseq slot this dataset cannot fill honestly, and both routes
were measured before being abandoned.

Aligning the proteins (`DECIPHER::AlignSeqs` → `Treeline`, the 16S idiom) assumes the
features are homologous. They are not: the 200 most abundant proteins align to **4236
columns for a mean length of 589 — 86% gaps**, non-homologous sequences gapping past
each other. A tree from that is a similarity dendrogram, not a phylogeny.

A GTDB taxonomy cladogram over bins fails worse. Measured while Table 6's lineages were
still loaded: only **415 of 750 bins** carry one, so dropping the unplaceable bins costs a
**median 33% of each sample's mass**, and that loss is strongly **diet-associated** — mean
retention 31% on Yeast against 73% on Pea, ANOVA $p = 4\times10^{-24}$. Weighted UniFrac
on that tree would read a diet-confounded classification-coverage artifact as composition:
a larger confound than the sequencing depth rarefaction exists to remove. That measurement
is why the lineages are no longer carried at all.

So there is no weighted UniFrac here. This is the motivation for the method rather than
an obstacle to it — relatedness at protein resolution has essentially no tree-based
option, which is why the ground metric is a continuous representation space.

# The phyloseq object

```{r build}
ps <- phyloseq(otu_table(counts, taxa_are_rows = TRUE),
               sample_data(md),
               tax_table(tax),
               refseq(Biostrings::AAStringSet(setNames(raw$sequence, raw$protein_id))))
ps
```

```{r save}
saveRDS(ps, file.path(HERE, "ps.rds"))
```

# Checking against the published result

> "A Bray–Curtis dissimilarity matrix calculated from proteinaceous biomasses of species
> in the microbiota was analyzed, using the adonis2 function in the vegan package in R,
> to test for the effect of protein source, amount of protein, mouse group, or age (in
> weeks) on microbiota composition. The strata function was used to account for repeated
> measures."

Their `df` fixes the coding: protein source at df 7 means 8 levels, not the 6 diets, and
amount at df 2 means 3 — which only works if T0 is in the model with its own level of
each. So source is the six proteins plus `chow` and `control`, and all 111 samples are
used.

`tax_glom()` needs more than one rank to work with, so the proteins are summed onto
their real taxon directly — the finest lineage rank Table 6 resolved. A protein with no
Table 6 lineage falls back to its own `Bin` rather than a shared `Unclassified` bucket:
merging only happens where the lineage says two proteins are the same organism, so
proteins nobody could place don't get treated as each other's evidence. (A single pooled
`Unclassified` bucket was tried first and fits the published numbers visibly worse —
28% of proteins have no lineage, and merging them regardless of their real, probably
different, organisms adds noise a per-cage `Bin` fallback avoids.)

```{r species}
tx  <- as(tax_table(ps), "matrix")
lin <- tx[, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")]
taxon <- apply(lin, 1, function(x) paste(x[!is.na(x)], collapse = ";"))
taxon <- ifelse(nzchar(taxon), taxon, tx[, "Bin"])

taxa_ab <- rowsum(as(otu_table(ps), "matrix"), taxon)
n_taxa  <- nrow(taxa_ab)
taxa_ab <- t(taxa_ab) / colSums(taxa_ab)          # samples in rows, relative abundance

m <- data.frame(sample_data(ps))
m$source <- ifelse(m$phase == "baseline", "chow",
            ifelse(m$phase == "control", "control", m$diet))
m$amount <- ifelse(is.na(m$dose), "chow", m$dose)

set.seed(1)
a <- adonis2(vegdist(taxa_ab, "bray") ~ source + amount + sex + age_week,
             data = m, by = "terms",
             permutations = how(blocks = factor(m$subject), nperm = 9999))
a
```

```{r compare, echo = FALSE}
data.frame(term = c("source", "amount", "sex", "age_week"),
           paper_df = c(7, 2, 1, 1), paper_R2 = c(.39, .01, .21, .01),
           paper_F = c(15.0, 2.2, 56.9, 2.5),
           df = a$Df[1:4], R2 = round(a$R2[1:4], 3), F = round(a$F[1:4], 1))
```

Source, mouse group and age land close to the published values, against `r n_taxa`
taxa — real lineage where Table 6 resolved one, `Bin` otherwise, so still finer than
their 161 dereplicated species (per-cage assembly still splits organisms Table 6 couldn't
place, just fewer of them than before).

The one term that does not match is `amount`, at df 1 against their 2: coding T0 as its
own level of *both* source and amount makes the two collinear, and the sequential fit
gives the shared degree of freedom to whichever enters first. The variance explained is
unchanged, so this is a covariate coding detail — but their exact coding is not
recoverable from the published methods, and this is where that shows.

"Mouse group" is sex. The sexes came from different Jackson Labs rooms and started with
different microbiomes, so group, room, cage pair and sex are one factor and cannot be
separated. Under `blocks = subject` it is constant within a mouse and has no valid
permutation distribution — read its $R^2$, not its $p$.

**The object reproduces the paper.** Analysis continues in `case_study.Rmd`.
