---
title: "MCL parameters"
subtitle: "How the graph thresholds and inflation shape the bins"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 3.6,
    message = FALSE, warning = FALSE
)
```

`protein_similarity()` and `bin_proteins()` have three parameters: the
similarity threshold `min_sim`, the coverage threshold `min_coverage` and the
MCL `inflation`. This page scores a grid of the three on the universe of
experimentally characterised Swiss-Prot enzymes from
[Simulation 2](simulation_ec.html). MCL bins are meant as abundance units
that group related proteins, so no single score decides a setting: each one
trades fragmentation (singleton proteins) against EC3 precision and EC3
recall. EC4 F1 is kept as a stricter, exact-enzyme diagnostic.

```{r libs}
library(repdist)
library(ggplot2)
```

```{r data}
u <- readRDS(file.path("curated", "simulation_universe.rds"))
proteins <- u$proteins
ec4 <- setNames(u$annotation$ec4, u$annotation$protein)
ec3 <- setNames(u$annotation$ec3, u$annotation$protein)
length(proteins)
```

# Grid

```{r grid}
grid <- expand.grid(min_sim = c(0.6, 0.7, 0.8, 0.9),
    min_coverage = c(0, 0.5, 0.8), inflation = c(1.5, 2, 3, 4))
nrow(grid)
```

The two thresholds act independently on each pair, so one similarity matrix
per coverage threshold, computed at the lowest `min_sim`, gives every other
`min_sim` by zeroing the pairs below it. The check below confirms this for
the package defaults.

```{r similarity}
n_cores <- BiocParallel::multicoreWorkers()
base <- lapply(c(`0` = 0, `0.5` = 0.5, `0.8` = 0.8), function(cv) {
    protein_similarity(proteins, min_sim = 0.6, min_coverage = cv,
        n_cores = n_cores)
})
gate <- function(S, t) {
    S[S < t] <- 0
    diag(S) <- 1
    S
}
stopifnot(isTRUE(all.equal(gate(base[["0.5"]], 0.7),
    protein_similarity(proteins, min_sim = 0.7, min_coverage = 0.5,
        n_cores = n_cores))))
```

# Scores

Each setting is scored on protein pairs. **Precision** is the share of pairs
placed in one bin that carry the same label; **recall** is the share of
same-label pairs placed in one bin; F1 is their harmonic mean (all three from
`clusterCrit::extCriteria()`, F1 being its Czekanowski-Dice index). Pair counts grow with the square of bin size, so large
mixed bins weigh heavily on precision.

```{r scores}
pair_scores <- function(labels, membership) {
    unlist(clusterCrit::extCriteria(as.integer(factor(labels)),
        as.integer(factor(membership[names(labels)])),
        c("Precision", "Recall", "Czekanowski_Dice")))
}
results <- do.call(rbind, lapply(seq_len(nrow(grid)), function(k) {
    g <- grid[k, ]
    bins <- bin_proteins(gate(base[[as.character(g$min_coverage)]],
        g$min_sim), inflation = g$inflation)
    sizes <- lengths(bins$clusters)
    s3 <- pair_scores(ec3, bins$membership)
    s4 <- pair_scores(ec4, bins$membership)
    data.frame(g, bins = length(sizes),
        `singletons (%)` = 100 * sum(sizes == 1) / length(proteins),
        `largest bin` = max(sizes),
        `EC3 precision` = s3[["precision"]], `EC3 recall` = s3[["recall"]],
        `EC4 precision` = s4[["precision"]], `EC4 recall` = s4[["recall"]],
        `EC4 F1` = s4[["czekanowski_dice"]], check.names = FALSE)
}))
```

The three objectives against `min_sim`, by coverage threshold (columns) and
inflation (colour); the dotted line marks the default `min_sim`:

```{r tradeoff-plot, fig.height = 6.5, fig.alt = "EC3 precision, EC3 recall and singleton proteins against the similarity threshold, by coverage threshold and inflation"}
objectives <- c("EC3 precision", "EC3 recall", "singletons (%)")
long <- do.call(rbind, lapply(objectives, function(m) {
    data.frame(results[c("min_sim", "min_coverage", "inflation")],
        objective = factor(m, objectives), value = results[[m]])
}))
ggplot(long, aes(min_sim, value, colour = factor(inflation))) +
    geom_vline(xintercept = 0.7, linetype = 3, colour = "grey50") +
    geom_line() +
    geom_point(size = 1.5) +
    facet_grid(objective ~ min_coverage, scales = "free_y",
        labeller = labeller(min_coverage = label_both)) +
    labs(x = "min_sim (predicted TM-score)", y = NULL,
        colour = "inflation") +
    theme_bw()
```

```{r table}
knitr::kable(results[order(results$min_coverage, results$min_sim,
    results$inflation), ], digits = 3, row.names = FALSE)
```

# Reading

```{r reading-values, include = FALSE}
def <- results[results$min_sim == 0.7 & results$min_coverage == 0.5 &
    results$inflation == 2, ]
nocov <- results[results$min_sim == 0.7 & results$min_coverage == 0 &
    results$inflation == 2, ]
covered <- results[results$min_coverage > 0, ]
```

- **The coverage gate matters most.** Without it, low similarity thresholds
  chain proteins that share a single domain into a few very large bins (the
  largest holds `r nocov[["largest bin"]]` proteins at `min_sim` 0.7 and
  inflation 2) and EC3 precision drops to
  `r round(nocov[["EC3 precision"]], 2)`, against
  `r round(def[["EC3 precision"]], 2)` with `min_coverage` 0.5.
- **With the coverage gate, the settings trade smoothly.** Raising `min_sim`
  or inflation gives more singletons and higher EC3 precision but lower EC3
  recall. Across those settings EC3 precision runs from
  `r round(min(covered[["EC3 precision"]]), 2)` to
  `r round(max(covered[["EC3 precision"]]), 2)`, EC3 recall from
  `r round(min(covered[["EC3 recall"]]), 3)` to
  `r round(max(covered[["EC3 recall"]]), 3)`, and singleton proteins from
  `r round(min(covered[["singletons (%)"]]))`% to
  `r round(max(covered[["singletons (%)"]]))`%.
- **The defaults are a compromise, not an optimum.** `min_sim` 0.7,
  `min_coverage` 0.5 and inflation 2 give EC3 precision
  `r round(def[["EC3 precision"]], 2)`, EC3 recall
  `r round(def[["EC3 recall"]], 3)` and
  `r round(def[["singletons (%)"]])`% singleton proteins: inside the stable
  region, neither the most precise nor the least fragmented setting. This
  page reports the trade-off; it does not change them.
- **EC4 is a stricter diagnostic.** EC4 F1 stays between
  `r round(min(covered[["EC4 F1"]]), 2)` and
  `r round(max(covered[["EC4 F1"]]), 2)` across the gated settings (defaults
  `r round(def[["EC4 F1"]], 2)`), with EC4 precision low at every setting: a
  bin often holds one structural family whose members catalyse related but
  different reactions. No setting turns the bins into EC4 classes; the
  thresholds choose the granularity of the grouping.

```{r session}
sessionInfo()
```
