---
title: "Reference"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

This page explains what each `repdist` function computes, in the order of a
typical analysis. Full argument lists are in the package help
(`?sample_mmd`, `?protein_similarity`, ...).

`embed_proteins()` returns an `AAStringSet` containing the protein sequences
and one learned representation per protein. In the examples below this object
is called `proteins` (in the help pages it is the argument `x`); the
embedding matrix is `S4Vectors::mcols(proteins)$embedding`, and $z_i$ is its
row for protein $i$.

| Name | Holds |
|---|---|
| `proteins` | sequences and their representations (`AAStringSet`) |
| `counts` | samples × proteins counts |
| `D` | distance between samples (`dist`) |
| `S` | protein × protein similarity graph (matrix) |
| `bins` | MCL result of `bin_proteins()` |
| `bin_counts` | samples × bins counts |

| Step | Function | Input | Output |
|---|---|---|---|
| Representations | `embed_proteins()`, `read_embeddings()` | sequences, or a TM-Vec database | `proteins` |
| Sample distance | `sample_mmd()` | `counts`, `proteins` | `D` |
| Similarity graph | `protein_similarity()` | `proteins` | `S` |
| Protein bins | `bin_proteins()` | `S` | `bins` |
| Aggregation | `bin_glom()` | `counts`, `bins` | `bin_counts` |
| QC | `plot_tsne()`, `plot_bin_similarity()` | `proteins`, `bins`, `S` | ggplot objects |

```{r libs}
library(repdist)
library(ggplot2)
```

The executable examples use the three largest enzyme classes (EC3) of the
[Simulation 2](simulation_ec.html) universe: experimentally characterised
Swiss-Prot enzymes whose embeddings were computed with `embed_proteins()`.

```{r example-data}
u <- readRDS(file.path("curated", "simulation_universe.rds"))
classes <- names(sort(table(u$annotation$ec3), decreasing = TRUE))[1:3]
keep <- u$annotation$ec3 %in% classes
proteins <- u$proteins[keep]
ec3 <- setNames(u$annotation$ec3[keep], u$annotation$protein[keep])
table(ec3)
```

# Protein representations {#protein-representations}

Every other function takes the same object: an `AAStringSet` of protein
sequences, named by accession, with one embedding row per protein in
`mcols(proteins)$embedding`. Rows are unit-normalised.

```{r proteins}
proteins
dim(S4Vectors::mcols(proteins)$embedding)
```

There are two ways to get it.

**`embed_proteins()`: sequences to embeddings inside R.** It accepts a named
character vector, an `AAStringSet` or a FASTA path, and a `model` key from
`known_models()`. Models run in a Python environment that `repdist` builds
and pins itself through `basilisk`, so users do not install or configure
Python. Weights are downloaded on first use and pinned to fixed revisions
(HuggingFace commits, or a SHA-256-checked file). Every sequence is embedded
whole; nothing is truncated or dropped.

```{r embed, eval = FALSE}
proteins <- embed_proteins("proteins.fasta") # model = "tmvec"
proteins <- embed_proteins(seqs, model = "esm2-650m", device = "cpu")
```

**`read_embeddings()`: embeddings computed elsewhere.** TM-Vec users can embed
outside R with the `tmvec` command-line tool
(<https://github.com/valentynbez/tmvec>) and import its `build-db` database
(`.npz`), giving the FASTA it was built from. An HDF5 file with one group per
protein (datasets `seq` and `emb`) is also read; per-residue embeddings are
averaged.

```{r read, eval = FALSE}
proteins <- read_embeddings("db/proteins.npz", seqs = "proteins.fasta")
```

**`known_models()`** lists the registered models. The default, `tmvec`, is a
TM-Vec head [@hamamsy2024tmvec] on the ProtT5-XL encoder
[@elnaggar2022prottrans]. Its head is trained so that the cosine similarity of
two embeddings predicts the TM-score of the two structures
[@zhang2004tmscore]; on this site, "predicted TM-score" means that cosine.
The `esm2-*` models give general sequence representations without this
calibration.

```{r models}
m <- known_models()
data.frame(
    model = names(m),
    backbone = vapply(m, `[[`, "", "backbone"),
    trained_to = vapply(m, function(s) s$training_max_length, 0),
    row.names = NULL
)
```

`trained_to` is the longest sequence in the model's training data. It is
recorded for reference and not enforced.

# Sample-level distance: MMD {#sample-level-distance-mmd}

`sample_mmd(counts, proteins)` treats sample $P$ as a weighted set of protein
embeddings $\{(z_i, p_i)\}$, where $p_i = c_i / \sum_j c_j$ is protein $i$'s
share of the sample's counts. Two samples are compared by the maximum mean
discrepancy [@gretton2012mmd; @muandet2017kme]:

$$
\mathrm{MMD}^2(P,Q) =
\sum_{i,j} p_i p_j\, k(z_i, z_j)
- 2 \sum_{i,j} p_i q_j\, k(z_i, z_j)
+ \sum_{i,j} q_i q_j\, k(z_i, z_j),
$$

with the Gaussian (RBF) kernel on Euclidean distance

$$
k(z_i, z_j) = \exp\!\left(-\frac{\lVert z_i - z_j \rVert^2}{2\sigma^2}\right).
$$

The sums run over all proteins observed in either sample, and the function
returns $\mathrm{MMD}(P,Q)$ (the square root) as a `dist`. For unit-norm
embeddings $\lVert z_i - z_j \rVert^2 = 2(1 - s_{ij})$, where $s_{ij}$ is the
cosine similarity, so $k(z_i, z_j) = \exp(-(1 - s_{ij})/\sigma^2)$; with
TM-Vec, $s_{ij}$ is the predicted TM-score.

**Why a kernel distance.** If $k$ were 1 for identical proteins and 0
otherwise, the expression would reduce to the Euclidean distance between the
two relative-abundance vectors, an identity-only comparison. With the RBF
kernel, the cross term lets different accessions with close embeddings
partly match, in proportion to their abundances. MMD compares the two
weighted distributions through their kernel mean embeddings; unlike
optimal-transport distances it solves no transport problem and is computed
directly from the protein kernel matrix. The choice is methodological; it is
not a claim that transport-based distances cannot be used.

In the example, samples `s1` and `s2` share no accession but draw on the
same enzyme class, while `s3` draws on another class:

```{r mmd-toy}
pick <- function(class, i) names(ec3)[ec3 == class][i]
counts <- matrix(0, 3, length(proteins),
    dimnames = list(c("s1", "s2", "s3"), names(proteins)))
counts["s1", pick(classes[1], 1:10)] <- 10
counts["s2", pick(classes[1], 11:20)] <- 10
counts["s3", pick(classes[2], 1:10)] <- 10

D <- sample_mmd(counts, proteins)
D
vegan::vegdist(counts, "bray") # 1 for every pair: no shared accession
```

**Bandwidth.** With `sigma = NULL` (default), $\sigma$ is the median pairwise
Euclidean distance among the proteins observed in any sample (the median
heuristic). It therefore changes when samples bring in new proteins. The value
used is returned as `attr(D, "sigma")`; pass it back to hold the kernel fixed
across analyses. For TM-Vec embeddings, `sigma = sqrt(1 - t)` places the
kernel's $1/e$ point at predicted TM-score $t$.

```{r sigma}
attr(D, "sigma")
D2 <- sample_mmd(counts[1:2, ], proteins, sigma = attr(D, "sigma"))
```

`weighted = FALSE` gives every detected protein equal weight (a
presence/absence version). `sample_mmd()` holds the protein × protein kernel
in memory, so memory grows with the square of the number of proteins.

# Protein similarity graph {#protein-similarity-graph}

```
protein embeddings
        │  cosine similarity
pairwise similarity s_ij
        │  similarity and coverage thresholds
weighted sparse graph W
        │  Markov clustering
protein bins ── bin_glom() ──► samples × bins counts
```

## `protein_similarity()` {#protein-similarity}

`protein_similarity(proteins, min_sim = 0.7, min_coverage = 0.5)` returns the
weighted adjacency matrix

$$
W_{ij} =
\begin{cases}
s_{ij}, & s_{ij} \ge t_{\mathrm{sim}} \text{ and } \mathrm{cov}_{ij} \ge t_{\mathrm{cov}},\\
0, & \text{otherwise,}
\end{cases}
$$

where $s_{ij}$ is the cosine similarity of the embeddings,
$t_{\mathrm{sim}}$ = `min_sim` and $t_{\mathrm{cov}}$ = `min_coverage`. With
TM-Vec this reads $W_{ij} = \mathrm{TM}_{ij}$ when the predicted TM-score
$\mathrm{TM}_{ij} \ge t_{\mathrm{TM}}$ and the coverage passes.

Coverage is computed only for pairs that pass `min_sim`: the two sequences
are aligned locally (Smith-Waterman [@smith1981local]; BLOSUM62, gap opening
11, extension 1), and $\mathrm{cov}_{ij}$ is the smaller of the two
fractions of each sequence that the alignment spans. A pair that shares one domain but differs
over the rest of two long proteins can still have a high predicted TM-score;
the coverage threshold keeps such a pair from becoming an edge. Coverage is a
hard gate: it removes edges but never rescales their weights.
`min_coverage = 0` skips the alignments; `n_cores` runs them in parallel.

The result is an ordinary symmetric matrix (1 on the diagonal, 0 for every
non-edge), so it can be clustered by other graph algorithms as well;
`bin_proteins()` is one choice, not a requirement.

```{r similarity}
S <- protein_similarity(proteins)
c(proteins = nrow(S), edges = sum(S[upper.tri(S)] > 0))
```

# MCL clustering and aggregation {#mcl-clustering-and-aggregation}

## `bin_proteins()` {#bin-proteins}

`bin_proteins(S, inflation = 2)` partitions the graph by Markov clustering
(MCL) [@vandongen2000mcl]. MCL simulates random walks on the weighted graph,
alternating expansion (matrix powers) with inflation (elementwise powers
followed by renormalisation), so that flow concentrates within densely
connected regions. Clustering proteins by MCL on a pairwise-similarity network
goes back to TRIBE-MCL [@enright2002tribemcl] and OrthoMCL
[@li2003orthomcl]. Here, each connected component is clustered separately,
each protein's self-loop takes the weight of its heaviest edge (as in the
`mcl` program), and proteins without edges become singleton bins. Higher
inflation gives more, smaller bins.

```{r bins}
bins <- bin_proteins(S)
bins$graph
head(lengths(bins$clusters))
```

The result lists the members of each bin (`clusters`), each protein's bin
(`membership`), a medoid per bin (`representatives`) and a summary of the
graph (`graph`).

**Thresholds define edges, not bins.** If A–B and B–C are edges, MCL can put
A, B and C in one bin even when A–C is below `min_sim`. A bin therefore does
not guarantee that every pair of its members passes the thresholds. Three
toy proteins with embeddings at 0°, 40° and 80° on a circle, so neighbours
have cosine 0.77 and the two ends 0.17:

```{r chain}
toy <- Biostrings::AAStringSet(c(A = "MKVLA", B = "MKVLG", C = "MKVLS"))
angle <- c(0, 40, 80) * pi / 180
S4Vectors::mcols(toy)$embedding <- cbind(cos(angle), sin(angle))
S_toy <- protein_similarity(toy, min_coverage = 0)
round(S_toy, 2) # A-C: cosine 0.17, below min_sim, so no edge
bin_proteins(S_toy)$clusters
```

## `bin_glom()` {#bin-glom}

`bin_glom(counts, bins)` sums the counts of each bin's members:

```
counts (samples × proteins) ──► membership ──► counts (samples × bins)
```

Sample totals are unchanged and a sparse `Matrix` stays sparse. Give it raw
counts when the next step is a count model such as DESeq2.

```{r glom}
bin_counts <- bin_glom(counts, bins)
bin_counts[, 1:5]
```

# QC and visualization {#qc-and-visualization}

## `plot_tsne()` {#plot-tsne}

`plot_tsne(proteins, color = NULL)` draws a two-dimensional t-SNE
[@vandermaaten2008tsne] of the embeddings computed on cosine distance,
$1 - s_{ij}$ (for TM-Vec, one minus the predicted TM-score). `color` takes
labels named by protein; the 12 most frequent are coloured. t-SNE preserves
local neighbourhoods, not distances between clusters or their sizes, and the
layout depends on the random seed.

```{r tsne, fig.height = 4.2}
set.seed(1)
plot_tsne(proteins, color = ec3)
```

## `plot_bin_similarity()` {#plot-bin-similarity}

`plot_bin_similarity(bins, bin, S, proteins)` shows every pairwise similarity
among one bin's members. Cell colour is the raw similarity from the
embeddings, before any threshold; a cross marks each pair that is not an edge
of `S`, that is, did not pass both thresholds. Crossed pairs are in the bin
through shared neighbours, or passed `min_sim` but failed the coverage gate.

```{r bin-similarity, fig.width = 6.5, fig.height = 5.5}
edge_share <- vapply(bins$clusters, function(m) {
    E <- S[m, m]
    mean(E[upper.tri(E)] > 0)
}, numeric(1))
sizes <- lengths(bins$clusters)
shown <- names(which.min(edge_share[sizes >= 5 & sizes <= 15]))
plot_bin_similarity(bins, shown, S, proteins)
```

The bin shown is the one between 5 and 15 members with the smallest share of
pairs that are edges.

# References {-}
