---
title: "Simulation: shared function without shared accessions"
author: "Yixuan Yang"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 10,
    fig.height = 3.4, message = FALSE, warning = FALSE
)
```

# The question

When different proteins contribute the same function, can a beta-diversity
method tell two groups apart? Bray-Curtis compares samples through accession
identity, so a functional shift spread across *non-overlapping* homologs is
invisible to it by construction. This page builds exactly that case and
runs three distances over it:

1. **RBF-MMD** on TM-Vec embeddings (`sample_mmd()`).
2. **Bray-Curtis on MCL bins**, from a graph joining proteins with a predicted
    TM-score of at least 0.7.
3. **Bray-Curtis at the protein level** — the conventional baseline.

The first two read protein relatedness, each through a different lens; the
third reads identity only.

```{r libs}
library(repdist)
library(ggplot2)
```

# The protein universe

Ten CATH FunFams — functional families of structural domains — with 150
members each, embedded with TM-Vec. The membership table and embeddings are
`validation/funfam_universe.rds` on this branch; `validation/README.md`
describes their source and `validation/prepare_funfam_universe.R` their
reconstruction. These coordinates come from the older `scikit-bio/tmvec` head, not the
current default, and their model revisions were not recorded: they are a fixed
example, not a reference for new inference.

```{r universe}
u <- readRDS(file.path("validation", "funfam_universe.rds"))
members <- u$members
x <- u$proteins
ids <- members$protein
table(members$label)
```

# The design

Twenty samples, ten per group, and **every sample contains every protein**, so
presence/absence carries no information and only abundance can be read. Each
sample has ten proteins upregulated twenty-fold over lognormal baseline
abundances:

- **Group A** draws its ten at random across all ten families — diffuse noise
    with no functional theme.
- **Group B** draws its ten from a single family (SHMT) — a coherent
    functional shift.

Within each group the upregulated sets are **disjoint**, and group A never
draws a protein group B uses, so *no two samples share a single upregulated
accession*. Group B's samples have nothing in common at the accession level
and everything in common at the family level — the case an identity-based
distance cannot see.

```{r design}
n_per_group <- 10L
n_up <- 10L
fold <- 20
depth <- 100000L
set.seed(20260901)

group <- factor(rep(c("A", "B"), each = n_per_group))
b_family <- members$family[members$label == "SHMT"][1]

# Disjoint, not merely independent: draw once, then cut into blocks.
draw_blocks <- function(pool) {
    unname(split(
        sample(pool, n_per_group * n_up), rep(seq_len(n_per_group), each = n_up)
    ))
}
up_B <- draw_blocks(ids[members$family == b_family])
up_A <- draw_blocks(setdiff(ids, unlist(up_B)))
up <- c(up_A, up_B)

w <- matrix(rlnorm(length(group) * length(ids), sdlog = 0.7),
    nrow = length(group),
    dimnames = list(sprintf("s%02d", seq_along(group)), ids)
)
for (i in seq_along(group)) w[i, up[[i]]] <- w[i, up[[i]]] * fold
counts <- t(apply(w, 1, function(x) rmultinom(1, depth, x)[, 1]))

c(
    shared_upregulated_accessions = sum(duplicated(unlist(up))),
    observed_in_every_sample = mean(colSums(counts > 0) == length(group))
)
```

# Three distances

One similarity matrix feeds the bins: `protein_similarity()` keeps every pair
of proteins with a predicted TM-score of at least 0.7, and an alignment
covering half of both sequences, as an edge, and `bin_proteins()` splits the
graph by Markov clustering (MCL) at inflation 2. The 0.7 is an edge floor, not
an all-pairs guarantee: a bin can hold pairs below it, joined through shared
neighbours. MMD builds its euclidean ground metric from the same embeddings
internally.

```{r structures}
S <- protein_similarity(x, min_sim = 0.7)
b <- bin_proteins(S, inflation = 2)
binned <- bin_glom(counts, b)
sizes <- lengths(b$clusters)
c(b$graph, bins = length(sizes), largest_bin = max(sizes),
    singleton_fraction = sum(sizes == 1) / sum(sizes))
```

The graph leaves no protein isolated and joins some families into shared
components; MCL splits those again, and the ten families come back as
`r ncol(binned)` bins without ever being told there were ten.

## What RBF-MMD computes

Each sample is a weighted point cloud in embedding space: protein $a$ enters
with its relative abundance $w_a$. For samples $p$ and $q$,

$$\mathrm{MMD}^2(p,q)\;=\;\sum_{a,b} w^p_a w^p_b\,k(a,b)
\;-\;2\sum_{a,b} w^p_a w^q_b\,k(a,b)
\;+\;\sum_{a,b} w^q_a w^q_b\,k(a,b),$$

under a Gaussian kernel on the euclidean ground metric,
$k(a,b)=\exp\!\left(-\lVert z_a-z_b\rVert^2/2\sigma^2\right)$, which on the
unit-norm rows `embed_proteins()` returns is exactly
$\exp\!\left(-(1-\mathrm{TM}_{ab})/\sigma^2\right)$.

The middle term is the whole difference from Bray-Curtis. Two *different*
proteins with $k(a,b)\approx 1$ cancel there almost as completely as the same
accession would. Shrink $k$ to the indicator of $a=b$ and the expression
collapses to $\lVert w^p - w^q\rVert^2$ — an identity-only distance, blind to
any relationship between distinct proteins. That is the corner this simulation
puts the protein-level baseline in by construction.

`sigma` defaults to the median pairwise distance among the retained proteins,
the usual median heuristic; the value actually used comes back as
`attr(D, "sigma")`, and passing it back holds the bandwidth fixed across runs
or studies.

```{r distances}
D <- list(
    `RBF-MMD` = sample_mmd(counts, x),
    `Bray-Curtis (MCL bins)` = vegan::vegdist(binned, "bray"),
    `Bray-Curtis (protein)` = vegan::vegdist(counts, "bray")
)
```

# PERMANOVA

```{r permanova}
md <- data.frame(group = group)
perma <- do.call(rbind, lapply(names(D), function(n) {
    set.seed(1)
    a <- vegan::adonis2(D[[n]] ~ group, data = md, permutations = 999)
    data.frame(
        distance = n, R2 = round(a$R2[1], 3),
        F = round(a[["F"]][1], 2), p = a$`Pr(>F)`[1]
    )
}))
knitr::kable(perma, row.names = FALSE)
```

Both distances that can read protein relatedness — through the kernel or the
bins — find the group difference decisively. Protein-level Bray-Curtis finds
nothing:
each sample's upregulated set is private, so to an identity-based comparison
every sample is equally unlike every other, in both groups.

# Ordination

PCoA on each distance, in the order above. The percentages in each panel
heading are that axis's share of the total variation, computed per distance
because each has its own eigenvalues.

```{r pcoa}
#| fig.width: 10
#| fig.height: 4
#| fig.alt: >
#|   PCoA of the simulated samples under RBF-MMD, MCL-bin Bray-Curtis,
#|   and protein-level Bray-Curtis, coloured by group
ord <- do.call(rbind, lapply(names(D), function(n) {
    e <- cmdscale(D[[n]], k = 2, eig = TRUE)
    rel <- 100 * e$eig / sum(e$eig[e$eig > 0]) # negative eigenvalues excluded
    data.frame(
        Axis1 = e$points[, 1], Axis2 = e$points[, 2], group = group,
        panel = sprintf("%s\n(PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2])
    )
}))
ord$panel <- factor(ord$panel, unique(ord$panel))

ggplot(ord, aes(Axis1, Axis2, colour = group)) +
    geom_point(size = 2, alpha = 0.85) +
    stat_ellipse(type = "t", linewidth = 0.3) +
    facet_wrap(~panel, nrow = 1, scales = "free") +
    labs(x = "PCo1", y = "PCo2", colour = "group") +
    theme_bw() +
    theme(legend.position = "bottom")
```

In the first two panels group B collapses into a tight cluster away from
group A, and the first axis — about half of the total variation — carries the
separation. In the protein-level Bray-Curtis panel the two groups
sit interleaved, and the leading axes hold a few percent each: the distance is
near-constant across all pairs, within group and between.

# Reading

The two structure-aware routes are not interchangeable in general — MMD
weighs graded similarity through a kernel, the bins through a graph floor and
MCL — but on a coherent functional shift they agree, and the conventional
baseline is blind to it. When homologs rather than shared accessions carry a
signal, the choice of representation is not a refinement; it is the difference
between detecting the effect and missing it entirely.

```{r session}
sessionInfo()
```
