---
title: "Simulation 3: MMD, optimal transport and UniFrac"
subtitle: "Embedding- and graph-based sample distances on one FunFam contrast"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

Group A's ten samples each enrich a different domain of one CATH FunFam;
group B's each enrich a different domain of a second FunFam. Four sample
distances that use the protein embeddings or the MCL graph are compared on
the same counts.

```{r libs}
library(repdist)
library(ggplot2)
library(vegan)
n_cores <- BiocParallel::multicoreWorkers()
```

# Simulation

The FunFam domain universe and the negative-binomial count model of
[Simulation 1](simulation_funfam.html): every domain is present at a
log-normal baseline in every sample, and each sample puts $\rho$ = 2% of its
expected abundance on its enriched domain. The $k$-th enriched domain of
group B takes the baseline abundance of the $k$-th of group A. The two
FunFams are drawn with a fixed seed from those with at least ten domains.

```{r simulate}
ff <- readRDS(file.path("curated", "simulation_funfam.rds"))
ann <- ff$annotation
stopifnot(identical(ann$domain, names(ff$proteins)))
nb_counts <- function(up, ids, rho, pairs, depth = 2e4) {
    n <- length(up)
    pi <- setNames(rlnorm(length(ids)), ids)
    pi[pairs[, 2]] <- pi[pairs[, 1]] # matched baseline of enriched domains
    pi <- pi / sum(pi)
    s <- rlnorm(n, 0, 0.25)
    e <- rlnorm(length(ids), 0, 0.5)
    p <- matrix((1 - rho) * pi, n, length(ids), byrow = TRUE,
        dimnames = list(names(up), ids))
    for (j in seq_len(n)) {
        p[j, up[[j]]] <- p[j, up[[j]]] + rho / length(up[[j]])
    }
    mu <- s * depth * p
    phi <- (0.1 + 1 / mu) * rep(e, each = n)
    matrix(rnbinom(length(mu), mu = mu, size = 1 / phi), n,
        dimnames = dimnames(p))
}
n_per_group <- 10L
size <- table(ann$family)
set.seed(20261009)
fams <- sample(names(size)[size >= n_per_group], 2)
ff$families[match(fams, ff$families$family), c("family", "name")]
pairs <- vapply(fams, function(f) {
    sample(ann$domain[ann$family == f], n_per_group)
}, character(n_per_group))
up <- as.list(c(pairs))
names(up) <- sprintf("s%02d", seq_along(up))
counts <- nb_counts(up, ann$domain, 0.02, pairs)
group <- factor(rep(c("A", "B"), each = n_per_group))
```

# Distances

On unit-norm TM-Vec embeddings the Euclidean distance is
$\sqrt{2(1 - \widehat{TM})}$, a metric in the predicted TM-score.

- **MMD**: `sample_mmd()`, the maximum mean discrepancy under a Gaussian
  kernel on the embeddings [@gretton2012mmd].
- **OT (embedding)**: optimal transport, the exact 1-Wasserstein distance
  between the samples' relative abundances, with the Euclidean distance
  between embeddings as the cost [@peyre2019cot], solved by
  `transport::wasserstein()` [@schuhmacher2024transport]. Under a metric
  cost only the mass that differs needs to move, so each pair is solved on
  $(p - q)^+$ against $(p - q)^-$.
- **OT (MCL graph)**: the same, with the cost read from the weighted graph
  that MCL clusters (`protein_similarity()`, package defaults). An edge
  costs $1 - \widehat{TM}$, two domains cost their shortest path, and
  domains with no path, or a path longer than 1, cost 1. With no edges every
  pair costs 1 and this is Bray-Curtis on relative abundances.
- **Weighted UniFrac (MCL tree)**: the MCL graph is compressed into a tree.
  Each MCL bin (package defaults) becomes one tip, and the bins are joined
  by average linkage (UPGMA) on the mean embedding distance between their
  domains, i.e. UPGMA on the domains started from the bins. Weighted UniFrac
  [@lozupone2007unifrac] is computed on the bin counts with `GUniFrac`
  [@chen2012gunifrac]. It is the 1-Wasserstein distance under the tree's
  path metric [@evans2012kr], so it differs from OT only in the ground
  metric.

```{r distances}
S <- protein_similarity(ff$proteins, n_cores = n_cores)
bins <- bin_proteins(S)
Z <- S4Vectors::mcols(ff$proteins)$embedding
Z <- Z / sqrt(rowSums(Z^2))
rownames(Z) <- names(ff$proteins)
C <- as.matrix(dist(Z)) # sqrt(2 (1 - predicted TM))
g <- bins$membership[rownames(C)]
sums <- rowsum(t(rowsum(C, g)), g) # bin x bin sums of domain distances
n <- as.vector(table(g)[rownames(sums)])
tree <- ape::as.phylo(hclust(as.dist(sums / outer(n, n)), "average",
    members = n))
w1 <- function(a, b, C) {
    d <- a - b
    P <- d > 0
    N <- d < 0
    m <- sum(d[P])
    m * transport::wasserstein(d[P] / m, -d[N] / sum(-d[N]),
        costm = C[P, N, drop = FALSE], p = 1)
}
edges <- which(upper.tri(S) & S > 0, arr.ind = TRUE)
graph <- igraph::graph_from_data_frame(data.frame(rownames(S)[edges[, 1]],
    rownames(S)[edges[, 2]]), directed = FALSE, vertices = rownames(S))
C_graph <- pmin(igraph::distances(graph, weights = 1 - S[edges]), 1)
p <- counts / rowSums(counts)
ot <- function(C) {
    ij <- combn(nrow(p), 2)
    OT <- matrix(0, nrow(p), nrow(p), dimnames = list(rownames(p),
        rownames(p)))
    OT[t(ij[2:1, ])] <- apply(ij, 2, function(k) w1(p[k[1], ], p[k[2], ], C))
    as.dist(OT)
}
D <- list(
    MMD = sample_mmd(counts, ff$proteins),
    `OT (embedding)` = ot(C),
    `OT (MCL graph)` = ot(C_graph),
    `weighted UniFrac (MCL tree)` = as.dist(GUniFrac::GUniFrac(
        bin_glom(counts, bins), tree, alpha = 1,
        verbose = FALSE)$unifracs[, , "d_1"])
)
```

```{r checks, include = FALSE}
# transporting (p - q)+- equals the full-support distance
stopifnot(identical(rownames(C_graph), colnames(p)),
    all.equal(as.matrix(D$`OT (embedding)`)[2, 1],
        transport::wasserstein(p[1, ], p[2, ], costm = C, p = 1)),
    all.equal(as.matrix(D$`OT (MCL graph)`)[2, 1],
        transport::wasserstein(p[1, ], p[2, ], costm = C_graph, p = 1)))
# weighted UniFrac = W1 under the tree's path metric / (2 x tree height)
pb <- bin_glom(counts, bins)
pb <- pb / rowSums(pb)
path <- ape::cophenetic.phylo(tree)[colnames(pb), colnames(pb)]
H <- max(ape::node.depth.edgelength(tree))
stopifnot(all.equal(as.matrix(D$`weighted UniFrac (MCL tree)`)[2, 1],
    w1(pb[1, ], pb[2, ], path) / (2 * H)))
```

# PCoA

Ellipses are 95% ellipses per group (multivariate t).

```{r pcoa, echo = FALSE, fig.width = 8.5, fig.height = 7.5, fig.alt = "PCoA of the simulated FunFam contrast under MMD, optimal transport on the embeddings and on the MCL graph, and weighted UniFrac on the MCL tree, coloured by group, with 95% ellipses"}
ord <- do.call(rbind, lapply(names(D), function(m) {
    e <- cmdscale(D[[m]], k = 2, eig = TRUE)
    rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
    data.frame(PCo1 = e$points[, 1], PCo2 = e$points[, 2], group = group,
        panel = sprintf("%s\nPCo1 %.0f%%, PCo2 %.0f%%", m, rel[1], rel[2]))
}))
ord$panel <- factor(ord$panel, unique(ord$panel))
ggplot(ord, aes(PCo1, PCo2, colour = group)) +
    geom_point(size = 2) +
    stat_ellipse(linewidth = 0.4) +
    facet_wrap(~panel, scales = "free", nrow = 2) +
    labs(colour = NULL) +
    theme_bw() +
    theme(legend.position = "bottom")
```

# PERMANOVA

`adonis2()`, 999 permutations [@anderson2001permanova; @oksanen2025vegan].

```{r permanova}
knitr::kable(t(vapply(D, function(d) {
    set.seed(1)
    a <- adonis2(d ~ group, data = data.frame(group = group),
        permutations = 999)
    c(R2 = a$R2[1], F = a$F[1], p = a$`Pr(>F)`[1])
}, numeric(3))), digits = 3)
```

```{r session}
sessionInfo()
```

# References {-}
