---
title: "Simulation 2: Sequence-disjoint functional contrasts"
subtitle: "Enzyme classes carried by different sequence clusters in every sample"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

**Question.** Can representation-aware analysis recover a condition-level
functional difference when the proteins carrying each function belong to
different sequence clusters in every sample?

Two coherent functional states are compared symmetrically: condition A is
enriched for one EC3, condition B for another, and within a condition no
50% MMseqs2 sequence cluster is enriched in two samples. An EC3 class spans
many sequence families and often several folds, so this is harder and
closer to metaproteomics than the FunFam positive control of
[Simulation 1](simulation_funfam.html). The page ends with how the MCL bins
of the same universe group proteins against the EC labels.

```{r libs}
library(repdist)
library(ggplot2)
library(vegan)
```

# Count model

Twenty samples, two conditions of ten, follow the negative-binomial model of
[Simulation 1](simulation_funfam.html#count-model), repeated here so this
page stands on its own. Every protein of the universe is present at baseline
in every sample; sample $i$ has expected composition

$$
p_i = (1 - \rho)\,\pi + \rho\, q_i ,
$$

where $\pi_j \propto \mathrm{LogNormal}(0, 1)$ is the common baseline and
$q_i$ spreads its mass evenly over sample $i$'s enriched proteins. The signal
strength $\rho$ is the expected share of abundance on the enriched proteins:
a defined mass, not a fold change. Counts have mean
$\mu_{ij} = s_i D p_{ij}$ with depth $D = 2 \times 10^4$ and size factor
$s_i \sim \mathrm{LogNormal}(0, 0.25)$, and variance $\mu + \phi\mu^2$, with
a DESeq2-like dispersion trend at each count's own mean,
$\phi_{ij} = (0.1 + 1/\mu_{ij})\, e_j$, $e_j \sim \mathrm{LogNormal}(0, 0.5)$.
The $k$-th enriched protein of the second group takes the baseline
abundance of the $k$-th of the first, so the groups also match on the
baseline of the proteins that carry the signal.

The representative signal strength is $\rho$ = 2%, fixed before any result:
in the source study of the [case study](case_study.html), glycoside hydrolases rose from under
1% to over 2.5% of the metaproteome on the yeast diet
[@blakeleyruiz2025diet].

```{r count-model}
depth <- 2e4
rho_example <- 0.02
nb_counts <- function(up, ids, rho, pairs) {
    n <- length(up)
    pi <- setNames(rlnorm(length(ids)), ids)
    pi[pairs[, 2]] <- pi[pairs[, 1]] # matched baseline of enriched proteins
    pi <- pi / sum(pi)
    s <- rlnorm(n, 0, 0.25)
    e <- rlnorm(length(ids), 0, 0.5)
    p <- matrix((1 - rho) * pi, n, length(ids), byrow = TRUE,
        dimnames = list(names(up), ids))
    for (j in seq_len(n)) {
        p[j, up[[j]]] <- p[j, up[[j]]] + rho / length(up[[j]])
    }
    mu <- s * depth * p
    phi <- (0.1 + 1 / mu) * rep(e, each = n)
    matrix(rnbinom(length(mu), mu = mu, size = 1 / phi), n,
        dimnames = dimnames(p))
}
```

# Distances and tests

- **MMD**: `sample_mmd()`, the maximum mean discrepancy between samples under
  a Gaussian kernel on the protein embeddings (see the
  [Reference](reference.html#sample-level-distance-mmd)).
- **Bray-Curtis (repdist MCL bins)**: `protein_similarity()` →
  `bin_proteins()` → `bin_glom()`, then Bray-Curtis on the bin counts. The
  bins depend only on the universe, so they are built once, with package
  defaults (predicted TM-score ≥ 0.7, coverage ≥ 0.5, inflation 2).
- **Bray-Curtis (MMseqs2 bins)**: counts summed within 50% MMseqs2 sequence
  clusters [@steinegger2017mmseqs2], then Bray-Curtis. The same 50%
  clustering constructs the benchmark (no 50% cluster is enriched twice
  within a condition), so its failure here is partly by design.
- **Bray-Curtis (MMseqs2 30% bins)**: the same at 30% identity, a
  clustering not used in the construction: the more independent sequence
  comparison.
- **Bray-Curtis (protein)**: Bray-Curtis on the protein counts.

PERMANOVA (`adonis2()`) [@anderson2001permanova] tests whether the groups
differ in their distance-based centroids; PERMDISP (`betadisper()` and
`permutest()`) [@anderson2006permdisp] tests whether they differ in spread.
Both use 999 permutations in `vegan` [@oksanen2025vegan]; a test detects at
p ≤ 0.05.

```{r tests}
group_tests <- function(D, group) {
    set.seed(1)
    a <- adonis2(D ~ group, data = data.frame(group = group),
        permutations = 999)
    set.seed(1)
    pd <- permutest(betadisper(D, group), permutations = 999)
    c(`PERMANOVA R2` = a$R2[1], `PERMANOVA p` = a$`Pr(>F)`[1],
        `PERMDISP p` = pd$tab[1, "Pr(>F)"])
}
test_all <- function(D, group) {
    t(vapply(D, group_tests, numeric(3), group = group))
}
```

```{r pcoa-helper, echo = FALSE}
pcoa_plot <- function(D, group) {
    ord <- do.call(rbind, lapply(names(D), function(n) {
        e <- cmdscale(D[[n]], k = 2, eig = TRUE)
        rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
        data.frame(PCo1 = e$points[, 1], PCo2 = e$points[, 2], group = group,
            panel = sprintf("%s\nPCo1 %.0f%%, PCo2 %.0f%%", n, rel[1], rel[2]))
    }))
    ord$panel <- factor(ord$panel, unique(ord$panel))
    ggplot(ord, aes(PCo1, PCo2, colour = group)) +
        geom_point(size = 2) +
        stat_ellipse(linewidth = 0.4) +
        facet_wrap(~panel, scales = "free", nrow = 1) +
        labs(colour = NULL) +
        theme_bw() +
        theme(legend.position = "bottom")
}
```

# Protein universe

```{r universe}
u <- readRDS(file.path("curated", "simulation_universe.rds"))
mm <- readRDS(file.path("curated", "simulation_mmseqs2.rds"))
# the sequence clusters were computed on this exact universe file
stopifnot(identical(mm$provenance$universe_md5,
    unname(tools::md5sum(file.path("curated", "simulation_universe.rds")))))
ec_proteins <- u$proteins
ann <- u$annotation
stopifnot(identical(ann$protein, names(ec_proteins)),
    identical(mm$clusters$protein, names(ec_proteins)))
ann$mmseqs50 <- mm$clusters$mmseqs50
ann$mmseqs30 <- mm$clusters$mmseqs30
c(proteins = length(ec_proteins), EC3 = length(unique(ann$ec3)),
    EC4 = length(unique(ann$ec4)))
u$provenance$steps
```

The proteins are UniProtKB/Swiss-Prot entries [@uniprot2025] (release
`r u$provenance$uniprot_release`) from Bacteria and Archaea, non-fragments,
with an experimentally supported catalytic activity. An entry is kept only
if every EC number it carries, on any reaction or in its name, is one
complete four-level number (EC4) and at least one reaction with that number
has experimental evidence (ECO:0000269). Full-length sequences of 50 to 1000
residues (the TM-Vec head's training range) without selenocysteine or
pyrrolysine are kept, one protein per UniRef50 cluster
[@suzek2015uniref], and embedded with `embed_proteins(model = "tmvec")`
(`data-raw/simulation_universe.R`).

# Sequence clusters

`data-raw/simulation_mmseqs2.R` clusters the same sequences with MMseqs2
`r mm$provenance$mmseqs_version`. The primary clustering uses 50% identity,
the usual protein-family level (as UniRef50); 30% identity is a permissive
sensitivity analysis. Both require the alignment to cover 80% of both
sequences (`--cov-mode 0`) and use the most sensitive prefilter (`-s 7.5`):

```{r mmseqs}
mm$provenance[c("cluster_50", "cluster_30")]
mm$provenance$n_clusters
# every 50% cluster lies inside one 30% cluster
all(tapply(ann$mmseqs30, ann$mmseqs50, function(v) length(unique(v))) == 1)
```

Because the universe is already one protein per UniRef50 cluster, the 50%
clustering is close to the proteins themselves.

# Eligible classes and contrasts

Each sample is enriched in two proteins of its condition's EC3, and every
enriched protein of a condition comes from a different 50% cluster. An EC3
is eligible if it spans at least 10 × 2 = 20 distinct 50% clusters.
Eligibility uses only the EC labels and the MMseqs2 clusters, never the
embeddings, the MCL bins or any result. The eligible EC3s are shuffled with
a fixed seed and paired in order, so each EC3 enters exactly one contrast.

```{r eligible}
n_per_group <- 10L
n_enriched <- 2L
k <- n_per_group * n_enriched
clusters_per <- function(level) {
    tapply(ann$mmseqs50, ann[[level]], function(v) length(unique(v)))
}
n50 <- clusters_per("ec3")
eligible <- sort(names(n50)[n50 >= k])
c(EC3 = length(n50), eligible = length(eligible),
    `eligible EC4` = sum(clusters_per("ec4") >= k))
set.seed(20261008)
shuffled <- sample(eligible)
contrasts <- data.frame(A = shuffled[c(TRUE, FALSE)],
    B = shuffled[c(FALSE, TRUE)])
nrow(contrasts)
```

Only `r sum(clusters_per("ec4") >= k)` EC4 classes meet the same rule, too
few for an EC4 simulation, so EC4 is used to score the bins below.

# Design

For contrast $c$ (seed $c$), condition A draws 20 distinct 50% clusters of
EC3-A and one protein from each, two per sample. To keep the enrichment from
collapsing into one enzyme, clusters are taken one per EC4 of the class
before a second from any EC4, in random order. Condition B does the same in
EC3-B, excluding any cluster A used. The conditions match on everything in
the count model; only the class carrying the signal differs.

$\rho$ takes the values 0, 1%, 2%, 5% and 10%. At $\rho = 0$ both
conditions have the same distribution: a null sanity check. Each contrast has
one simulated count realisation per $\rho$, so the counts below are shares of
contrasts detected, not estimates of type-I error or power. A contrast
uses the same seed at every $\rho$, so its enriched proteins, baseline, size
factors and dispersions do not change with the signal strength.

```{r ec-design}
rhos <- c(0, 0.01, 0.02, 0.05, 0.10)
mm50 <- setNames(ann$mmseqs50, ann$protein)
mm30 <- setNames(ann$mmseqs30, ann$protein)
draw_enriched <- function(ec, exclude = character()) {
    cand <- unique(ann[ann$ec3 == ec & !ann$mmseqs50 %in% exclude,
        c("mmseqs50", "ec4")])
    cand <- cand[sample.int(nrow(cand)), ]
    cand <- cand[!duplicated(cand$mmseqs50), ]
    # one cluster per EC4 before a second from any EC4
    cand <- cand[order(ave(seq_len(nrow(cand)), cand$ec4, FUN = seq_along)), ]
    vapply(seq_len(k), function(r) {
        p <- ann$protein[ann$mmseqs50 == cand$mmseqs50[r] &
            ann$ec4 == cand$ec4[r]]
        p[sample.int(length(p), 1L)]
    }, "")
}
enriched <- function(i) {
    set.seed(i)
    a <- draw_enriched(contrasts$A[i])
    b <- draw_enriched(contrasts$B[i], exclude = mm50[a])
    # no 50% cluster is enriched twice within a condition
    stopifnot(!anyDuplicated(mm50[a]), !anyDuplicated(mm50[b]))
    up <- unname(split(c(a, b), rep(seq_len(2 * n_per_group),
        each = n_enriched)))
    names(up) <- sprintf("s%02d", seq_along(up))
    list(up = up, pairs = cbind(a, b),
        group = factor(rep(c("A", "B"), each = n_per_group)))
}
simulate_ec <- function(i, rho) {
    d <- enriched(i) # sets the seed; the counts continue the same stream
    d$counts <- nb_counts(d$up, ann$protein, rho, d$pairs)
    d
}
```

```{r ec-bins}
S <- protein_similarity(ec_proteins, n_cores = BiocParallel::multicoreWorkers())
bins <- bin_proteins(S)
glom <- function(counts, cluster) t(rowsum(t(counts), cluster[colnames(counts)]))
distances_ec <- function(counts) {
    list(
        MMD = sample_mmd(counts, ec_proteins),
        `Bray-Curtis (repdist MCL bins)` = vegdist(bin_glom(counts, bins),
            "bray"),
        `Bray-Curtis (MMseqs2 bins)` = vegdist(glom(counts, mm50), "bray"),
        `Bray-Curtis (protein)` = vegdist(counts, "bray"),
        `Bray-Curtis (MMseqs2 30% bins)` = vegdist(glom(counts, mm30), "bray")
    )
}
```

The fifth distance, MMseqs2 bins at 30% identity, is the sensitivity
analysis; it is reported in the tables but not in the ordination.

# Representative contrast

The first contrast of the seeded pairing, at $\rho$ = 2%; both were fixed
before any result.

```{r ec-example}
sim <- simulate_ec(1, rho_example)
unlist(contrasts[1, ])
enr <- data.frame(sample = rep(names(sim$up), lengths(sim$up)),
    condition = rep(sim$group, lengths(sim$up)), protein = unlist(sim$up))
enr <- cbind(enr, ann[match(enr$protein, ann$protein),
    c("ec4", "name", "mmseqs50", "mmseqs30")])
enr$mcl_bin <- bins$membership[enr$protein]
# distinct labels among each condition's 20 enriched proteins
t(sapply(split(enr, enr$condition), function(d) c(
    `50% clusters` = length(unique(d$mmseqs50)),
    `30% clusters` = length(unique(d$mmseqs30)),
    `MCL bins` = length(unique(d$mcl_bin)), EC4 = length(unique(d$ec4)))))
# realised share of each sample's counts on its enriched proteins
tapply(vapply(seq_along(sim$up), function(i) {
    sum(sim$counts[i, sim$up[[i]]]) / sum(sim$counts[i, ])
}, numeric(1)), sim$group, median)
D <- distances_ec(sim$counts)
res <- test_all(D, sim$group)
knitr::kable(res, digits = 3)
```

Ellipses are 95% ellipses per condition (multivariate t).

```{r ec-pcoa, echo = FALSE, fig.width = 11, fig.height = 3.8, fig.alt = "PCoA of Simulation 2's representative contrast under four distances, coloured by condition, with 95% ellipses"}
pcoa_plot(D[1:4], factor(paste0(sim$group, ": EC ",
    unlist(contrasts[1, ])[as.character(sim$group)])))
```

# All contrasts

The representative is one of `r nrow(contrasts)` contrasts. Every contrast is
simulated at every $\rho$, so the result can be checked beyond one pair of
enzyme classes.

```{r ec-benchmark}
bench2 <- do.call(rbind, lapply(seq_len(nrow(contrasts)), function(i) {
    do.call(rbind, lapply(rhos, function(rho) {
        s <- simulate_ec(i, rho)
        r <- test_all(distances_ec(s$counts), s$group)
        data.frame(contrast = i, A = contrasts$A[i], B = contrasts$B[i],
            rho = rho, method = rownames(r), r, row.names = NULL,
            check.names = FALSE)
    }))
}))
bench2$method <- factor(bench2$method, names(D))
# the representative contrast is the benchmark's contrast 1 at rho = 2%
stopifnot(isTRUE(all.equal(unname(res[, "PERMANOVA p"]),
    bench2$`PERMANOVA p`[bench2$contrast == 1 & bench2$rho == rho_example])))
```

Contrasts detected by PERMANOVA, out of `r nrow(contrasts)`, with the number
also significant by PERMDISP in parentheses. The $\rho = 0$ column is the
null sanity check (one realisation per contrast; about
`r 0.05 * nrow(contrasts)` detections would be expected by chance at
α = 0.05).

```{r ec-detection, echo = FALSE}
count_by <- function(v) tapply(v, list(bench2$method, bench2$rho), sum)
detect <- count_by(bench2$`PERMANOVA p` <= 0.05)
disp <- count_by(bench2$`PERMDISP p` <= 0.05)
rho_names <- sprintf("ρ = %g%%", 100 * rhos)
knitr::kable(matrix(sprintf("%d (%d)", detect, disp), nrow(detect),
    dimnames = list(rownames(detect), rho_names)))
```

# Protein clustering against functional labels

This benchmark is separate from the sample-level contrasts but uses the same
Swiss-Prot universe. Its MCL bins were built from embeddings and alignments alone;
EC labels are used only here. The two MMseqs2 clusterings are scored the
same way for comparison. With $n_k$ members of label $k$ in a group of size
$n$:

- **Purity** is $\max_k n_k / n$, averaged over groups with at least two
  members.
- **Precision** is the share of same-group protein pairs that share the label.
- **Recall** is the share of same-label protein pairs placed in one group.
- **F1** is $2PR/(P + R)$.

Purity and precision ignore singletons, so many small, pure groups score
high on both while splitting each class across groups; recall measures that
splitting.

```{r grouping-sizes}
groupings <- list(`repdist MCL bins` = bins$membership,
    `MMseqs2 50%` = mm50, `MMseqs2 30%` = mm30)
size_tab <- t(vapply(groupings, function(m) {
    sizes <- as.vector(table(m))
    c(groups = length(sizes), `multi-member` = sum(sizes > 1),
        `singleton proteins (%)` = 100 * sum(sizes == 1) / length(m),
        table(cut(sizes, c(0, 1, 2, 5, 10, 50, Inf),
            labels = c("1", "2", "3-5", "6-10", "11-50", ">50"))),
        largest = max(sizes))
}, numeric(10)))
knitr::kable(size_tab, digits = 1)
```

Columns `1` to `>50` count groups by size.

Pairwise precision, recall and F1 come from `clusterCrit::extCriteria()`
(F1 is its Czekanowski-Dice index).

```{r ec-scores}
labels <- list(EC3 = setNames(ann$ec3, ann$protein),
    EC4 = setNames(ann$ec4, ann$protein))
scores <- function(m, lab) {
    lab <- lab[names(m)]
    tab <- table(m, lab)
    multi <- rowSums(tab) > 1
    pr <- clusterCrit::extCriteria(as.integer(factor(lab)),
        as.integer(factor(m)), c("Precision", "Recall", "Czekanowski_Dice"))
    c(purity = mean(apply(tab[multi, ], 1, max) / rowSums(tab[multi, ])),
        precision = pr$precision, recall = pr$recall,
        F1 = pr$czekanowski_dice)
}
ec_scores <- do.call(rbind, lapply(names(labels), function(level) {
    do.call(rbind, lapply(names(groupings), function(g) {
        data.frame(level = level, grouping = g,
            t(scores(groupings[[g]], labels[[level]])))
    }))
}))
knitr::kable(ec_scores, digits = 3, row.names = FALSE)
```

The three groupings sit in different regimes rather than on one scale:

```{r pr-plot, echo = FALSE, fig.width = 7.5, fig.height = 3.4, fig.alt = "Pairwise precision against recall for repdist MCL bins and MMseqs2 clusters, at EC3 and EC4"}
ggplot(ec_scores, aes(recall, precision, colour = grouping)) +
    geom_point(size = 3) +
    facet_wrap(~level, scales = "free_x") +
    scale_x_continuous(limits = c(0, NA)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "pairwise recall", y = "pairwise precision", colour = NULL) +
    theme_bw()
```

MMseqs2 gives very pure sequence families that split each enzyme class into
many groups, most of them singletons. The MCL bins group across sequence-family
boundaries, with far fewer singletons, while keeping high EC3 coherence; their
EC3 recall is still low in absolute terms. At EC4 the MCL bins recover more
same-enzyme pairs but join more proteins of different EC4s, so the 30%
clusters are more precise there and score a higher F1: exact EC4
reconstruction is better served by sequence clustering in this benchmark.
repdist MCL bins occupy a middle ground between sequence-family fragmentation
and broad functional merging. They are representation-aware abundance units,
not reconstructions of EC classes.

# Summary

```{r summary-values, include = FALSE}
s2 <- function(m) paste(detect[m, as.character(rhos[-1])], collapse = ", ")
mmd_miss <- bench2[bench2$method == "MMD" & bench2$rho > 0 &
    bench2$`PERMANOVA p` > 0.05, ]
detected <- function(m) {
    bench2$`PERMANOVA p`[bench2$method == m & bench2$rho > 0] <= 0.05
}
gain30 <- sum(detected("Bray-Curtis (MMseqs2 30% bins)") &
    !(detected("MMD") & detected("Bray-Curtis (repdist MCL bins)")))
sc <- function(level, g, col) {
    sprintf("%.3f",
        ec_scores[ec_scores$level == level & ec_scores$grouping == g, col])
}
sing <- function(g) sprintf("%.1f%%", size_tab[g, "singleton proteins (%)"])
```

- **Contrasts.** At $\rho$ = 1, 2, 5 and 10%, MMD detected `r s2("MMD")`
  of `r nrow(contrasts)` contrasts; Bray-Curtis on MCL bins
  `r s2("Bray-Curtis (repdist MCL bins)")`, on MMseqs2 30% bins
  `r s2("Bray-Curtis (MMseqs2 30% bins)")`, on MMseqs2 50% bins
  `r s2("Bray-Curtis (MMseqs2 bins)")` and on proteins
  `r s2("Bray-Curtis (protein)")`. The $\rho = 0$ null sanity check gave
  `r sum(detect[, "0"])` detections over the five distances. MMD missed
  `r nrow(mmd_miss)` contrast–$\rho$ combinations, all at
  $\rho \le$ `r 100 * max(mmd_miss$rho)`% with p between
  `r sprintf("%.3f", min(mmd_miss$"PERMANOVA p"))` and
  `r sprintf("%.3f", max(mmd_miss$"PERMANOVA p"))`. PERMDISP was significant for at most
  `r max(disp[, -1])` of `r nrow(contrasts)` contrasts for any distance and
  $\rho > 0$, so the detections are shifts in location.
- **Sequence bins.** The 50% clustering helped construct the benchmark, so
  the failure of the 50% bins is partly by design, not a general comparison
  with MMseqs2. The 30% bins, not used in the construction, detected
  `r gain30` contrast–$\rho$ combinations that MMD or the MCL bins missed.
- **Protein clustering.** At EC3 the MCL bins keep purity
  `r sc("EC3", "repdist MCL bins", "purity")` and precision
  `r sc("EC3", "repdist MCL bins", "precision")` while recovering
  `r sc("EC3", "repdist MCL bins", "recall")` of same-EC3 pairs, against
  `r sc("EC3", "MMseqs2 30%", "recall")` for the 30% and
  `r sc("EC3", "MMseqs2 50%", "recall")` for the 50% clusters, with
  `r sing("repdist MCL bins")` singleton proteins against
  `r sing("MMseqs2 30%")` and `r sing("MMseqs2 50%")`. At EC4 the MCL bins
  recall more (`r sc("EC4", "repdist MCL bins", "recall")` against
  `r sc("EC4", "MMseqs2 30%", "recall")`) but the 30% clusters are more
  precise (`r sc("EC4", "MMseqs2 30%", "precision")` against
  `r sc("EC4", "repdist MCL bins", "precision")`) and score a higher F1
  (`r sc("EC4", "MMseqs2 30%", "F1")` against
  `r sc("EC4", "repdist MCL bins", "F1")`).

Under sequence-disjoint functional contrasts, graded representation
similarity retains relationships that protein identity and hard sequence
bins may discard, and hard representation bins retain part of them. The
result is conditional on that regime, built here by construction; it does not
show that structure is better than sequence. The [case study](case_study.html)
applies the same tools to a real metaproteomic contrast without known
functional truth, and [MCL parameters](parameters.html) shows how the graph
thresholds and inflation move the bins scored above.

```{r session}
sessionInfo()
```

# References {-}
