---
title: "Simulation"
subtitle: "Two benchmarks with known truth"
bibliography: references.bib
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
    message = FALSE, warning = FALSE
)
```

# Overview

Two simulations with known truth answer different questions; the
[case study](case_study.html) then applies repdist to real metaproteomes,
where there is none.

|  | Simulation 1 | Simulation 2 |
| --- | --- | --- |
| Question | Can a sample distance detect a shared latent relationship among enriched proteins when no accession is shared? | Can it separate two functional states when every sample carries its state through different sequence clusters? |
| Proteins | CATH FunFam domains | Swiss-Prot enzymes, full length |
| Truth | FunFam membership | experimental EC3 |
| Groups | one FunFam (coherent) against a matched null | EC3-A against EC3-B, symmetric |
| Sequence clusters | QC: every enriched domain in its own 50% MMseqs2 cluster | the test: no 50% cluster enriched twice within a condition |
| Distances | MMD, Bray-Curtis on MCL bins and on proteins | the same, plus Bray-Curtis on MMseqs2 bins |

Simulation 1 is a controlled positive benchmark: FunFams are tight
structural and functional families, so the hidden coherence is strong.
Simulation 2 is harder and closer to metaproteomics: an EC3 class spans many
sequence families and often several folds. Neither asks whether structure
beats sequence. Both ask whether a representation-aware distance keeps the
relationship between different proteins that protein identity, or hard
sequence bins, discard.

```{r libs}
library(repdist)
library(ggplot2)
library(vegan)
```

## Count model

Both simulations use twenty samples, two groups of ten, and the same
negative-binomial model. Every protein of a universe is present at baseline
in every sample; sample $i$ has expected composition

$$
p_i = (1 - \rho)\,\pi + \rho\, q_i ,
$$

where $\pi_j \propto \mathrm{LogNormal}(0, 1)$ is the common baseline and
$q_i$ spreads its mass evenly over sample $i$'s enriched proteins. The signal
strength $\rho$ is the expected share of abundance on the enriched proteins:
a defined mass, not a fold change. Counts have mean
$\mu_{ij} = s_i D p_{ij}$ with depth $D = 2 \times 10^4$ and size factor
$s_i \sim \mathrm{LogNormal}(0, 0.25)$, and variance $\mu + \phi\mu^2$, with
a DESeq2-like dispersion trend at each count's own mean,
$\phi_{ij} = (0.1 + 1/\mu_{ij})\, e_j$, $e_j \sim \mathrm{LogNormal}(0, 0.5)$.
The $k$-th enriched protein of the second group takes the baseline
abundance of the $k$-th of the first, so the groups also match on the
baseline of the proteins that carry the signal.

The representative signal strength is $\rho$ = 2%, fixed before any result:
in the source study of the case study, glycoside hydrolases rose from under
1% to over 2.5% of the metaproteome on the yeast diet
[@blakeleyruiz2025diet].

```{r count-model}
depth <- 2e4
rho_example <- 0.02
nb_counts <- function(up, ids, rho, pairs) {
    n <- length(up)
    pi <- setNames(rlnorm(length(ids)), ids)
    pi[pairs[, 2]] <- pi[pairs[, 1]] # matched baseline of enriched proteins
    pi <- pi / sum(pi)
    s <- rlnorm(n, 0, 0.25)
    e <- rlnorm(length(ids), 0, 0.5)
    p <- matrix((1 - rho) * pi, n, length(ids), byrow = TRUE,
        dimnames = list(names(up), ids))
    for (j in seq_len(n)) {
        p[j, up[[j]]] <- p[j, up[[j]]] + rho / length(up[[j]])
    }
    mu <- s * depth * p
    phi <- (0.1 + 1 / mu) * rep(e, each = n)
    matrix(rnbinom(length(mu), mu = mu, size = 1 / phi), n,
        dimnames = dimnames(p))
}
```

## Distances and tests

- **MMD**: `sample_mmd()`, the maximum mean discrepancy between samples under
  a Gaussian kernel on the protein embeddings (see the
  [Reference](reference.html#sample-level-distance-mmd)).
- **Bray-Curtis (repdist MCL bins)**: `protein_similarity()` →
  `bin_proteins()` → `bin_glom()`, then Bray-Curtis on the bin counts. The
  bins depend only on the universe, so they are built once per universe,
  with package defaults (predicted TM-score ≥ 0.7, coverage ≥ 0.5,
  inflation 2).
- **Bray-Curtis (MMseqs2 bins)**: counts summed within MMseqs2 sequence
  clusters [@steinegger2017mmseqs2], then Bray-Curtis (Simulation 2 only).
- **Bray-Curtis (protein)**: Bray-Curtis on the protein counts.

PERMANOVA [@anderson2001permanova] tests whether the groups differ in their
distance-based centroids; PERMDISP [@anderson2006permdisp] tests whether they
differ in spread (mean distance to the centroid). Both use 999 permutations
in `vegan` [@oksanen2025vegan]; a test detects at p ≤ 0.05.

```{r tests}
group_tests <- function(D, group) {
    set.seed(1)
    a <- adonis2(D ~ group, data = data.frame(group = group),
        permutations = 999)
    bd <- betadisper(D, group)
    set.seed(1)
    pd <- permutest(bd, permutations = 999)
    disp <- tapply(bd$distances, group, mean)
    c(R2 = a$R2[1], F = a$F[1], p = a$`Pr(>F)`[1],
        setNames(disp, paste("dispersion", names(disp))),
        PERMDISP_p = pd$tab[1, "Pr(>F)"])
}
test_all <- function(D, group) {
    t(vapply(D, group_tests, numeric(6), group = group))
}
```

```{r pcoa-helper, echo = FALSE}
pcoa_plot <- function(D, group) {
    ord <- do.call(rbind, lapply(names(D), function(n) {
        e <- cmdscale(D[[n]], k = 2, eig = TRUE)
        rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
        data.frame(PCo1 = e$points[, 1], PCo2 = e$points[, 2], group = group,
            panel = sprintf("%s\nPCo1 %.0f%%, PCo2 %.0f%%", n, rel[1], rel[2]))
    }))
    ord$panel <- factor(ord$panel, unique(ord$panel))
    ggplot(ord, aes(PCo1, PCo2, colour = group)) +
        geom_point(size = 2) +
        facet_wrap(~panel, scales = "free", nrow = 1) +
        labs(colour = NULL) +
        theme_bw() +
        theme(legend.position = "bottom")
}
dir.create("supplement", showWarnings = FALSE)
```

# Simulation 1 — Hidden coherence across distinct proteins

**Question.** Can a representation-aware sample distance detect a shared
latent protein relationship when protein-accession identity contains no
shared signal?

- **H0 (null group):** the enriched proteins share no imposed relationship.
- **H1 (coherent group):** the enriched proteins all belong to one FunFam,
  but no two samples enrich the same domain, or the same sequence cluster.

At the accession level both groups look equally heterogeneous: ten samples,
ten different enriched domains. Only the coherent group hides a shared
family. This is a positive control, not a symmetric biological comparison,
and FunFam membership is a structural and functional family, not a proxy for
function in general.

## Domain universe

```{r ff-universe}
ff <- readRDS(file.path("curated", "simulation_funfam.rds"))
x1 <- ff$proteins
a1 <- ff$annotation
# every domain has its own 50% cluster and its own UniProt accession
stopifnot(identical(a1$domain, names(x1)), !anyDuplicated(a1$mmseqs50),
    !anyDuplicated(a1$accession))
c(FunFams = nrow(ff$families), domains = length(x1))
ff$provenance$steps
```

FunFams are domain-level families, so the universe holds domain sequences,
not full-length proteins. `data-raw/simulation_funfam.R` takes, in each of
the `r ff$provenance$rules$n_superfamilies` CATH v4.3 superfamilies
[@sillitoe2021cath] with the most S35 representatives, the largest FunFam
[@das2015funfam] whose seed alignment has DOPS ≥ 70. Its Bacteria and
Archaea members of 50 to 1000 residues, with standard residues only, are
clustered with MMseqs2 at 50% identity and 80% coverage of both sequences,
and one domain is kept per cluster (clusters spanning two FunFams are
dropped), then one per UniProt accession, at most
`r ff$provenance$rules$max_members` per FunFam (the largest has
`r max(table(a1$family))`). FunFams are tight orthologue groups: the
`r ff$provenance$steps[["unique_sequences"]]` unique prokaryotic domain
sequences fall into `r ff$provenance$steps[["one_per_50pct_cluster"]]`
clusters at 50% identity. Every domain was embedded with
`embed_proteins(model = "tmvec")`. One FunFam per superfamily keeps
different FunFams structurally distinct.

A FunFam is a target if it has at least ten domains, so its ten enriched
domains come from ten different 50% clusters and ten different proteins. Eligibility uses only FunFam
membership and the clusters, never the embeddings or any result; every
FunFam stays in the universe as background.

```{r ff-targets}
n1 <- 10L
targets <- sort(names(which(table(a1$family) >= n1)))
length(targets)
```

## Design

For target $t$ (seed $t$): the coherent group's ten samples each enrich one
domain of the target FunFam, drawn without replacement. Each null sample
enriches one domain from a different non-target FunFam, so the ten null
domains come from ten FunFams of ten other superfamilies; each is the domain
closest in length to one coherent domain (ties at random), and takes that
domain's baseline abundance. Both groups use the same $\rho$ = 2%, depth,
size factors and dispersion model.

```{r ff-design}
draw_ff <- function(i) {
    set.seed(i)
    coherent <- a1$domain[a1$family == targets[i]]
    coherent <- coherent[sample.int(length(coherent), n1)]
    pool <- a1[a1$family != targets[i], ]
    null <- character(n1)
    for (j in sample.int(n1)) {
        gap <- abs(pool$length - a1$length[match(coherent[j], a1$domain)])
        best <- which(gap == min(gap))
        best <- best[sample.int(length(best), 1L)]
        null[j] <- pool$domain[best]
        pool <- pool[pool$family != pool$family[best], ]
    }
    up <- as.list(c(null, coherent))
    names(up) <- sprintf("s%02d", seq_along(up))
    list(up = up, pairs = cbind(coherent, null),
        group = factor(rep(c("null", "coherent"), each = n1),
            c("null", "coherent")))
}
simulate_ff <- function(i, rho = rho_example) {
    d <- draw_ff(i) # sets the seed; the counts continue the same stream
    d$counts <- nb_counts(d$up, a1$domain, rho, d$pairs)
    d
}
```

```{r ff-bins}
bins1 <- bin_proteins(protein_similarity(x1,
    n_cores = BiocParallel::multicoreWorkers()))
distances_ff <- function(counts) {
    list(
        MMD = sample_mmd(counts, x1),
        `Bray-Curtis (repdist MCL bins)` = vegdist(bin_glom(counts, bins1),
            "bray"),
        `Bray-Curtis (protein)` = vegdist(counts, "bray")
    )
}
```

## Representative FunFam

The representative target is drawn with a fixed seed before any simulation
is run.

```{r ff-example}
set.seed(20261008)
ex1 <- sample(length(targets), 1)
sim1 <- simulate_ff(ex1)
ff$families[ff$families$family == targets[ex1], c("family", "name")]
# no accession or 50% cluster repeats; lengths are matched
stopifnot(!anyDuplicated(unlist(sim1$up)),
    !anyDuplicated(a1$mmseqs50[match(unlist(sim1$up), a1$domain)]))
sapply(list(coherent = sim1$pairs[, 1], null = sim1$pairs[, 2]),
    function(d) summary(a1$length[match(d, a1$domain)]))
D1 <- distances_ff(sim1$counts)
knitr::kable(test_all(D1, sim1$group), digits = 3)
```

```{r ff-pcoa, echo = FALSE, fig.width = 10, fig.height = 3.8, fig.alt = "PCoA of Simulation 1's representative target under three distances, coloured by group"}
pcoa_plot(D1, sim1$group)
```

## All eligible FunFams

```{r ff-all}
within_tm <- function(ids) {
    Z <- S4Vectors::mcols(x1)$embedding[ids, ]
    Z <- Z / sqrt(rowSums(Z^2))
    s <- tcrossprod(Z)
    median(s[upper.tri(s)])
}
bench1 <- do.call(rbind, lapply(seq_along(targets), function(i) {
    s <- simulate_ff(i)
    r <- test_all(distances_ff(s$counts), s$group)
    data.frame(target = targets[i],
        name = ff$families$name[match(targets[i], ff$families$family)],
        within_TM = within_tm(a1$domain[a1$family == targets[i]]),
        method = rownames(r), R2 = r[, "R2"], p = r[, "p"],
        PERMDISP_p = r[, "PERMDISP_p"],
        dispersion_null = r[, "dispersion null"],
        dispersion_coherent = r[, "dispersion coherent"], row.names = NULL)
}))
bench1$method <- factor(bench1$method, names(D1))
write.csv(bench1, file.path("supplement", "simulation1_funfams.csv"),
    row.names = FALSE)
summ1 <- do.call(rbind, lapply(split(bench1, bench1$method), function(d) {
    data.frame(method = d$method[1], targets = nrow(d),
        `PERMANOVA p ≤ 0.05` = sum(d$p <= 0.05),
        `PERMDISP p ≤ 0.05` = sum(d$PERMDISP_p <= 0.05),
        `PERMANOVA only` = sum(d$p <= 0.05 & d$PERMDISP_p > 0.05),
        `median R2` = median(d$R2), check.names = FALSE)
}))
knitr::kable(summ1, digits = 3, row.names = FALSE)
```

`PERMANOVA only` counts targets detected by PERMANOVA without a significant
dispersion difference. The number of MCL bins holding each target FunFam's
domains, counted over targets:

```{r ff-bins-per-target}
bins_per_target <- vapply(targets, function(t) {
    length(unique(bins1$membership[a1$domain[a1$family == t]]))
}, integer(1))
table(bins_per_target)
c(MCL_bins = length(bins1$clusters))
```

Each target's PERMANOVA $R^2$ (filled: p ≤ 0.05), against the median
predicted TM-score between the target FunFam's domains:

```{r ff-plot, echo = FALSE, fig.width = 9, fig.height = 3.6, fig.alt = "PERMANOVA R2 per target FunFam against its within-family predicted TM-score, by distance"}
ggplot(bench1, aes(within_TM, R2, shape = p <= 0.05)) +
    geom_point(size = 2) +
    scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 16),
        labels = c("p > 0.05", "p ≤ 0.05"), name = NULL) +
    facet_wrap(~method, nrow = 1) +
    labs(x = "median within-FunFam predicted TM-score",
        y = "PERMANOVA R2") +
    theme_bw() +
    theme(legend.position = "bottom")
```

Every target, with its R2, p and PERMDISP p per distance:
[`supplement/simulation1_funfams.csv`](supplement/simulation1_funfams.csv).

# Simulation 2 — Sequence-disjoint functional contrasts

**Question.** Can representation-aware analysis recover a condition-level
functional difference when the proteins carrying each function belong to
different sequence clusters in every sample?

Two coherent functional states are compared symmetrically: condition A is
enriched for one EC3, condition B for another, and within a condition no
MMseqs2 sequence cluster is enriched in two samples.

## Protein universe

```{r universe}
u <- readRDS(file.path("curated", "simulation_universe.rds"))
mm <- readRDS(file.path("curated", "simulation_mmseqs2.rds"))
# the sequence clusters were computed on this exact universe file
stopifnot(identical(mm$provenance$universe_md5,
    unname(tools::md5sum(file.path("curated", "simulation_universe.rds")))))
x <- u$proteins
ann <- u$annotation
stopifnot(identical(ann$protein, names(x)),
    identical(mm$clusters$protein, names(x)))
ann$mmseqs50 <- mm$clusters$mmseqs50
ann$mmseqs30 <- mm$clusters$mmseqs30
c(proteins = length(x), EC3 = length(unique(ann$ec3)),
    EC4 = length(unique(ann$ec4)))
u$provenance$steps
```

The proteins are UniProtKB/Swiss-Prot entries [@uniprot2025] (release
`r u$provenance$uniprot_release`) from Bacteria and Archaea, non-fragments,
with an experimentally supported catalytic activity. An entry is kept only
if every EC number it carries, on any reaction or in its name, is one
complete four-level number (EC4) and at least one reaction with that number
has experimental evidence (ECO:0000269). Full-length sequences of 50 to 1000
residues (the TM-Vec head's training range) without selenocysteine or
pyrrolysine are kept, one protein per UniRef50 cluster
[@suzek2015uniref], and embedded with `embed_proteins(model = "tmvec")`
(`data-raw/simulation_universe.R`).

## Sequence clusters

`data-raw/simulation_mmseqs2.R` clusters the same sequences with MMseqs2
`r mm$provenance$mmseqs_version`. The primary clustering uses 50% identity,
the usual protein-family level (as UniRef50); 30% identity is a permissive
sensitivity analysis. Both require the alignment to cover 80% of both
sequences (`--cov-mode 0`) and use the most sensitive prefilter (`-s 7.5`):

```{r mmseqs}
mm$provenance[c("cluster_50", "cluster_30")]
mm$provenance$n_clusters
# every 50% cluster lies inside one 30% cluster
all(tapply(ann$mmseqs30, ann$mmseqs50, function(v) length(unique(v))) == 1)
```

Because the universe is already one protein per UniRef50 cluster, the 50%
clustering is close to the proteins themselves.

## Eligible classes and contrasts

Each sample is enriched in two proteins of its condition's EC3, and every
enriched protein of a condition comes from a different 50% cluster. An EC3
is eligible if it spans at least 10 × 2 = 20 distinct 50% clusters.
Eligibility uses only the EC labels and the MMseqs2 clusters, never the
embeddings, the MCL bins or any result. The eligible EC3s are shuffled with
a fixed seed and paired in order, so each EC3 enters exactly one contrast.

```{r eligible}
n_per_group <- 10L
n_enriched <- 2L
k <- n_per_group * n_enriched
clusters_per <- function(level) {
    tapply(ann$mmseqs50, ann[[level]], function(v) length(unique(v)))
}
n50 <- clusters_per("ec3")
eligible <- sort(names(n50)[n50 >= k])
c(EC3 = length(n50), eligible = length(eligible),
    `eligible EC4` = sum(clusters_per("ec4") >= k))
set.seed(20261008)
shuffled <- sample(eligible)
contrasts <- data.frame(A = shuffled[c(TRUE, FALSE)],
    B = shuffled[c(FALSE, TRUE)])
nrow(contrasts)
```

Only `r sum(clusters_per("ec4") >= k)` EC4 classes meet the same rule, too
few for an EC4 simulation, so EC4 is used to score the bins below.

## Design

For contrast $c$ (seed $c$), condition A draws 20 distinct 50% clusters of
EC3-A and one protein from each, two per sample. To keep the enrichment from
collapsing into one enzyme, clusters are taken one per EC4 of the class
before a second from any EC4, in random order. Condition B does the same in
EC3-B, excluding any cluster A used. The conditions match on everything in
the count model; only the class carrying the signal differs.

$\rho$ takes the values 0, 1%, 2%, 5% and 10%; at $\rho = 0$ both conditions
have the same distribution, so detections are false positives. A contrast
uses the same seed at every $\rho$, so its enriched proteins, baseline, size
factors and dispersions do not change with the signal strength.

```{r ec-design}
rhos <- c(0, 0.01, 0.02, 0.05, 0.10)
mm50 <- setNames(ann$mmseqs50, ann$protein)
mm30 <- setNames(ann$mmseqs30, ann$protein)
draw_enriched <- function(ec, exclude = character()) {
    cand <- unique(ann[ann$ec3 == ec & !ann$mmseqs50 %in% exclude,
        c("mmseqs50", "ec4")])
    cand <- cand[sample.int(nrow(cand)), ]
    cand <- cand[!duplicated(cand$mmseqs50), ]
    # one cluster per EC4 before a second from any EC4
    cand <- cand[order(ave(seq_len(nrow(cand)), cand$ec4, FUN = seq_along)), ]
    vapply(seq_len(k), function(r) {
        p <- ann$protein[ann$mmseqs50 == cand$mmseqs50[r] &
            ann$ec4 == cand$ec4[r]]
        p[sample.int(length(p), 1L)]
    }, "")
}
enriched <- function(i) {
    set.seed(i)
    a <- draw_enriched(contrasts$A[i])
    b <- draw_enriched(contrasts$B[i], exclude = mm50[a])
    up <- unname(split(c(a, b), rep(seq_len(2 * n_per_group),
        each = n_enriched)))
    names(up) <- sprintf("s%02d", seq_along(up))
    list(up = up, pairs = cbind(a, b),
        group = factor(rep(c("A", "B"), each = n_per_group)))
}
simulate_ec <- function(i, rho) {
    d <- enriched(i) # sets the seed; the counts continue the same stream
    d$counts <- nb_counts(d$up, ann$protein, rho, d$pairs)
    d
}
```

```{r ec-bins}
S <- protein_similarity(x, n_cores = BiocParallel::multicoreWorkers())
bins <- bin_proteins(S)
glom <- function(counts, cluster) t(rowsum(t(counts), cluster[colnames(counts)]))
distances_ec <- function(counts) {
    list(
        MMD = sample_mmd(counts, x),
        `Bray-Curtis (repdist MCL bins)` = vegdist(bin_glom(counts, bins),
            "bray"),
        `Bray-Curtis (MMseqs2 bins)` = vegdist(glom(counts, mm50), "bray"),
        `Bray-Curtis (protein)` = vegdist(counts, "bray"),
        `Bray-Curtis (MMseqs2 30% bins)` = vegdist(glom(counts, mm30), "bray")
    )
}
```

The fifth distance, MMseqs2 bins at 30% identity, is the sensitivity
analysis; it is reported in the tables but not in the ordination.

## Representative contrast

The first contrast of the seeded pairing, at $\rho$ = 2%; both were fixed
before any result.

```{r ec-example}
sim <- simulate_ec(1, rho_example)
unlist(contrasts[1, ])
enr <- data.frame(sample = rep(names(sim$up), lengths(sim$up)),
    condition = rep(sim$group, lengths(sim$up)), protein = unlist(sim$up))
enr <- cbind(enr, ann[match(enr$protein, ann$protein),
    c("ec4", "name", "mmseqs50", "mmseqs30")])
enr$mcl_bin <- bins$membership[enr$protein]
write.csv(enr, file.path("supplement", "simulation2_example_enriched.csv"),
    row.names = FALSE)
# no 50% cluster is enriched twice within a condition
stopifnot(!anyDuplicated(enr$mmseqs50[enr$condition == "A"]),
    !anyDuplicated(enr$mmseqs50[enr$condition == "B"]))
# distinct labels among each condition's 20 enriched proteins
t(sapply(split(enr, enr$condition), function(d) c(
    `50% clusters` = length(unique(d$mmseqs50)),
    `30% clusters` = length(unique(d$mmseqs30)),
    `MCL bins` = length(unique(d$mcl_bin)), EC4 = length(unique(d$ec4)))))
# realised share of each sample's counts on its enriched proteins
tapply(vapply(seq_along(sim$up), function(i) {
    sum(sim$counts[i, sim$up[[i]]]) / sum(sim$counts[i, ])
}, numeric(1)), sim$group, median)
D <- distances_ec(sim$counts)
res <- test_all(D, sim$group)
knitr::kable(res, digits = 3)
```

Each sample's enriched proteins:
[`supplement/simulation2_example_enriched.csv`](supplement/simulation2_example_enriched.csv).

```{r ec-pcoa, echo = FALSE, fig.width = 11, fig.height = 3.8, fig.alt = "PCoA of Simulation 2's representative contrast under four distances, coloured by condition"}
pcoa_plot(D[1:4], factor(paste0(sim$group, ": EC ",
    unlist(contrasts[1, ])[as.character(sim$group)])))
```

## Sequence divergence of the enriched proteins

For every contrast and condition, the 180 pairs of enriched proteins from
different samples are checked against the two clusterings, the MCL bins, the
EC4 labels and an all-vs-all MMseqs2 search (`-s 7.5`, E-value ≤ 10^-3^; a
pair without a hit has no detectable sequence similarity). This depends only
on the design, not on $\rho$ or any test.

```{r divergence}
hit_key <- paste(mm$hits$p1, mm$hits$p2)
cross_pairs <- function(sets) {
    s <- rep(seq_along(sets), lengths(sets))
    p <- unlist(sets)
    ij <- which(outer(s, s, "<"), arr.ind = TRUE)
    data.frame(a = p[ij[, 1]], b = p[ij[, 2]])
}
ec4 <- setNames(ann$ec4, ann$protein)
divergence <- do.call(rbind, lapply(seq_len(nrow(contrasts)), function(i) {
    d <- enriched(i)
    do.call(rbind, lapply(c("A", "B"), function(g) {
        pr <- cross_pairs(d$up[d$group == g])
        h <- mm$hits$fident[match(paste(pmin(pr$a, pr$b), pmax(pr$a, pr$b)),
            hit_key)]
        data.frame(contrast = i, condition = g, EC3 = contrasts[i, g],
            `same 50% cluster` = mean(mm50[pr$a] == mm50[pr$b]),
            `same 30% cluster` = mean(mm30[pr$a] == mm30[pr$b]),
            `sequence hit` = mean(!is.na(h)),
            `max identity` = if (all(is.na(h))) NA else max(h, na.rm = TRUE),
            `same MCL bin` = mean(bins$membership[pr$a] ==
                bins$membership[pr$b]),
            `same EC4` = mean(ec4[pr$a] == ec4[pr$b]),
            `distinct EC4` = length(unique(ec4[unlist(d$up[d$group == g])])),
            check.names = FALSE)
    }))
}))
stopifnot(all(divergence$`same 50% cluster` == 0))
knitr::kable(t(vapply(divergence[-(1:3)], function(v) {
    quantile(v, c(0, 0.5, 1), na.rm = TRUE)
}, numeric(3))), digits = 3)
```

Rows give, per condition, the share of cross-sample pairs in one 50%
cluster, in one 30% cluster, with a detectable sequence hit, in one MCL bin
and with one EC4; the highest identity of any hit pair; and the number of
distinct EC4s among the 20 enriched proteins. Columns are the minimum,
median and maximum over the `r nrow(divergence)` conditions.

## All contrasts

```{r ec-benchmark}
bench2 <- do.call(rbind, lapply(seq_len(nrow(contrasts)), function(i) {
    do.call(rbind, lapply(rhos, function(rho) {
        s <- simulate_ec(i, rho)
        r <- test_all(distances_ec(s$counts), s$group)
        data.frame(contrast = i, A = contrasts$A[i], B = contrasts$B[i],
            rho = rho, method = rownames(r), R2 = r[, "R2"], p = r[, "p"],
            PERMDISP_p = r[, "PERMDISP_p"], row.names = NULL)
    }))
}))
bench2$method <- factor(bench2$method, names(D))
# the representative contrast is the benchmark's contrast 1 at rho = 2%
stopifnot(isTRUE(all.equal(unname(res[, "p"]),
    bench2$p[bench2$contrast == 1 & bench2$rho == rho_example])))
write.csv(bench2, file.path("supplement", "simulation2_contrasts.csv"),
    row.names = FALSE)
```

Contrasts detected by PERMANOVA, out of `r nrow(contrasts)`, with the number
also significant by PERMDISP in parentheses. At $\rho = 0$ these are false
positives; about `r 0.05 * nrow(contrasts)` are expected at α = 0.05.

```{r ec-detection, echo = FALSE}
count_by <- function(v) tapply(v, list(bench2$method, bench2$rho), sum)
detect <- count_by(bench2$p <= 0.05)
disp <- count_by(bench2$PERMDISP_p <= 0.05)
rho_names <- sprintf("ρ = %g%%", 100 * rhos)
knitr::kable(matrix(sprintf("%d (%d)", detect, disp), nrow(detect),
    dimnames = list(rownames(detect), rho_names)))
```

MMseqs2 50% bins and proteins coincide at zero in the plot.

```{r ec-detection-plot, echo = FALSE, fig.width = 7.5, fig.height = 3.6, fig.alt = "Share of EC3 contrasts detected by PERMANOVA against signal strength, by distance"}
det <- aggregate(cbind(detected = p <= 0.05) ~ method + rho, bench2, mean)
ggplot(det, aes(100 * rho, detected, colour = method)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = "grey50") +
    geom_line() +
    geom_point(size = 1.8) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "signal strength ρ (%)", y = "share of contrasts detected",
        colour = NULL) +
    theme_bw()
```

Median PERMANOVA $R^2$ over contrasts. $R^2$ is not on a common scale across
distances: Bray-Curtis on many sparse features has a higher floor at
$\rho = 0$.

```{r ec-r2, echo = FALSE}
r2 <- tapply(bench2$R2, list(bench2$method, bench2$rho), median)
colnames(r2) <- rho_names
knitr::kable(r2, digits = 3)
```

## MMseqs2 bins against repdist

Per signal strength, the number of contrasts that one distance detects and
the other does not:

```{r ec-versus, echo = FALSE}
wide <- reshape(bench2[bench2$rho > 0, c("contrast", "rho", "method", "p")],
    idvar = c("contrast", "rho"), timevar = "method", direction = "wide")
hit <- function(m) wide[[paste0("p.", m)]] <= 0.05
only <- function(m1, m2) tapply(hit(m1) & !hit(m2), wide$rho, sum)
mmseqs <- "Bray-Curtis (MMseqs2 bins)"
mcl <- "Bray-Curtis (repdist MCL bins)"
versus <- rbind(
    `MMD, not MMseqs2 bins` = only("MMD", mmseqs),
    `MMseqs2 bins, not MMD` = only(mmseqs, "MMD"),
    `MCL bins, not MMseqs2 bins` = only(mcl, mmseqs),
    `MMseqs2 bins, not MCL bins` = only(mmseqs, mcl),
    `MMseqs2 30% bins, not MCL bins` = only("Bray-Curtis (MMseqs2 30% bins)",
        mcl)
)
colnames(versus) <- sprintf("ρ = %g%%", 100 * as.numeric(colnames(versus)))
knitr::kable(versus)
```

Every contrast, signal strength and distance, with R2, p and PERMDISP p:
[`supplement/simulation2_contrasts.csv`](supplement/simulation2_contrasts.csv).

# Protein clustering against functional labels

This benchmark is separate from the sample-level simulations. The MCL bins
of Simulation 2's universe were built from embeddings and alignments alone;
EC labels are used only here. The two MMseqs2 clusterings are scored the
same way for comparison. With $n_k$ members of label $k$ in a group of size
$n$:

- **Purity** is $\max_k n_k / n$, averaged over groups with at least two
  members.
- **Precision** is the share of same-group protein pairs that share the label.
- **Recall** is the share of same-label protein pairs placed in one group.
- **F1** is $2PR/(P + R)$.

Purity and precision ignore singletons, so many small, pure groups score
high on both while splitting each class across groups; recall measures that
splitting. Each score is also given with the labels shuffled across proteins
and the groups held fixed (mean of 100 shuffles).

```{r grouping-sizes}
groupings <- list(`repdist MCL bins` = bins$membership,
    `MMseqs2 50%` = mm50, `MMseqs2 30%` = mm30)
knitr::kable(t(vapply(groupings, function(m) {
    sizes <- as.vector(table(m))
    c(groups = length(sizes), `multi-member` = sum(sizes > 1),
        `singleton proteins (%)` = 100 * sum(sizes == 1) / length(m),
        table(cut(sizes, c(0, 1, 2, 5, 10, 50, Inf),
            labels = c("1", "2", "3-5", "6-10", "11-50", ">50"))),
        largest = max(sizes))
}, numeric(10))), digits = 1)
```

Columns `1` to `>50` count groups by size.

```{r ec-scores}
labels <- list(EC3 = setNames(ann$ec3, ann$protein), EC4 = ec4)
scores <- function(m, lab) {
    lab <- lab[names(m)]
    nk <- table(paste(m, lab, sep = "\r"))
    group_of <- sub("\r.*", "", names(nk))
    n <- table(m)
    multi <- names(n)[n > 1]
    top <- tapply(as.vector(nk), group_of, max)
    same <- sum(choose(nk, 2))
    precision <- same / sum(choose(n, 2))
    recall <- same / sum(choose(table(lab), 2))
    c(purity = mean(top[multi] / n[multi]), precision = precision,
        recall = recall, F1 = 2 * precision * recall / (precision + recall))
}
set.seed(1)
ec_scores <- do.call(rbind, lapply(names(labels), function(level) {
    do.call(rbind, lapply(names(groupings), function(g) {
        m <- groupings[[g]]
        obs <- scores(m, labels[[level]])
        shuf <- rowMeans(replicate(100, scores(m,
            setNames(sample(labels[[level]]), names(labels[[level]])))))
        data.frame(level = level, grouping = g, t(obs),
            `purity, shuffled` = shuf[["purity"]],
            `precision, shuffled` = shuf[["precision"]],
            `recall, shuffled` = shuf[["recall"]], check.names = FALSE)
    }))
}))
knitr::kable(ec_scores, digits = 3, row.names = FALSE)
```

# Summary

```{r summary-values, include = FALSE}
s1 <- function(m, col) summ1[summ1$method == m, col]
s2 <- function(m) paste(detect[m, as.character(rhos[-1])], collapse = ", ")
mmd_miss <- bench2[bench2$method == "MMD" & bench2$rho > 0 &
    bench2$p > 0.05, ]
mmseqs_gain <- sum(versus[c("MMseqs2 bins, not MMD",
    "MMseqs2 bins, not MCL bins", "MMseqs2 30% bins, not MCL bins"), ])
sc <- function(level, g, col) {
    sprintf("%.3f",
        ec_scores[ec_scores$level == level & ec_scores$grouping == g, col])
}
one_ec4 <- sum(divergence$`distinct EC4` == 1)
null_wider <- with(bench1[bench1$method == "MMD" & bench1$PERMDISP_p <= 0.05, ],
    sum(dispersion_null > dispersion_coherent))
```

- **Simulation 1.** At $\rho$ = 2%, MMD detected
  `r s1("MMD", "PERMANOVA p ≤ 0.05")` of `r length(targets)` target
  FunFams; Bray-Curtis on MCL bins detected
  `r s1("Bray-Curtis (repdist MCL bins)", "PERMANOVA p ≤ 0.05")` and on
  proteins `r s1("Bray-Curtis (protein)", "PERMANOVA p ≤ 0.05")`. For MMD,
  PERMDISP was also significant for `r s1("MMD", "PERMDISP p ≤ 0.05")`
  targets, the null group being more spread in `r null_wider` of them, as
  expected when each null sample moves in its own direction: there PERMANOVA reflects spread as well
  as location, and `r s1("MMD", "PERMANOVA only")` detections show no
  dispersion difference. The MCL bins hold
  `r sum(bins_per_target == 1)` of `r length(targets)` target FunFams in one
  bin, so the bins do carry the hidden family; Bray-Curtis loses it because
  a 2% shift in one of `r length(bins1$clusters)` bins is small against the
  sampling noise summed over all of them. MMD's graded similarity uses the
  shared family directly.
- **Simulation 2.** At $\rho$ = 1, 2, 5 and 10%, MMD detected `r s2("MMD")`
  of `r nrow(contrasts)` contrasts, Bray-Curtis on MCL bins
  `r s2("Bray-Curtis (repdist MCL bins)")`, on MMseqs2 bins
  `r s2("Bray-Curtis (MMseqs2 bins)")` (30% bins:
  `r s2("Bray-Curtis (MMseqs2 30% bins)")`) and on proteins
  `r s2("Bray-Curtis (protein)")`. At $\rho = 0$ false positives totalled
  `r sum(detect[, "0"])` over the five distances and `r nrow(contrasts)`
  contrasts. MMD missed
  `r nrow(mmd_miss)` contrast–$\rho$ combinations, all at $\rho \le$
  `r 100 * max(mmd_miss$rho)`% with p between
  `r sprintf("%.3f", min(mmd_miss$p))` and
  `r sprintf("%.3f", max(mmd_miss$p))`. PERMDISP was significant for at most
  `r max(disp[, -1])` of `r nrow(contrasts)` contrasts for any distance and
  $\rho > 0$, so the detections are shifts in location. MMseqs2 bins
  detected `r mmseqs_gain` contrast–$\rho$ combinations that MMD or the MCL
  bins missed. A condition's 20 enriched proteins span a median of
  `r median(divergence$"distinct EC4")` EC4s; `r one_ec4` of
  `r nrow(divergence)` conditions are a single EC4, an EC3 holding one
  enzyme with enough sequence clusters.
- **Protein clustering.** MCL bins are functionally homogeneous but
  fragment broader EC classes: EC3 purity `r sc("EC3", "repdist MCL bins", "purity")`
  and precision `r sc("EC3", "repdist MCL bins", "precision")`, but recall
  `r sc("EC3", "repdist MCL bins", "recall")`. They do not reconstruct EC
  classes. MMseqs2 clusters are purer still and fragment more (EC3 recall
  `r sc("EC3", "MMseqs2 50%", "recall")` at 50% and
  `r sc("EC3", "MMseqs2 30%", "recall")` at 30%). At EC4 the 30% clusters
  score a higher F1 than the MCL bins
  (`r sc("EC4", "MMseqs2 30%", "F1")` against
  `r sc("EC4", "repdist MCL bins", "F1")`), with higher precision
  (`r sc("EC4", "MMseqs2 30%", "precision")` against
  `r sc("EC4", "repdist MCL bins", "precision")`): an MCL bin often joins
  different EC4s of one class (EC3 precision
  `r sc("EC3", "repdist MCL bins", "precision")`).

Both simulations are built in the regime the method is meant for: related
proteins spread over sequence clusters that do not overlap by construction.
There, a distance on graded representation similarity keeps relationships
between different proteins that protein identity and hard sequence bins
discard, and hard representation bins keep part of them. This does not show
that structure beats sequence in general. The [case study](case_study.html)
goes one step further, to a real metaproteomic contrast with no known
functional truth.

# Supplementary files {-}

- [`supplement/simulation1_funfams.csv`](supplement/simulation1_funfams.csv):
  Simulation 1, every target FunFam and distance.
- [`supplement/simulation2_contrasts.csv`](supplement/simulation2_contrasts.csv):
  Simulation 2, every contrast, signal strength and distance.
- [`supplement/simulation2_example_enriched.csv`](supplement/simulation2_example_enriched.csv):
  the representative contrast's enriched proteins, with their EC4, MMseqs2
  clusters and MCL bin.

```{r session}
sessionInfo()
```

# References {-}
