---
title: "Homology-aware beta diversity with repdist"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Homology-aware beta diversity with repdist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
```

# Why a representation distance

Bray-Curtis, Jaccard and their relatives compare samples through protein
*identity*: two proteins are either the same accession or maximally different.
Metaproteomes break that assumption. Two communities can carry the same
enzymatic function on accessions that never overlap, and an identity-based
distance cannot tell that case apart from any other difference — it simply
reports the maximum and stops.

`repdist` replaces the identity comparison with a distance in a protein
language model's representation space. Each sample becomes an
abundance-weighted distribution over protein embeddings, and two samples are
compared by the maximum mean discrepancy (MMD) between those distributions
under an RBF kernel. The ground distance is defined for *any* pair of
sequences, including proteins with no annotation, no tree position, and no
homolog in the other sample.

This vignette runs the whole workflow on real data:

1. **Beta diversity** with the MMD distance and a PERMANOVA in
   `r BiocStyle::CRANpkg("vegan")`.
2. **Structural bins** at a predicted TM-score of 0.7.
3. **Differential abundance** on the bins with
   `r BiocStyle::Biocpkg("DESeq2")`.
4. **Reading the winner**: the functional annotations of the proteins in the
   most responsive bin.

A companion vignette, *Simulation: shared function without shared accessions*,
shows on simulated data exactly when this matters — and when Bray-Curtis
cannot see a real difference at all.

```{r libs}
library(repdist)
library(ggplot2)
```

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("Yixuan39/repdist")
```

# The example dataset

The study is Blakeley-Ruiz *et al.*, *ISME J* **19**(1) wraf048: mouse faecal
metaproteomes across six dietary protein sources, quantified by spectral
counting. Does dietary protein source restructure the gut metaproteome, and
which proteins carry the signal?

```{r data}
d <- readRDS(system.file("extdata", "diet_metaproteome.rds", package = "repdist"))
names(d)
```

`counts` is rarefied to even depth and is what the beta-diversity step uses;
`counts_raw` is the same samples and proteins before rarefaction, which is what
the differential-abundance step uses. Rarefaction buys every sample the same
footing for a whole-community distance, which is exactly what a per-bin model
does not need — it fits its own per-sample normalization instead.

```{r dims}
dim(d$counts)
table(d$metadata$diet)
```

Each mouse contributes several samples, so `subject` is carried through
everything below as a blocking factor. The full study is 30,561 proteins;
shipped here are the 2000 most abundant after rarefaction, which hold 80% of
the total spectral counts. `inst/scripts/prepare_vignette_data.R` in the source
repository is the script that made this object.

# Embedding

`embed_proteins()` downloads a protein language model and runs it through a
pinned Python environment managed by `r BiocStyle::Biocpkg("basilisk")`; you do
not manage Python packages yourself. Every sequence is embedded whole and every
sequence comes back, in input order — nothing is truncated, subset, or silently
dropped. Because the default model is several gigabytes, the embeddings here
are precomputed; on your own data this is the one call that needs the model:

```{r embed, eval = FALSE}
Z <- embed_proteins(d$sequences)      # default model: "tmvec"
```

```{r Z}
Z <- d$embeddings
## Normalize rounded cached embeddings to the current unit-row contract.
Z <- Z / sqrt(rowSums(Z^2))
dim(Z)
```

`known_models()` lists the registered models, each with a `description` of
what its embedding measures. Those keys are the whole of what `model` accepts:
the pinned environment fixes which architectures load at all, so an arbitrary
HuggingFace repo is rejected rather than embedded on a stack nobody has
checked it against. TM-Vec heads are the reason the scale is
interpretable: the head was trained to regress the TM-score of protein pairs,
so cosine similarity between two TM-Vec embeddings *is* a predicted TM-score.
When structural similarity is not the question, the `esm2-*` models embed
general sequence features instead — at the cost of that calibration.

Other trained heads exist, and some of them cluster proteins better than a
TM-score head does. They are not the default here because their distance has
no referent outside the model: it means similarity as that model was trained
to define it, so a threshold on it cannot be read as anything in particular.
A predicted TM-score can.

```{r models}
names(known_models())
```

# The protein ground distance

`repdist_matrix()` builds the protein-by-protein distance that binning and the
QC plots consume. Compute it once and reuse it — at real proteome sizes it is
the expensive object (roughly 400 MB at 10,000 proteins).

```{r ground}
G_cos <- repdist_matrix(Z)                      # 1 - predicted TM-score
```

`bin_proteins()` reads the *value* as a TM-score, which only `"cosine"` gives.
`sample_repdist()` does not take this object: MMD needs a positive-semi-definite
kernel, which a Gaussian RBF only guarantees on a true metric, and `1 - cosine`
is not one. It computes the Euclidean RBF from the embeddings in two-dimensional
tiles, so that matrix is never held whole. On the unit-norm rows
`embed_proteins()` returns the two are monotone transforms of each other
(`||a - b|| = sqrt(2 * (1 - cos))`) and rank protein pairs identically, so
nothing is given up.

```{r scales}
c(median_predicted_TM = round(1 - median(G_cos), 3),
  pairs_above_0.7     = sum(G_cos <= 0.3))
```

# Step 1: beta diversity

Two distances over the same rarefied counts. Bray-Curtis is the protein-level
baseline: it can only ask whether two samples hold the *same accession*. MMD
asks how far apart the proteins they hold actually are.

```{r distances}
D <- list(
  `Bray-Curtis` = vegan::vegdist(d$counts, "bray"),
  `MMD`         = sample_repdist(d$counts, Z))
```

`sample_repdist()` treats each sample as an abundance-weighted distribution over
embeddings and returns the maximum mean discrepancy between them. Both are
`dist` objects, so they drop straight into `r BiocStyle::CRANpkg("vegan")`.

Both are also abundance-weighted, so both inherit some dependence on how many
proteins a sample carries — MMD doubly so, since a sample's own within-sample
dispersion enters its self-term. One line before reading any $R^2$: how much of
each distance richness alone explains.

```{r qc-richness}
rich <- vegan::specnumber(d$counts)
round(vapply(D, function(x) cor(rich, rowMeans(as.matrix(x))), numeric(1)), 2)
```

Richness still spans `r round(max(rich) / min(rich), 1)`-fold after rarefaction
and tracks both distances, MMD slightly *less* than the protein-identity
baseline. That baseline is the thing to check against: a representation distance
more richness-coupled than Bray-Curtis would be measuring depth, not structure.

## PERMANOVA

Amount of protein, sex and age stay in the model as covariates and diet goes
last, so its reported $R^2$ is what survives *after* them — the conservative
reading. Permutations are blocked within mouse.

```{r permanova}
md <- d$metadata
md$subject <- factor(md$subject)
perm <- permute::how(blocks = md$subject, nperm = 999)

res <- do.call(rbind, lapply(names(D), function(n) {
  set.seed(1)
  a <- vegan::adonis2(D[[n]] ~ dose + sex + age_week + diet, data = md,
                      by = "terms", permutations = perm)
  data.frame(distance = n, R2 = round(a["diet", "R2"], 3),
             F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"])
}))
knitr::kable(res, row.names = FALSE)
```

Both find dietary protein source highly significant. That is worth being
precise about: these are cage-housed lab mice searched against one protein
database, so samples really do share accessions, and accession identity is
something Bray-Curtis can read directly. The representation distance is not
rescuing a failed analysis here — it is being checked against a case where the
conventional method already works, and it agrees. The simulation vignette
shows the case where only one of them survives.

## Ordination

PCoA on each distance, one panel each, coloured by diet. The percentage in each
panel heading is that axis's share of the total variation, computed per distance
because each one has its own eigenvalues.

```{r pcoa, fig.height = 3.6, fig.width = 6.5, fig.alt = "PCoA of the samples under Bray-Curtis and MMD, coloured by dietary protein source"}
ord <- do.call(rbind, lapply(names(D), function(n) {
  e   <- cmdscale(D[[n]], k = 2, eig = TRUE)
  rel <- 100 * e$eig / sum(e$eig[e$eig > 0])   # negative eigenvalues excluded
  data.frame(
    Axis1 = e$points[, 1], Axis2 = e$points[, 2],
    diet  = md[rownames(e$points), "diet"],
    panel = sprintf("%s  (PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2]))
}))

ggplot(ord, aes(Axis1, Axis2, colour = diet)) +
  geom_point(size = 1.6, alpha = 0.85) +
  stat_ellipse(type = "t", linewidth = 0.3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "PCo1", y = "PCo2", colour = "diet") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Step 2: structural bins at TM-score 0.7

PERMANOVA says diet structures the community; it says nothing about which
proteins carry it. Testing 2000 proteins one at a time would pay a
multiple-testing price on units that are not independent — homologous proteins
are the same signal counted repeatedly.

`bin_proteins()` collapses them into bins by complete-linkage clustering on the
ground distance, so **every** pair inside a bin meets the threshold, not just
nearest neighbours. Cosine similarity is a predicted TM-score, so
`min_sim = 0.7` is a stricter cut than TM-score's conventional same-fold
boundary of 0.5 — tighter, more homogeneous bins, at the cost of more of them.

Two QC plots inform the choice of threshold. `plot_similarity_profile()`
histograms each protein's similarity to its nearest neighbour: everything left
of the cut will end up a singleton bin, and a pile-up at 1 would mean
near-duplicate sequences or a saturated model. `plot_bin_profile()` sweeps the
threshold over the tree `bin_proteins()` computes anyway, and brackets the two
ways binning stops saying anything: proteins collapsing into one bin as the
threshold drops (chaining) and proteins falling out as singletons as it rises
(no aggregation left). Pick from the valley between them, on a plateau of the
bin count.

```{r bins}
b <- bin_proteins(G_cos, min_sim = 0.7)
```

```{r qc-bins, fig.height = 5, fig.alt = "Nearest-neighbour similarity histogram, and bin count with chaining and singleton shares across binning thresholds"}
plot_similarity_profile(G_cos, min_sim = 0.7)
plot_bin_profile(b, min_sim = 0.7)
```

Rolling the unrarefied counts up to bin level is one `rowsum()` over the
membership:

```{r bin-counts}
bin <- setNames(rep(names(b$clusters), lengths(b$clusters)), unlist(b$clusters))
bin_counts <- t(rowsum(t(d$counts_raw), bin[colnames(d$counts_raw)]))
c(proteins = ncol(d$counts_raw), bins = ncol(bin_counts))
```

```{r bin-sizes}
sizes <- lengths(b$clusters)
table(cut(sizes, c(0, 1, 2, 5, 10, Inf),
          labels = c("1", "2", "3-5", "6-10", ">10")))
```

`r sum(sizes == 1)` of the `r ncol(d$counts_raw)` proteins have no structural
relative at this threshold and stay in a bin of their own. The rest fall into
`r sum(sizes > 1)` multi-member bins, which is where the collapse buys
something: the number of things to test drops by a factor of
`r round(ncol(d$counts_raw) / ncol(bin_counts), 1)`, and each surviving unit is
a structural group rather than an accession.

# Step 3: differential abundance with DESeq2

`r BiocStyle::Biocpkg("DESeq2")` fits a negative-binomial model per bin on the
**unrarefied** counts — its size factors are the model-based counterpart to
rarefaction. The design is paired: `subject` enters as a fixed blocking factor,
absorbing sex, cage, and every other between-mouse difference, and diet is
tested *within* mice. Size factors use the `"poscounts"` estimator because,
as in most metaproteomes, few bins are observed in every sample.

Diet has six levels, so "the diet effect" on a bin is a likelihood-ratio test
against the model without diet — the counterpart to `adonis2`'s `by = "terms"`,
not a single contrast.

```{r deseq}
dds <- DESeq2::DESeqDataSetFromMatrix(t(bin_counts), md,
                                      design = ~ subject + diet)
dds <- DESeq2::DESeq(dds, test = "LRT", reduced = ~ subject,
                     sfType = "poscounts", quiet = TRUE)
res_lrt <- DESeq2::results(dds)

c(bins_tested = nrow(res_lrt),
  diet_significant = sum(res_lrt$padj <= 0.05, na.rm = TRUE))
```

Nothing about the bins is DESeq2-specific — they are ordinary count columns, so
a compositional model such as LinDA or ALDEx2 drops in the same way.

The LRT has no size to it — no single number is "the" effect when diet has five
contrasts against the Casein reference. So each significant bin is given, as
its effect size, the largest fold change it shows against Casein on any diet.
Two refinements matter here. `nbinomWaldTest()` re-uses the dispersions already
fitted to give each contrast its own test, which the volcano below needs; and
`lfcShrink()` pulls fold changes toward zero in proportion to how little
supports them, so a bin seen in a handful of samples cannot post an enormous
log-ratio on almost no evidence.

```{r effects}
lev  <- setdiff(levels(md$diet), "Casein")
ddsw <- DESeq2::nbinomWaldTest(dds, quiet = TRUE)
shrunk <- lapply(lev, function(l)
  DESeq2::lfcShrink(ddsw, coef = paste0("diet_", l, "_vs_Casein"),
                    type = "normal", quiet = TRUE))
names(shrunk) <- lev
fc <- vapply(shrunk, function(o) o$log2FoldChange, numeric(nrow(res_lrt)))
rownames(fc) <- rownames(res_lrt)

sig <- rownames(res_lrt)[!is.na(res_lrt$padj) & res_lrt$padj <= 0.05]
top <- data.frame(
  bin     = sig,
  members = sizes[sig],
  padj    = formatC(res_lrt[sig, "padj"], digits = 2, format = "e"),
  diet    = lev[apply(abs(fc[sig, ]), 1, which.max)],
  log2FC  = round(fc[cbind(sig, lev[apply(abs(fc[sig, ]), 1, which.max)])], 2))
top <- top[order(-abs(top$log2FC)), ]
knitr::kable(head(top, 8), row.names = FALSE)
```

A volcano plot needs one effect size per point, so the five diet-vs-Casein
contrasts are shown separately: one panel per diet, shrunken log2 fold change
on the x-axis, each point one bin, red where that specific contrast rejects at
`padj <= 0.05`.

```{r volcano, fig.height = 3.4, fig.width = 9, fig.alt = "Volcano plot of bins, one panel per diet: shrunken log2 fold change against Casein versus -log10 p, red where that contrast rejects"}
volc <- do.call(rbind, lapply(lev, function(l) {
  o <- shrunk[[l]]
  data.frame(bin = rownames(o), diet = l, log2FoldChange = o$log2FoldChange,
             pvalue = o$pvalue, reject = !is.na(o$padj) & o$padj <= 0.05)
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                     guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) +
  theme_bw()
```

# Step 4: the three most responsive bins, read functionally

The very top of that table is telling in its own right: several of the most
responsive bins contain only proteins with **no database annotation at all** —
groups a lineage- or profile-based method cannot form, held together here by
structure alone. But a bin with no annotations cannot demonstrate what the
annotations of a bin look like, so take the three most responsive bins whose
members carry them:

```{r pick}
n_ann <- vapply(b$clusters[top$bin],
                function(m) sum(!is.na(d$taxa[m, "annotation"])), integer(1))
picks <- head(top[top$members > 1 & n_ann >= 2, ], 3)
knitr::kable(picks, row.names = FALSE)
```

Their members, with the counts that produced those fold changes:

```{r members}
info <- do.call(rbind, lapply(seq_len(nrow(picks)), function(i) {
  m <- b$clusters[[picks$bin[i]]]
  data.frame(
    bin         = picks$bin[i],
    genus       = d$taxa[m, "genus"],
    annotation  = substr(d$taxa[m, "annotation"], 1, 52),
    mean_Casein = round(colMeans(d$counts_raw[md$diet == "Casein", m, drop = FALSE]), 1),
    mean_top    = round(colMeans(d$counts_raw[md$diet == picks$diet[i], m, drop = FALSE]), 1))
}))
knitr::kable(info, row.names = FALSE)
```

Read the annotation column one bin at a time. In **`r picks$bin[1]`** all four
members carry the same call, *endo-alpha-N-acetylgalactosaminidase* — one
mucin-O-glycan-cleaving activity supplied by four distinct accessions from four
different genome bins. **`r picks$bin[2]`** is the same situation with the
synonyms left in: a component of the oligopeptide permease and a
dipeptide-binding protein are both substrate-binding subunits of peptide ABC
importers, one job under two names, and the third member has no annotation at
all. **`r picks$bin[3]`** is the pattern at scale — nine members, every one a
*SusD-like protein*, the outer-membrane glycan-binding lipoprotein of the
Bacteroidetes starch-utilization system, drawn from four named genera and two
unclassified bins.

That is the workflow's point. At the protein level each of these is several weak
tests on units that are not independent; as a bin it is one unit, one test, and
one functional reading — including for the members with no annotation, which
inherit a specific, checkable hypothesis from their bin-mates instead of
silence. All three fall on the same diet and in the same direction — glycan
degradation and peptide uptake moving together — which is a property of that
diet's proteome rather than of any one accession.

The complete-linkage guarantee is that every member is within the threshold of
every *other* member, not merely of a nearest neighbour. That is true by
construction, but `plot_bin_similarity()` is where you see how far above the
threshold a bin actually sits, and whether it is one tight group or two
sub-groups that only just met it:

```{r floor, fig.height = 4.5, fig.alt = "Pairwise predicted TM-score heatmap among the members of the most responsive annotated bin"}
plot_bin_similarity(b, picks$bin[1], G_cos)
```

# Practical notes

**Rarefy first** for the whole-community distance. `sample_repdist()` is
abundance-weighted, so it inherits the usual depth sensitivity; rarefy with
`vegan::rrarefy()` when samples were not collected at a common depth. The
differential-abundance model wants the opposite — raw counts, with depth
handled by its size factors.

**Memory.** `repdist_matrix()` is quadratic in the number of retained proteins
and the index cap is 46,341, so bin the *identified* subset of a catalog, never
a whole search database. `sample_repdist()` holds nothing quadratic: it
accumulates 1000-by-1000 kernel tiles directly into the sample Gram matrix,
keeps sparse counts sparse, and reads its default bandwidth off a 2000-protein
subsample.

**Bandwidth.** `sample_repdist()` returns the bandwidth it used as
`attr(, "sigma")`; pass it back to compare across studies. With the default the
kernel is nearly linear in TM-score, so the distance is close to the euclidean
distance between abundance-weighted mean embeddings, and on this data that is
also where separation peaks. Checking that is one line: the largest exponent the
kernel ever sees, `max(G_cos) / attr(D$MMD, "sigma")^2` (the euclidean
`||a - b||^2 / (2 sigma^2)` on unit-norm rows), is
`r round(max(G_cos) / attr(D$MMD, "sigma")^2, 2)` here, so `exp()` never
leaves its near-linear stretch. A much larger value means the kernel is
saturating and only near-identical proteins still count.

The full analysis runs this same workflow on every protein that survives
rarefaction and a 3,000aa ceiling, rather than the 2,000-protein package
subset. That ceiling is a choice about TM-Vec's training range, not something
`repdist` imposes -- `embed_proteins()` embeds every sequence whole and no
function here takes a maximum length. It is published at
<https://yixuan39.github.io/repdist/>, with its source on the `gh-pages`
branch.

```{r session}
sessionInfo()
```
