---
title: "Homology-aware beta diversity with repdist"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Homology-aware beta diversity with repdist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4, message = FALSE)
```

# Why a representation distance

Bray-Curtis, Jaccard and their relatives compare samples through protein
*identity*: two proteins are either the same accession or maximally different.
Metaproteomes break that assumption. Two communities can carry the same
enzymatic function on accessions that never overlap, and an identity-based
distance cannot tell that case apart from any other difference — it simply
reports the maximum and stops.

`repdist` replaces the identity comparison with a distance in a protein
language model's representation space. Each sample becomes an
abundance-weighted distribution over protein embeddings, and two samples are
compared by the maximum mean discrepancy (MMD) between those distributions
under an RBF kernel. The ground distance is defined for *any* pair of
sequences, including proteins with no annotation, no tree position, and no
homolog in the other sample.

```{r load}
library(repdist)
```

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("Yixuan39/repdist")
```

# The workflow

Three steps, and the middle one is the expensive object worth reusing.

1. **Embed** sequences with `embed_proteins()`, or import embeddings you
   already have with `read_embeddings()`.
2. **Build the protein ground distance** once with `seq_repdist()`.
3. **Compare samples** with `sample_repdist()`, and/or **collapse proteins into
   structural bins** with `repdist_bin()`.

## Embedding

`embed_proteins()` downloads a protein language model and runs it through a
pinned Python environment managed by `r BiocStyle::Biocpkg("basilisk")`; you do
not manage Python packages yourself. Because the default model is several
gigabytes, the call is not evaluated here.

```{r embed, eval = FALSE}
seqs <- read_fasta("proteins.fasta")
Z <- embed_proteins(seqs)                       # default: tmvec-swissmodel-large
```

`known_models()` lists the registered models. TM-Vec heads are the reason the
cosine scale is interpretable: the head was trained to regress the TM-score of
protein pairs, so cosine similarity between two TM-Vec embeddings *is* a
predicted TM-score.

```{r models}
names(known_models())
```

Sequences are embedded complete, never truncated. `memory_fraction` is the
single batching control — batch sizes are derived from currently available CPU,
CUDA or MPS memory. A protein that cannot fit alone, or that exceeds the
backbone's architectural context, is skipped with a warning rather than
silently clipped.

Embeddings computed elsewhere can be imported directly. Official TM-Vec `.npz`
databases use `headers` rather than `ids`; both are detected.

```{r read-emb, eval = FALSE}
Z <- read_embeddings("protein_embeddings.npz")
```

# A worked example

Everything below runs without a model download, on a small simulated protein
universe built to make the identity/representation contrast visible.

Four structural families, twelve proteins each. Embeddings are unit-norm, as
`embed_proteins()` returns them.

```{r universe}
set.seed(1)
n_fam <- 4L; per_fam <- 12L; d <- 32L

centres <- matrix(0.12, n_fam, d)
for (f in seq_len(n_fam)) centres[f, ((f - 1L) * 8L + 1L):(f * 8L)] <- 1.5
family <- rep(seq_len(n_fam), each = per_fam)

Z <- centres[family, ] +
  matrix(abs(rnorm(n_fam * per_fam * d, 0, 0.2)), n_fam * per_fam, d)
Z <- Z / sqrt(rowSums(Z^2))
rownames(Z) <- sprintf("prot%02d", seq_len(nrow(Z)))
dim(Z)
```

Eight samples in two groups. Group A expresses families 1 and 2, group B
families 3 and 4 — but **every sample is given its own private set of
accessions**, so no two samples share a single protein identifier.

```{r counts}
counts <- matrix(0L, 8L, nrow(Z),
                 dimnames = list(sprintf("s%d", 1:8), rownames(Z)))
for (i in seq_len(8L)) {
  fams <- if (i <= 4L) c(1L, 2L) else c(3L, 4L)
  k <- if (i <= 4L) i else i - 4L
  idx <- unlist(lapply(fams,
    function(f) (f - 1L) * per_fam + ((k - 1L) * 3L + 1:3)))
  counts[i, idx] <- rpois(length(idx), 30) + 10L
}
group <- factor(rep(c("A", "B"), each = 4))

# no accession is shared by any pair of samples
max(sapply(1:7, function(i) max(sapply((i + 1):8,
  function(j) sum(counts[i, ] > 0 & counts[j, ] > 0)))))
```

## The protein ground distance

`seq_repdist()` builds the protein-by-protein distance everything else
consumes. Compute it once and pass it around: at real proteome sizes it is the
expensive object (roughly 400 MB at 10,000 proteins).

```{r ground}
G <- seq_repdist(Z)          # cosine: 1 - predicted TM-score
M <- as.matrix(G)
c(within_family  = 1 - mean(M[1:12, 1:12][lower.tri(diag(12))]),
  between_family = 1 - mean(M[1:12, 13:24]))
```

Within a family the predicted TM-score is 0.98; across families 0.37. That
straddles 0.5, TM-score's conventional same-fold threshold.

## Beta diversity

`sample_repdist()` returns a `dist`, so it drops straight into
`r BiocStyle::CRANpkg("vegan")`. Compare it against Bray-Curtis on the same
counts.

```{r beta}
D  <- sample_repdist(counts, Z)
bc <- vegan::vegdist(counts, "bray")

c(bray_curtis_min = min(bc), bray_curtis_max = max(bc))
```

Bray-Curtis is exactly 1 for every pair. With no shared accessions it has
nothing to work with, and the two groups are indistinguishable from any other
pair of samples.

```{r beta-repdist}
Dm <- as.matrix(D)
c(within_group  = mean(Dm[1:4, 1:4][lower.tri(diag(4))]),
  between_group = mean(Dm[1:4, 5:8]))
```

`sample_repdist()` separates them by an order of magnitude, because it compares
where the samples sit in representation space rather than which accessions they
happen to carry.

```{r permanova}
md <- data.frame(group = group)
rbind(
  `Bray-Curtis` = vegan::adonis2(bc ~ group, data = md, permutations = 999)[1, c("R2", "Pr(>F)")],
  `repdist`     = vegan::adonis2(D  ~ group, data = md, permutations = 999)[1, c("R2", "Pr(>F)")])
```

## The Euclidean ground metric

`sample_repdist()` always uses Euclidean distance, while `seq_repdist()` and
`repdist_bin()` default to cosine distance. That split is deliberate.

MMD is a distance only when the kernel is positive semi-definite, and a
Gaussian RBF guarantees that only on a true metric. `1 - cosine` is not one: it
violates the triangle inequality, and the kernel it induces can be indefinite —
on real TM-Vec embeddings its smallest eigenvalue runs to `-5e-1`. Rather than
return a meaningless number, `sample_repdist()` refuses.

Nothing is lost by the split. For unit-norm embeddings — what `embed_proteins()`
returns — the two metrics are monotone transforms of one another:

```{r equivalence}
dc <- seq_repdist(Z, "cosine")
de <- seq_repdist(Z, "euclidean")
max(abs(as.vector(de) - sqrt(2 * as.vector(dc))))          # ||a-b|| = sqrt(2(1-cos))
cor(as.vector(dc), as.vector(de), method = "spearman")
```

They rank protein pairs identically; euclidean simply expresses that ordering
on a scale that is a true metric. Use `"cosine"` where the *value* has to be
read as a TM-score — that is, for `repdist_bin()`'s `cluster_threshold` — and
`"euclidean"` wherever the distance feeds a kernel.

This equivalence holds only for unit-norm rows. On an embedding matrix with
varying row norms, Euclidean distance also picks up magnitude. A precomputed
non-Euclidean `dist` is rejected by `sample_repdist()`.

# Structural bins

`repdist_bin()` collapses proteins into bins by complete-linkage clustering on
the ground distance, so that *every* pair inside a bin meets the requested
TM-score threshold. This is the route to differential abundance at the level of
structure rather than accession, which is where biomarker discovery usually
wants to be: a signal spread thinly across fifty homologous accessions is
invisible to a per-protein test after multiple-testing correction, but
concentrated in one bin it is not.

```{r bins}
b <- repdist_bin(counts, G, cluster_threshold = 0.5)
table(b$bins$bin)
```

The four families are recovered exactly, twelve proteins each, without ever
being told there were four. The bin-level count table shows the group structure
that the accession-level table could not:

```{r bin-counts}
b$bins$counts
```

`repdist_bin()` also returns the complete-linkage tree over the original
proteins, the same tree pruned to bin representatives, and — when `seqs` is
supplied — the medoid sequence of each bin as an `AAStringSet`, ready for
annotation.

```{r bin-parts}
b$tree
b$bins$reps
```

The pruned tree is a similarity hierarchy, not a phylogeny; `repdist` fits it,
and an external implementation such as `phyloseq::UniFrac()` can consume it.

For a bin-resolution MMD, hand `sample_repdist()` the medoid *embeddings*
rather than `bin_dist`. `bin_dist` is on the cosine scale by construction — the
scale `cluster_threshold` is read against — and so is not the right input to a
kernel; the medoid rows carry the same information and let the euclidean
default apply.

```{r bin-mmd}
Z_bin <- Z[b$bins$reps$protein, ]
rownames(Z_bin) <- b$bins$reps$bin
D_bin <- sample_repdist(b$bins$counts, Z_bin)
round(as.matrix(D_bin), 3)
```

The within-group distances are near zero and the between-group ones sit at
0.62, on four bins rather than forty-eight proteins.

Several thresholds can share one pairwise distance by passing a vector, which
avoids recomputing the expensive object:

```{r bins-multi}
b2 <- repdist_bin(counts, G, cluster_threshold = c(0.5, 0.7))
vapply(b2$bins, function(x) ncol(x$counts), integer(1))
```

# Quality control

Embedding models have a training length range, and proteins beyond it are the
ones most likely to be mis-embedded. `length_abundance_profile()` shows how
much of the catalog *and* how much of the abundance lies past a candidate
cutoff — the second curve is the one that matters, since a handful of very long
proteins can carry a large share of the signal.

```{r qc, fig.alt = "Proportion of proteins and of abundance longer than each length"}
set.seed(2)
lengths <- setNames(round(exp(rnorm(nrow(Z), log(320), 0.6))) + 40, rownames(Z))
length_abundance_profile(counts, lengths, max_length = 1000)
```

# Practical notes

**Rarefy first.** `sample_repdist()` is abundance-weighted, so it inherits the
usual depth sensitivity. Rarefy with `vegan::rrarefy()` when samples were not
collected at a common depth.

**Memory.** The RBF kernel materializes a protein-by-protein Gram matrix, so
cost is quadratic in the number of retained proteins. `repdist_bin()` is the
coarser route when that matrix is too large; in practice the ceiling for
binning is around 75,000 proteins on 32 GB. Bin the *identified* subset of a
catalog, never a whole search database.

**Reuse a compatible ground distance.** `sample_repdist()` accepts
`seq_repdist(Z, "euclidean")`; `repdist_bin()` uses the cosine default. They are
different objects because only the Euclidean one defines a valid RBF-MMD.

```{r session}
sessionInfo()
```
