---
title: "Structure-aware beta diversity with repdist"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Structure-aware beta diversity with repdist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>", fig.width = 7,
    fig.height = 4, message = FALSE, warning = FALSE
)
```

# Why a representation distance

Bray-Curtis, Jaccard and their relatives compare samples through protein
*identity*: two proteins are either the same accession or maximally different.
Metaproteomes break that assumption. Two communities can carry the same
enzymatic function on accessions that never overlap, and an identity-based
distance cannot tell that case apart from any other difference — it simply
reports the maximum and stops.

`repdist` replaces the identity comparison with a distance in a protein
language model's representation space. Each sample becomes an
abundance-weighted distribution over protein embeddings, and two samples are
compared by the maximum mean discrepancy (MMD) between those distributions
under an RBF kernel. The ground distance is defined for *any* pair of
sequences, including proteins with no annotation, no tree position, and no
homolog in the other sample.

This vignette runs the whole workflow on real data:

1. **Beta diversity** with the MMD distance and a PERMANOVA in
    `r BiocStyle::CRANpkg("vegan")`.
2. **Structural bins** by Markov clustering of the graph joining proteins
    with a predicted TM-score of at least 0.7.
3. **Differential abundance** on the bins with
    `r BiocStyle::Biocpkg("DESeq2")`.
4. **Reading the winner**: the functional annotations of the proteins in the
    most responsive bin.

The [simulation study](https://yixuan39.github.io/repdist/simulation.html)
on the package website shows on simulated data exactly when this matters — and when Bray-Curtis
cannot see a real difference at all.

```{r libs}
library(repdist)
library(ggplot2)
```

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("repdist")
```

# The example dataset

The study is Blakeley-Ruiz *et al.*, *ISME J* **19**(1) wraf048: mouse faecal
metaproteomes across six dietary protein sources, quantified by spectral
counting. Does dietary protein source restructure the gut metaproteome, and
which proteins carry the signal?

```{r data}
d <- readRDS(system.file("extdata", "diet_metaproteome.rds",
    package = "repdist"
))
names(d)
```

`counts` was rarefied to even depth on the full protein catalog and then cut
to the 2000 most abundant proteins, so the shipped table's row totals differ —
it is **not** itself equal-depth. `sample_mmd()` normalizes each row
anyway. `counts_raw` is the same samples and proteins before rarefaction, which
is what the differential-abundance step uses — it fits its own per-sample
normalization. See `?diet_metaproteome` for every field and the installed
`scripts/DATA_PROVENANCE.md` for source, licence and processing.

```{r dims}
dim(d$counts)
table(d$metadata$diet)
```

Each mouse contributes several samples, so `subject` is carried through
everything below as a blocking factor. The full study is 30,561 proteins;
shipped here are the 2000 most abundant after rarefaction, which hold 80% of
the total spectral counts. `inst/scripts/prepare_vignette_data.R` in the source
repository is the script that made this object.

# Embedding

`embed_proteins()` downloads a protein language model and runs it through a
pinned Python environment managed by `r BiocStyle::Biocpkg("basilisk")`; you do
not manage Python packages yourself. Every sequence is embedded whole and every
sequence comes back, in input order — nothing is truncated, subset, or silently
dropped. Because the default model is several gigabytes, the embeddings here
are precomputed; on your own data this is the one call that needs the model.
Embeddings made with the `tmvec` command-line tool
(<https://github.com/valentynbez/tmvec>) are read with `read_embeddings()`
instead:

```{r embed, eval = FALSE}
x <- embed_proteins("proteins.fasta") # default model: "tmvec"
x <- read_embeddings("db/proteins.npz", seqs = "proteins.fasta")
```

Either way the result is an `AAStringSet`: the sequences, named by accession,
with one embedding row per protein in `mcols(x)$embedding`. The shipped data
holds the same object.

```{r x}
x <- d$proteins
x
dim(S4Vectors::mcols(x)$embedding)
```

`known_models()` lists the registered models, each with a `description` of
what its embedding measures. Those keys are the whole of what `model` accepts:
the pinned environment fixes which architectures load at all, so an arbitrary
HuggingFace repo is rejected rather than embedded on a stack nobody has
checked it against. TM-Vec heads are the reason the scale is
interpretable: the head was trained to regress the TM-score of protein pairs,
so cosine similarity between two TM-Vec embeddings *is* a predicted TM-score.
When structural similarity is not the question, the `esm2-*` models embed
general sequence features instead — at the cost of that calibration.

Other trained heads exist, and some of them cluster proteins better than a
TM-score head does. They are not the default here because their distance has
no referent outside the model: it means similarity as that model was trained
to define it, so a threshold on it cannot be read as anything in particular.
A predicted TM-score can.

```{r models}
names(known_models())
```

# The protein similarity graph

`protein_similarity()` builds the protein-by-protein similarity -- for
TM-Vec, the predicted TM-score -- that binning reads. Pairs below `min_sim` are
set to 0, and so are pairs whose local alignment (BLOSUM62 unless
`substitutionMatrix` says otherwise) covers less than `min_coverage` (default
0.5) of either sequence, so two proteins sharing only one domain are not
joined. Only pairs above `min_sim` are aligned; `n_cores` spreads the
alignments over several cores.

Build it once with no gates, to see every predicted TM-score, and once with
the gates you bin at:

```{r ground}
S_all <- protein_similarity(x, min_sim = 0, min_coverage = 0) # every pair
S <- protein_similarity(x, min_sim = 0.7) # TM >= 0.7, half of both covered
```

`sample_mmd()` does not take this matrix: MMD needs a
positive-semi-definite kernel, which the Gaussian RBF gives on Euclidean
distances but not on `1 - cosine`, so it builds the Euclidean distance from the
embeddings itself. On the unit-norm rows `embed_proteins()` returns the two are
monotone transforms of each other (`||a - b|| = sqrt(2 * (1 - cos))`) and rank
protein pairs identically, so nothing is given up.

```{r scales}
pairs <- S_all[upper.tri(S_all)]
c(
    median_predicted_TM = round(median(pairs), 3),
    pairs_above_0.7 = sum(pairs >= 0.7)
)
```

# Step 1: beta diversity

Two distances over the same rarefied counts. Bray-Curtis is the protein-level
baseline: it can only ask whether two samples hold the *same accession*. MMD
asks how far apart the proteins they hold actually are.

```{r distances}
D <- list(
    `Bray-Curtis` = vegan::vegdist(d$counts, "bray"),
    `MMD`         = sample_mmd(d$counts, x)
)
```

`sample_mmd()` treats each sample as an abundance-weighted distribution over
embeddings and returns the maximum mean discrepancy between them. Both are
`dist` objects, so they drop straight into `r BiocStyle::CRANpkg("vegan")`.

Both are also abundance-weighted, so both inherit some dependence on how many
proteins a sample carries — MMD doubly so, since a sample's own within-sample
dispersion enters its self-term. One line before reading any $R^2$: how much of
each distance richness alone explains.

```{r qc-richness}
rich <- vegan::specnumber(d$counts)
round(vapply(D, function(x) cor(rich, rowMeans(as.matrix(x))), numeric(1)), 2)
```

Richness still spans `r round(max(rich) / min(rich), 1)`-fold after rarefaction
and tracks both distances, MMD slightly *less* than the protein-identity
baseline. That baseline is the thing to check against: a representation
distance more richness-coupled than Bray-Curtis would be the first thing to
suspect of tracking depth rather than structure.

## PERMANOVA

Amount of protein, sex and age stay in the model as covariates and diet goes
last, so its reported $R^2$ is what survives *after* them — the conservative
reading. Permutations are blocked within mouse.

```{r permanova}
md <- d$metadata
md$subject <- factor(md$subject)
perm <- permute::how(blocks = md$subject, nperm = 999)

res <- do.call(rbind, lapply(names(D), function(n) {
    set.seed(1)
    a <- vegan::adonis2(D[[n]] ~ dose + sex + age_week + diet,
        data = md,
        by = "terms", permutations = perm
    )
    data.frame(
        distance = n, R2 = round(a["diet", "R2"], 3),
        F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"]
    )
}))
knitr::kable(res, row.names = FALSE)
```

Both find dietary protein source highly significant. That is worth being
precise about: these are cage-housed lab mice searched against one protein
database, so samples really do share accessions, and accession identity is
something Bray-Curtis can read directly. The representation distance is not
rescuing a failed analysis here — it is being checked against a case where the
conventional method already works, and it agrees. The simulation study on
the package website shows the case where only one of them survives.

## Ordination

PCoA on each distance, one panel each, coloured by diet. The percentage in each
panel heading is that axis's share of the total variation, computed per distance
because each one has its own eigenvalues.

```{r pcoa}
#| fig.height: 3.6
#| fig.width: 6.5
#| fig.alt: >
#|   PCoA of the samples under Bray-Curtis and MMD, coloured by dietary
#|   protein source
ord <- do.call(rbind, lapply(names(D), function(n) {
    e <- cmdscale(D[[n]], k = 2, eig = TRUE)
    rel <- 100 * e$eig / sum(e$eig[e$eig > 0]) # negative eigenvalues excluded
    data.frame(
        Axis1 = e$points[, 1], Axis2 = e$points[, 2],
        diet = md[rownames(e$points), "diet"],
        panel = sprintf("%s  (PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2])
    )
}))

ggplot(ord, aes(Axis1, Axis2, colour = diet)) +
    geom_point(size = 1.6, alpha = 0.85) +
    stat_ellipse(type = "t", linewidth = 0.3) +
    facet_wrap(~panel, scales = "free") +
    labs(x = "PCo1", y = "PCo2", colour = "diet") +
    theme_bw() +
    theme(legend.position = "bottom")
```

# Step 2: structural bins at TM-score 0.7

PERMANOVA says diet structures the community; it says nothing about which
proteins carry it. Testing 2000 proteins one at a time would pay a
multiple-testing price on units that are not independent — homologous proteins
are the same signal counted repeatedly.

`bin_proteins()` collapses them into bins: every nonzero pair of `S`, a
predicted TM-score of at least 0.7 with half of both sequences aligned, is an
edge, and Markov clustering (MCL) at inflation 2 splits the graph. The 0.7
sits above TM-score's conventional same-fold boundary of 0.5, but it is an
edge floor, not an all-pairs guarantee: a bin can hold pairs below it, joined
through shared neighbours.

```{r bins}
b <- bin_proteins(S, inflation = 2)
b$graph
```

`b$graph` describes the graph MCL split. Isolated proteins have no edge at
this floor and become singleton bins; a largest component holding most of the
proteins would mean the floor is too low to separate anything before MCL
starts.

`plot_tsne()` maps the embeddings in two dimensions, here with the 12
largest bins coloured. A bin should sit in one neighbourhood; one split across
two islands, as a couple are here, is worth a look with
`plot_bin_similarity()`. t-SNE keeps neighbours together but not the distances
between groups, so read it for compactness, not for how far apart bins sit.

```{r tsne}
#| fig.height: 5
#| fig.alt: >
#|   t-SNE of the protein embeddings, the 12 largest bins coloured
set.seed(1)
plot_tsne(x, color = b$membership)
```

Rolling the unrarefied counts up to bin level is one `bin_glom()` call:

```{r bin-counts}
binned <- bin_glom(d$counts_raw, b)
c(proteins = ncol(d$counts_raw), bins = ncol(binned))
```

```{r bin-sizes}
sizes <- lengths(b$clusters)
table(cut(sizes, c(0, 1, 2, 5, 10, Inf),
    labels = c("1", "2", "3-5", "6-10", ">10")
))
```

`r sum(sizes == 1)` of the `r ncol(d$counts_raw)` proteins have no structural
relative at this floor and stay in a bin of their own. The rest fall into
`r sum(sizes > 1)` multi-member bins, which is where the collapse buys
something: the number of things to test drops by a factor of
`r round(ncol(d$counts_raw) / ncol(binned), 1)`, and each surviving unit is
a structural group rather than an accession.

# Step 3: differential abundance with DESeq2

`r BiocStyle::Biocpkg("DESeq2")` fits a negative-binomial model per bin on the
**unrarefied** counts — its size factors are the model-based counterpart to
rarefaction. The design is paired: `subject` enters as a fixed blocking factor,
absorbing sex, cage, and every other between-mouse difference, and diet is
tested *within* mice. Size factors use the `"poscounts"` estimator because,
as in most metaproteomes, few bins are observed in every sample.

Diet has six levels, so "the diet effect" on a bin is a likelihood-ratio test
against the model without diet — the counterpart to `adonis2`'s `by = "terms"`,
not a single contrast.

```{r deseq}
dds <- DESeq2::DESeqDataSetFromMatrix(t(binned), md,
    design = ~ subject + diet
)
dds <- DESeq2::DESeq(dds,
    test = "LRT", reduced = ~subject,
    sfType = "poscounts", quiet = TRUE
)
res_lrt <- DESeq2::results(dds)

c(
    bins_tested = nrow(res_lrt),
    diet_significant = sum(res_lrt$padj <= 0.05, na.rm = TRUE)
)
```

Nothing about the bins is DESeq2-specific — they are ordinary count columns, so
a compositional model such as LinDA or ALDEx2 drops in the same way.

The LRT has no size to it — no single number is "the" effect when diet has five
contrasts against the Casein reference. So each significant bin is given, as
its effect size, the largest fold change it shows against Casein on any diet.
Two refinements matter here. `nbinomWaldTest()` re-uses the dispersions already
fitted to give each contrast its own test, which the volcano below needs; and
`lfcShrink()` pulls fold changes toward zero in proportion to how little
supports them, so a bin seen in a handful of samples cannot post an enormous
log-ratio on almost no evidence.

```{r effects}
lev <- setdiff(levels(md$diet), "Casein")
ddsw <- DESeq2::nbinomWaldTest(dds, quiet = TRUE)
shrunk <- lapply(lev, function(l) {
    DESeq2::lfcShrink(ddsw,
        coef = paste0("diet_", l, "_vs_Casein"),
        type = "normal", quiet = TRUE
    )
})
names(shrunk) <- lev
fc <- vapply(shrunk, function(o) o$log2FoldChange, numeric(nrow(res_lrt)))
rownames(fc) <- rownames(res_lrt)

sig <- rownames(res_lrt)[!is.na(res_lrt$padj) & res_lrt$padj <= 0.05]
top <- data.frame(
    bin     = sig,
    members = sizes[sig],
    padj    = formatC(res_lrt[sig, "padj"], digits = 2, format = "e"),
    diet    = lev[apply(abs(fc[sig, ]), 1, which.max)],
    log2FC  = round(fc[cbind(sig, lev[apply(abs(fc[sig, ]), 1, which.max)])], 2)
)
top <- top[order(-abs(top$log2FC)), ]
knitr::kable(head(top, 8), row.names = FALSE)
```

A volcano plot needs one effect size per point, so the five diet-vs-Casein
contrasts are shown separately: one panel per diet, shrunken log2 fold change
on the x-axis, each point one bin, red where that specific contrast rejects at
`padj <= 0.05`.

```{r volcano}
#| fig.height: 3.4
#| fig.width: 9
#| fig.alt: >
#|   Volcano plot of bins, one panel per diet: shrunken log2 fold change
#|   against Casein versus -log10 p, red where that contrast rejects
volc <- do.call(rbind, lapply(lev, function(l) {
    o <- shrunk[[l]]
    data.frame(
        bin = rownames(o), diet = l, log2FoldChange = o$log2FoldChange,
        pvalue = o$pvalue, reject = !is.na(o$padj) & o$padj <= 0.05
    )
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
    geom_point(alpha = 0.6, size = 1) +
    facet_wrap(~diet, nrow = 1) +
    scale_color_manual(
        values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
        guide = "none"
    ) +
    labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) +
    theme_bw()
```

# Step 4: the three most responsive bins, read functionally

The very top of that table is telling in its own right: several of the most
responsive bins contain only proteins with **no database annotation at all** —
groups a lineage- or profile-based method cannot form, held together here by
structure alone. But a bin with no annotations cannot demonstrate what the
annotations of a bin look like, so take the three most responsive bins whose
members carry them:

```{r pick}
n_ann <- vapply(
    b$clusters[top$bin],
    function(m) sum(!is.na(d$taxa[m, "annotation"])), integer(1)
)
picks <- head(top[top$members > 1 & n_ann >= 2, ], 3)
knitr::kable(picks, row.names = FALSE)
```

Their members, with the counts that produced those fold changes:

```{r members}
info <- do.call(rbind, lapply(seq_len(nrow(picks)), function(i) {
    m <- b$clusters[[picks$bin[i]]]
    data.frame(
        bin         = picks$bin[i],
        genus       = d$taxa[m, "genus"],
        annotation  = substr(d$taxa[m, "annotation"], 1, 52),
        mean_Casein = round(colMeans(
            d$counts_raw[md$diet == "Casein", m, drop = FALSE]
        ), 1),
        mean_top = round(colMeans(
            d$counts_raw[md$diet == picks$diet[i], m, drop = FALSE]
        ), 1)
    )
}))
knitr::kable(info, row.names = FALSE)
```

Read the annotation column one bin at a time. In **`r picks$bin[1]`** all four
members carry the same call, *endo-alpha-N-acetylgalactosaminidase* — one
mucin-O-glycan-cleaving activity supplied by four distinct accessions from four
different genome bins. **`r picks$bin[2]`** is the caution: a dihydrolipoyl
dehydrogenase and a thioredoxin reductase are both flavoprotein disulfide
reductases, built on the same FAD- and NAD(P)H-binding domains but doing
different jobs, and the third member has no annotation at all.
**`r picks$bin[3]`** agrees again: three carbon monoxide dehydrogenases, none
from a named genus.

That is the workflow's point, and its limit. At the protein level each of these
is several weak tests on units that are not independent; as a bin it is one
unit, one test, and one structural reading — a shared activity where the
annotations agree, a shared family where they do not — including for the
members with no annotation, which inherit a specific, checkable hypothesis
from their bin-mates instead of silence. All three fall on the same diet and in
the same direction, which is a property of that diet's proteome rather than of
any one accession.

MCL makes no all-pairs promise: members are joined through the graph, so a
pair inside a bin need not be an edge. `plot_bin_similarity()` colours every
pair by its predicted TM-score and crosses out every pair that is not an edge,
so it shows whether a bin is one tight group or two sub-groups joined through a
few edges. A crossed pair above 0.7 failed the coverage gate:

```{r floor}
#| fig.height: 4.5
#| fig.alt: >
#|   Pairwise predicted TM-score heatmap among the members of the most
#|   responsive annotated bin
plot_bin_similarity(b, picks$bin[1], S, x)
```

# Practical notes

**Sampling depth.** `sample_mmd()` normalizes each row, which removes
total-count scaling but not the effect of depth on which proteins were seen at
all; check depth and richness as confounders, and rarefy with
`vegan::rrarefy()` if you want them off the table, at the cost of discarded
counts. The differential-abundance model wants the opposite — raw counts, with
depth handled by its size factors.

**Memory.** `protein_similarity()` and `sample_mmd()` hold
protein-by-protein matrices, so bin the *identified* subset of a catalog, never
a whole search database.

**Bandwidth.** `sample_mmd()` returns the bandwidth it used as
`attr(, "sigma")`; pass it back to compare across studies. With the default the
kernel is nearly linear in TM-score, so the distance is close to the euclidean
distance between abundance-weighted mean embeddings, and on this data that is
also where separation peaks. Checking that is one line: the largest exponent the
kernel ever sees, `(1 - min(S_all)) / attr(D$MMD, "sigma")^2` (the euclidean
`||a - b||^2 / (2 sigma^2)` on unit-norm rows), is
`r round((1 - min(S_all)) / attr(D$MMD, "sigma")^2, 2)` here, so `exp()` never
leaves its near-linear stretch. A much larger value means the kernel is
saturating and only near-identical proteins still count.

The full analysis runs this same workflow on every protein that survives
rarefaction and a 3,000aa ceiling, rather than the 2,000-protein package
subset. That ceiling is a choice about TM-Vec's training range, not something
`repdist` imposes -- `embed_proteins()` embeds every sequence whole and no
function here takes a maximum length. It is published at
<https://yixuan39.github.io/repdist/>, with its source on the `gh-pages`
branch.

# References

- Blakeley-Ruiz et al. (2025). Dietary protein source alters gut microbiota
    composition and function. <https://doi.org/10.1093/ismejo/wraf048>.
- Hamamsy et al. (2024). Protein remote homology detection and structural
    alignment using deep learning. <https://doi.org/10.1038/s41587-023-01917-2>.
- Gretton et al. (2012). A Kernel Two-Sample Test.
    <https://jmlr.org/papers/v13/gretton12a.html>.

# Session information

```{r session}
sessionInfo()
```
