---
title: "Simulation: shared function without shared accessions"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Simulation: shared function without shared accessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 10,
                      fig.height = 3.4, message = FALSE, warning = FALSE)
```

# The question

When different proteins contribute the same function, can a beta-diversity
method tell two groups apart? Bray-Curtis compares samples through accession
identity, so a functional shift spread across *non-overlapping* homologs is
invisible to it by construction. This vignette builds exactly that case and
runs four distances over it:

1. **RBF-MMD** on TM-Vec embeddings (`sample_repdist()`).
2. **Weighted UniFrac** on the structure tree `bin_proteins()` computes.
3. **Bray-Curtis on structural bins** cut at a predicted TM-score of 0.7.
4. **Bray-Curtis at the protein level** — the conventional baseline.

The first three read protein relatedness, each through a different lens; the
fourth reads identity only.

```{r libs}
library(repdist)
library(ggplot2)
```

# The protein universe

Ten CATH FunFams — functional families of structural domains — with 150
members each, embedded with TM-Vec. The membership table and embeddings ship
with the package; `inst/scripts/prepare_simulation_data.R` in the source
repository is the script that made the object.

```{r universe}
u <- readRDS(system.file("extdata", "funfam_universe.rds", package = "repdist"))
members <- u$members
Z <- u$embeddings
## Normalize rounded cached embeddings to the current unit-row contract.
Z <- Z / sqrt(rowSums(Z^2))
ids <- members$protein
table(members$label)
```

# The design

Twenty samples, ten per group, and **every sample contains every protein**, so
presence/absence carries no information and only abundance can be read. Each
sample has ten proteins upregulated twenty-fold over lognormal baseline
abundances:

- **Group A** draws its ten at random across all ten families — diffuse noise
  with no functional theme.
- **Group B** draws its ten from a single family (SHMT) — a coherent
  functional shift.

Within each group the upregulated sets are **disjoint**, and group A never
draws a protein group B uses, so *no two samples share a single upregulated
accession*. Group B's samples have nothing in common at the accession level
and everything in common at the family level — the case an identity-based
distance cannot see.

```{r design}
n_per_group <- 10L; n_up <- 10L; fold <- 20; depth <- 100000L
set.seed(20260901)

group <- factor(rep(c("A", "B"), each = n_per_group))
b_family <- members$family[members$label == "SHMT"][1]

# Disjoint, not merely independent: draw once, then cut into blocks.
draw_blocks <- function(pool) unname(split(
  sample(pool, n_per_group * n_up), rep(seq_len(n_per_group), each = n_up)))
up_B <- draw_blocks(ids[members$family == b_family])
up_A <- draw_blocks(setdiff(ids, unlist(up_B)))
up <- c(up_A, up_B)

w <- matrix(rlnorm(length(group) * length(ids), sdlog = 0.7),
            nrow = length(group),
            dimnames = list(sprintf("s%02d", seq_along(group)), ids))
for (i in seq_along(group)) w[i, up[[i]]] <- w[i, up[[i]]] * fold
counts <- t(apply(w, 1, function(x) rmultinom(1, depth, x)[, 1]))

c(shared_upregulated_accessions = sum(duplicated(unlist(up))),
  observed_in_every_sample = mean(colSums(counts > 0) == length(group)))
```

# Four distances

One cosine ground distance feeds everything structural: `bin_proteins()` cuts
it at TM-score 0.7 for the bins *and* returns the complete-linkage tree, which
becomes the UniFrac tree via `r BiocStyle::CRANpkg("ape")`. The MMD builds its
euclidean ground metric from the same embeddings internally.

```{r structures}
G_cos <- repdist_matrix(Z)
b <- bin_proteins(G_cos, min_sim = 0.7)
bin <- setNames(rep(names(b$clusters), lengths(b$clusters)), unlist(b$clusters))
bin_counts <- t(rowsum(t(counts), bin[colnames(counts)]))
c(proteins = ncol(counts), bins = ncol(bin_counts))
```

The threshold nearly recovers the design: ten families come back as
`r ncol(bin_counts)` bins without ever being told there were ten.

## What RBF-MMD computes

Each sample is a weighted point cloud in embedding space: protein $a$ enters
with its relative abundance $w_a$. For samples $p$ and $q$,

$$\mathrm{MMD}^2(p,q)\;=\;\sum_{a,b} w^p_a w^p_b\,k(a,b)
\;-\;2\sum_{a,b} w^p_a w^q_b\,k(a,b)
\;+\;\sum_{a,b} w^q_a w^q_b\,k(a,b),$$

under a Gaussian kernel on the euclidean ground metric,
$k(a,b)=\exp\!\left(-\lVert z_a-z_b\rVert^2/2\sigma^2\right)$, which on the
unit-norm rows `embed_proteins()` returns is exactly
$\exp\!\left(-(1-\mathrm{TM}_{ab})/\sigma^2\right)$.

The middle term is the whole difference from Bray-Curtis. Two *different*
proteins with $k(a,b)\approx 1$ cancel there almost as completely as the same
accession would. Shrink $k$ to the indicator of $a=b$ and the expression
collapses to $\lVert w^p - w^q\rVert^2$ — an identity-only distance, blind to
any relationship between distinct proteins. That is the corner this simulation
puts the protein-level baseline in by construction.

`sigma` defaults to the median pairwise distance among the retained proteins
(over a 2000-protein subsample once there are more), the usual median heuristic; the value actually used comes back as
`attr(D, "sigma")`, and passing it back holds the bandwidth fixed across runs
or studies.

```{r distances}
tree <- ape::as.phylo(b$tree)
D <- list(
  `RBF-MMD`                 = sample_repdist(counts, Z),
  `UniFrac (structure tree)` = as.dist(
    GUniFrac::GUniFrac(counts, tree, alpha = 1, verbose = FALSE)$unifracs[, , "d_1"]),
  `Bray-Curtis (TM 0.7 bins)` = vegan::vegdist(bin_counts, "bray"),
  `Bray-Curtis (protein)`     = vegan::vegdist(counts, "bray"))
```

# PERMANOVA

```{r permanova}
md <- data.frame(group = group)
perma <- do.call(rbind, lapply(names(D), function(n) {
  set.seed(1)
  a <- vegan::adonis2(D[[n]] ~ group, data = md, permutations = 999)
  data.frame(distance = n, R2 = round(a$R2[1], 3),
             F = round(a[["F"]][1], 2), p = a$`Pr(>F)`[1])
}))
knitr::kable(perma, row.names = FALSE)
```

Every distance that can read protein relatedness — through the kernel, the
tree, or the bins — finds the group difference decisively. Protein-level
Bray-Curtis finds nothing: each sample's upregulated set is private, so to an
identity-based comparison every sample is equally unlike every other, in both
groups.

# Ordination

PCoA on each distance, in the order above. The percentages in each panel
heading are that axis's share of the total variation, computed per distance
because each has its own eigenvalues.

```{r pcoa, fig.width = 8, fig.height = 7, fig.alt = "PCoA of the simulated samples under RBF-MMD, structure-tree UniFrac, bin-level Bray-Curtis, and protein-level Bray-Curtis, coloured by group"}
ord <- do.call(rbind, lapply(names(D), function(n) {
  e   <- cmdscale(D[[n]], k = 2, eig = TRUE)
  rel <- 100 * e$eig / sum(e$eig[e$eig > 0])   # negative eigenvalues excluded
  data.frame(
    Axis1 = e$points[, 1], Axis2 = e$points[, 2], group = group,
    panel = sprintf("%s\n(PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2]))
}))
ord$panel <- factor(ord$panel, unique(ord$panel))

ggplot(ord, aes(Axis1, Axis2, colour = group)) +
  geom_point(size = 2, alpha = 0.85) +
  stat_ellipse(type = "t", linewidth = 0.3) +
  facet_wrap(~ panel, nrow = 2, scales = "free") +
  labs(x = "PCo1", y = "PCo2", colour = "group") +
  theme_bw() +
  theme(legend.position = "bottom")
```

In the first three panels group B collapses into a tight cluster away from
group A, and the first axis — a third to a half of the total variation —
carries the separation. In the protein-level Bray-Curtis panel the two groups
sit interleaved, and the leading axes hold a few percent each: the distance is
near-constant across all pairs, within group and between.

# Reading

The three structure-aware routes are not interchangeable in general — MMD
weighs graded similarity through a kernel, UniFrac through shared branch
length, and the bins through a hard threshold — but on a coherent functional
shift they agree, and the conventional baseline is blind to it. When homologs
rather than shared accessions carry a signal, the choice of representation is
not a refinement; it is the difference between detecting the effect and
missing it entirely.

```{r session}
sessionInfo()
```
