---
title: "Full case study 2: homology-aware beta diversity with repdist"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "../..")
```

# Why a representation distance

Bray-Curtis, Jaccard and their relatives compare samples through protein
*identity*: two proteins are either the same accession or maximally different.
Metaproteomes break that assumption. Two communities can carry the same
enzymatic function on accessions that never overlap, and an identity-based
distance cannot tell that case apart from any other difference.

`repdist` replaces identity with distance in a protein language model's
representation space. Each sample becomes an abundance-weighted distribution
over protein embeddings, and two samples are compared by maximum mean
discrepancy (MMD) under an RBF kernel.

This report follows the package vignette's workflow on a marine
plastic-substrate metaproteome:

1. **Beta diversity** with MMD and PERMANOVA.
2. **Structural bins** at a predicted TM-score of 0.7.
3. **Differential abundance** on the bins with DESeq2.
4. **Functional reading** of the most responsive annotated bins.

```{r libs}
library(repdist)
library(ggplot2)
library(phyloseq)
```

# The full dataset

The study contains three independent incubations for each of eight substrates:
four plastics, two PHBH formulations, filter paper and ambient sediment. PLA is
removed before rarefaction because its three runs have exceptionally low and
inconsistent depth (92, 766 and 1,832 counts). The other 21 samples are retained
and rarefied to the shallowest remaining sample.

```{r data}
path <- here::here("data", "case_study2")
MAX_AA <- 3000L

ps <- readRDS(file.path(path, "ps.rds"))
ps <- prune_samples(data.frame(sample_data(ps))$substrate != "PLA", ps)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
sample_data(ps)$substrate <- droplevels(sample_data(ps)$substrate)
floor_depth <- min(sample_sums(ps))

set.seed(1)
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)

n_after_rarefaction <- ntaxa(psd)
protein_length <- Biostrings::width(refseq(psd)[taxa_names(psd)])
too_long <- protein_length > MAX_AA
long_abundance <- sum(taxa_sums(psd)[too_long]) / sum(taxa_sums(psd))
psd <- prune_taxa(!too_long, psd)

cached <- readRDS(file.path(path, "embeddings_tmvec1_large.rds"))$Z
without_embeddings <- setdiff(taxa_names(psd), rownames(cached))
psd <- prune_taxa(taxa_names(psd) %in% rownames(cached), psd)
keep <- taxa_names(psd)
psu <- prune_taxa(taxa_names(ps) %in% keep, ps)

counts <- Matrix::Matrix(t(as(otu_table(psd), "matrix")), sparse = TRUE)
counts_raw <- t(as(otu_table(psu), "matrix"))[rownames(counts), keep, drop = FALSE]

md <- data.frame(sample_data(psd))[rownames(counts), ]
md$substrate <- relevel(factor(md$substrate), "Sediment")

tx <- as.data.frame(as(tax_table(psd), "matrix"))[keep, ]
annotation_cols <- c("MantisName", "MicrobeName", "PlasticDB",
                     "Depolymerase", "Description")
annotation <- apply(tx[, annotation_cols, drop = FALSE], 1, function(x) {
  x <- unique(x[!is.na(x) & nzchar(trimws(x)) & !x %in% c("-", "NA")])
  if (length(x)) x[1] else NA_character_
})

taxa <- data.frame(
  protein = keep,
  genome = tx$GenomeID,
  annotation = annotation,
  plastic_db = tx$PlasticDB,
  depolymerase = tx$Depolymerase,
  length = Biostrings::width(refseq(psd)[keep]),
  row.names = keep, stringsAsFactors = FALSE)

d <- list(counts = counts, counts_raw = counts_raw, metadata = md, taxa = taxa,
          sequences = setNames(as.character(refseq(psd)[keep]), keep))

c(samples = nrow(counts), proteins = ncol(counts),
  rarefaction_depth = floor_depth,
  removed_over_3000aa = sum(too_long),
  removed_abundance_pct = round(100 * long_abundance, 3))
table(md$substrate)
```

Rarefaction provides the abundance filter: proteins absent after subsampling
are removed. It retained `r n_after_rarefaction` proteins; the 3,000aa ceiling
then removed `r sum(too_long)` more, representing only
`r sprintf("%.3f", 100 * long_abundance)`% of rarefied abundance. The beta-diversity
matrix is sparse; `counts_raw` contains unrarefied counts for exactly the same
proteins and samples.

The current source files are reconstructed by
`Rscript inst/scripts/case_study2_annotation.R`. Their abundance values are mean
PUP over technical replicates; the legacy workflow rounds these values before
rarefaction. The original means are preserved in `mean_pup.rds`; rounded means
are not recovered raw spectral counts. Annotation comparisons do not refit DA.
With these files, `r length(without_embeddings)` retained proteins lack cached
embeddings and are excluded consistently from the current distance analysis.
Their IDs and abundances are exported by the reconstruction script. Changes
to the input protein set and its order can change the seeded rarefaction, so
bin numbers in a new render need not identify the same members as older renders.

# Embedding

`embed_proteins()` embeds every sequence whole. The full TM-Vec embeddings are
cached outside the render because model inference is not part of the statistical
analysis.

```{r Z}
stopifnot(all(keep %in% rownames(cached)))
Z <- cached[keep, , drop = FALSE]
dim(Z)
```

TM-Vec's head was trained to regress TM-score, so cosine similarity between two
embeddings is a predicted TM-score. `known_models()` lists the other registered
representations.

```{r models}
names(known_models())
```

# The protein ground distance

`repdist_matrix()` builds the condensed protein-by-protein cosine distance used
by binning and QC. This is the expensive object in the full analysis.

```{r ground}
G_cos <- repdist_matrix(Z)
```

`bin_proteins()` reads this scale as predicted TM-score. `sample_repdist()`
instead builds a Euclidean RBF internally so the kernel remains positive
semi-definite. On unit-norm TM-Vec embeddings the two ground distances are
monotone transforms of one another.

```{r scales}
c(median_predicted_TM = round(1 - median(G_cos), 3),
  pairs_above_0.7 = sum(G_cos <= 0.3))
```

# Step 1: beta diversity

Bray-Curtis is the protein-identity baseline. MMD asks how far apart different
proteins actually are. The sparse count matrix stays sparse throughout the MMD
calculation.

```{r distances}
D <- list(
  `Bray-Curtis` = vegan::vegdist(as.matrix(d$counts), "bray"),
  `MMD` = sample_repdist(d$counts, Z))
```

Both are abundance-weighted, so richness is checked before interpreting the
substrate effect.

```{r qc-richness}
rich <- vegan::specnumber(as.matrix(d$counts))
round(vapply(D, function(x) cor(rich, rowMeans(as.matrix(x))), numeric(1)), 2)
```

## PERMANOVA

The incubations are independent replicates, so permutations are not blocked.

```{r permanova}
res <- do.call(rbind, lapply(names(D), function(n) {
  set.seed(1)
  a <- vegan::adonis2(D[[n]] ~ substrate, data = md,
                      by = "terms", permutations = 999)
  data.frame(distance = n, R2 = round(a["substrate", "R2"], 3),
             F = round(a["substrate", "F"], 2), p = a["substrate", "Pr(>F)"])
}))
knitr::kable(res, row.names = FALSE)
```

## Ordination

```{r pcoa, fig.height = 3.8, fig.width = 7, fig.alt = "PCoA under Bray-Curtis and RBF-MMD, coloured by incubation substrate"}
ord <- do.call(rbind, lapply(names(D), function(n) {
  e <- cmdscale(D[[n]], k = 2, eig = TRUE)
  rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
  data.frame(
    Axis1 = e$points[, 1], Axis2 = e$points[, 2],
    substrate = md[rownames(e$points), "substrate"],
    panel = sprintf("%s  (PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2]))
}))

ggplot(ord, aes(Axis1, Axis2, colour = substrate)) +
  geom_point(size = 2, alpha = 0.85) +
  stat_ellipse(type = "t", linewidth = 0.3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "PCo1", y = "PCo2", colour = "substrate") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Step 2: structural bins at TM-score 0.7

`bin_proteins()` uses complete linkage, so every pair inside a bin meets the
threshold. A 0.7 cut is stricter than the conventional same-fold boundary of
0.5: bins are tighter, at the cost of more singletons.

```{r bins}
b <- bin_proteins(G_cos, min_sim = 0.7)
```

```{r qc-bins, fig.height = 5, fig.alt = "Nearest-neighbour similarity and structural-bin profiles"}
plot_similarity_profile(G_cos, min_sim = 0.7)
plot_bin_profile(b, min_sim = 0.7)
```

The unrarefied counts roll up to bin level with one `rowsum()`.

```{r bin-counts}
bin <- setNames(rep(names(b$clusters), lengths(b$clusters)), unlist(b$clusters))
bin_counts <- t(rowsum(t(d$counts_raw), bin[colnames(d$counts_raw)]))
c(proteins = ncol(d$counts_raw), bins = ncol(bin_counts))
```

```{r bin-sizes}
sizes <- lengths(b$clusters)
table(cut(sizes, c(0, 1, 2, 5, 10, Inf),
          labels = c("1", "2", "3-5", "6-10", ">10")))
```

`r sum(sizes == 1)` proteins remain singleton bins. The other
`r sum(sizes > 1)` bins combine structurally related accessions, reducing the
number of tested units by a factor of
`r round(ncol(d$counts_raw) / ncol(bin_counts), 1)`.

# Step 3: differential abundance with DESeq2

DESeq2 fits a negative-binomial model per bin on unrarefied counts. Substrate is
tested by likelihood ratio against an intercept-only model.

```{r deseq}
dds <- DESeq2::DESeqDataSetFromMatrix(t(bin_counts), md,
                                      design = ~ substrate)
dds <- DESeq2::DESeq(dds, test = "LRT", reduced = ~ 1,
                     sfType = "poscounts", quiet = TRUE)
res_lrt <- DESeq2::results(dds)

c(bins_tested = nrow(res_lrt),
  substrate_significant = sum(res_lrt$padj <= 0.05, na.rm = TRUE))
```

Each significant bin is assigned its largest shrunken substrate-vs-sediment
fold change for ranking and display.

```{r effects}
lev <- setdiff(levels(md$substrate), "Sediment")
coef_names <- setNames(paste0("substrate_", make.names(lev), "_vs_Sediment"), lev)
ddsw <- DESeq2::nbinomWaldTest(dds, quiet = TRUE)
stopifnot(all(coef_names %in% DESeq2::resultsNames(ddsw)))
shrunk <- lapply(coef_names, function(coef)
  DESeq2::lfcShrink(ddsw, coef = coef, type = "normal", quiet = TRUE))
names(shrunk) <- lev
fc <- vapply(shrunk, function(o) o$log2FoldChange, numeric(nrow(res_lrt)))
rownames(fc) <- rownames(res_lrt)

sig <- rownames(res_lrt)[!is.na(res_lrt$padj) & res_lrt$padj <= 0.05]
best <- apply(abs(fc[sig, , drop = FALSE]), 1, which.max)
top <- data.frame(
  bin = sig,
  members = sizes[sig],
  padj = formatC(res_lrt[sig, "padj"], digits = 2, format = "e"),
  substrate = lev[best],
  log2FC = round(fc[cbind(sig, lev[best])], 2))
top <- top[order(-abs(top$log2FC)), ]
knitr::kable(head(top, 8), row.names = FALSE)
```

```{r volcano, fig.height = 4.6, fig.width = 9, fig.alt = "Substrate-vs-sediment volcano plots for structural bins"}
volc <- do.call(rbind, lapply(lev, function(l) {
  o <- shrunk[[l]]
  data.frame(bin = rownames(o), substrate = l,
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue,
             reject = !is.na(o$padj) & o$padj <= 0.05)
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ substrate, nrow = 2) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                     guide = "none") +
  labs(x = "log2 fold change vs sediment", y = expression(-log[10](p))) +
  theme_bw()
```

# Step 4: responsive bins, read functionally

Select the three most responsive multi-member bins with at least two annotated
proteins.

```{r pick}
n_ann <- vapply(b$clusters[top$bin],
                function(m) sum(!is.na(d$taxa[m, "annotation"])), integer(1))
picks <- head(top[top$members > 1 & n_ann >= 2, ], 3)
knitr::kable(picks, row.names = FALSE)
```

```{r members}
info <- do.call(rbind, lapply(seq_len(nrow(picks)), function(i) {
  m <- b$clusters[[picks$bin[i]]]
  data.frame(
    bin = picks$bin[i],
    genome = d$taxa[m, "genome"],
    annotation = substr(d$taxa[m, "annotation"], 1, 52),
    PlasticDB = d$taxa[m, "plastic_db"],
    depolymerase = d$taxa[m, "depolymerase"],
    mean_Sediment = round(colMeans(d$counts_raw[md$substrate == "Sediment", m,
                                                  drop = FALSE]), 1),
    mean_top = round(colMeans(d$counts_raw[md$substrate == picks$substrate[i], m,
                                             drop = FALSE]), 1))
}))
knitr::kable(info, row.names = FALSE)
```

The complete-linkage guarantee can be inspected directly for the leading
annotated bin.

```{r floor, fig.height = 4.5, fig.alt = "Pairwise predicted TM-score heatmap for the leading annotated bin"}
plot_bin_similarity(b, picks$bin[1], G_cos)
```

# Annotation-focused clustering (optional comparison)

The TM70 bins above remain the units for differential abundance. Their
all-pairs structural similarity bound does not imply shared biochemical
function. For annotation, compare alternative partitions of **all retained
proteins together**: clustering each TM70 bin independently could never move
a misplaced protein to a sibling TM70 bin in the same TM50 group.

`bin_proteins()` now offers complete linkage only: MCL, DBSCAN*/DBSCAN,
HDBSCAN and an emergent SOM were scored against it on an EC benchmark and none
improved annotation purity at a matched bin count, while all gave up the
all-pairs similarity floor. The optional chunk below therefore sweeps the
threshold instead of the method, reusing `G_cos`, `keep` and `annotation` from
this report; it leaves `b`, bin counts and the DA model unchanged.

The completed method comparison on the current source files is in
`case_study2_annotation_report.html`, generated by
`case_study2_annotation.R` and `case_study2_annotation_report.Rmd` against the
earlier multi-method API. It scores KO annotations separately by source and
includes audited protein reassignment examples.

```{r annotation-comparison, eval = FALSE}
source("inst/scripts/annotation_clustering.R")

# Text agreement is only a screening proxy; synonyms can disagree and broad
# descriptions can agree without establishing the same molecular function.
labels <- setNames(tolower(trimws(annotation)), keep)
generic <- grepl("^(hypothetical|uncharacteri[sz]ed|unknown|predicted) protein$",
                  labels)
labels[generic] <- NA_character_
comparison <- compare_annotation_clusters(G_cos, labels, c(.5, .7, .9))
knitr::kable(comparison$summary, digits = 3, row.names = FALSE)

out <- file.path("results", "case_study2_annotation_clustering")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
write.csv(comparison$summary, file.path(out, "summary.csv"), row.names = FALSE)
write.csv(comparison$membership, file.path(out, "membership.csv"),
           row.names = FALSE)
```

`membership.csv` links every accession to every candidate bin, including the
TM50 and TM70 baselines (`run1` and `run2`), and marks density noise separately.
This lets collaborators trace splits and movement between sibling bins without
truncating accession IDs or annotation text. Unknown annotations are excluded
from agreement scores, while their proteins remain in the clustering. Purity
is scored only in bins with at least two labelled proteins; coverage, singleton
fraction and pair recall show the cost of splitting. Different text labels
are candidates for review, not confirmed functional conflicts. Curated EC/KO
labels should be evaluated separately at a consistent level of specificity
before propagating annotations.

# Practical notes

**Rarefy first** for whole-community distances. The differential-abundance
model uses the corresponding unrarefied counts and handles depth through size
factors.

**Memory.** `repdist_matrix()` remains quadratic because binning needs the
condensed protein distance. `sample_repdist()` does not: it accumulates
two-dimensional 1000-by-1000 kernel tiles directly into the sample Gram matrix,
and sparse counts remain sparse.

**Bandwidth.** `sample_repdist()` returns its bandwidth as `attr(, "sigma")`;
pass it back when comparisons across studies need one fixed kernel.

```{r session}
sessionInfo()
```
