---
title: "Simulation benchmark: a coherent functional shift on non-overlapping accessions"
author: "Yixuan Yang"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
```

The methodological test of `repdist`: a functional shift with a known truth,
carried by accessions no two samples share. The
[simulation page](../simulation.html) works one such design through in full on
ten CATH FunFams; this page repeats the test on experimentally characterised
enzyme classes, with every eligible EC3 as the target.

```{r libs}
library(ggplot2)
bm <- readRDS("simulation_benchmark.rds")
u <- readRDS("microbial_ec_universe.rds")
```

# Design

- **Universe.** `r nrow(u$members)` bacterial and archaeal UniProtKB enzymes
  (release `r u$provenance$uniprot_release`) whose EC number sits on a
  catalytic-activity reaction with experimental evidence (ECO:0000269), one
  EC3 each, one per UniRef50 cluster; every EC3 with at least 50 such
  proteins, 50 of each (`r length(unique(u$members$ec3))` EC3s). Embedded
  with TM-Vec. `README.md` beside this page gives every selection rule.
- **Groups.** Ten samples each. Every group A sample upregulates 5 private
  proteins of the target EC3; every group B sample 5 private proteins from
  distinct other EC3s. No accession is upregulated twice, every protein is
  in every sample, and group B's k-th sample takes the baseline abundances of
  group A's k-th, so the groups move the same amount of abundance.
- **Counts.** Negative binomial: lognormal baseline abundance, lognormal size
  factors, a DESeq2-style mean–dispersion trend with scatter, and a
  `r bm$settings$fold`-fold mean increase on upregulated proteins, which
  makes the target EC3's group-mean aggregate exactly 4 times baseline.
- **Methods.** Protein-level Bray-Curtis; RBF-MMD (`sample_mmd()`);
  Bray-Curtis on MCL bins (`protein_similarity()` with `min_sim`
  `r bm$settings$min_sim` and no coverage gate, then `bin_proteins()` with
  `inflation` `r bm$settings$inflation`); DESeq2 at
  protein and bin level. Summing by true EC3 is the oracle.
- **Repeats.** Every EC3 as the target, one seed each:
  `r nrow(bm$scores)` simulations.

The clustering never sees the EC3 labels; they are used only to score.

# Sample-level separation

```{r separation}
s <- bm$separation
tab <- do.call(rbind, lapply(split(s, s$distance), function(x) {
  data.frame(distance = x$distance[1], detected = mean(x$p < 0.05),
             median_R2 = median(x$R2),
             dispersion_differs = mean(x$permdisp_p < 0.05))
}))
tab <- tab[order(-tab$median_R2), ]
knitr::kable(tab, row.names = FALSE, digits = 3)
```

`detected`: share of targets with PERMANOVA $p < 0.05$ (999 permutations).
`dispersion_differs`: share with PERMDISP $p < 0.05$, where part of a
PERMANOVA difference can be spread rather than location. $R^2$ is unchanged
by multiplying a distance by a constant, so the rows can be compared
descriptively; the distances encode different geometries, so read gaps
between them with that in mind.

```{r separation-figure, fig.height = 3.4, fig.alt = "PERMANOVA R2 between groups for every target EC3 under each distance"}
s$distance <- factor(s$distance, rev(tab$distance))
ggplot(s, aes(R2, distance)) +
  geom_boxplot(outlier.shape = NA, colour = "grey40") +
  geom_jitter(height = 0.15, size = 1.3, alpha = 0.7) +
  labs(x = expression(PERMANOVA ~ R^2 * ", group A against B"), y = NULL) +
  theme_bw()
```

Separation follows the target's structural coherence, the median predicted
TM-score between its own members (Spearman correlation with $R^2$):

```{r coherence}
Z <- u$embeddings / sqrt(rowSums(u$embeddings^2))
ec3 <- setNames(u$members$ec3, u$members$protein)
coherence <- vapply(split(names(ec3), ec3), function(ids) {
  S <- tcrossprod(Z[ids, ])
  median(S[upper.tri(S)])
}, numeric(1))
round(vapply(split(bm$separation, bm$separation$distance), function(x)
  cor(x$R2, coherence[x$target], method = "spearman"), numeric(1)), 2)
```

# Differential abundance

```{r da}
sc <- bm$scores
knitr::kable(data.frame(
  measure = c("protein level: target proteins significant (total of 1,400)",
              "protein level: any protein significant (mean per target)",
              "MCL bins: targets with a significant target-dominated bin",
              "MCL bins: recall of the upregulated target proteins (mean)",
              "MCL bins: significant bins higher in group B (total)",
              "EC3 oracle: targets significant"),
  value = c(sum(sc$protein_sig_target), mean(sc$protein_sig),
            sum(sc$bin_sig_up_a_target > 0), mean(sc$bin_recall),
            sum(sc$bin_sig_up_b), sum(sc$oracle_padj < 0.05))
), digits = 3)
```

Per target, sorted by MCL-bin recall. `target_completeness` is the share of
the target EC3 in its largest bin; `target_bins` the number of bins it spans.

```{r per-target}
sc$coherence <- round(coherence[sc$target], 3)
knitr::kable(sc[order(-sc$bin_recall),
                c("target", "coherence", "target_bins", "target_completeness",
                  "bin_recall", "bin_sig_up_a_target", "target_bin_padj",
                  "target_bin_rank")],
             row.names = FALSE, digits = 3)
```

# Sensitivity to the edge floor and inflation

Averaged over targets and scored with the labels. The reference setting was
fixed before this ran; this table shows how the parameters move the result
and is not used to choose them.

```{r sensitivity}
sens <- aggregate(
  cbind(bins, in_multi_bins, ec3_purity, target_completeness, bin_recall,
        recovered = bin_sig_up_a_target > 0, bin_sig_up_b) ~
    min_sim + inflation,
  data = bm$sensitivity, FUN = mean)
knitr::kable(sens[order(sens$min_sim, sens$inflation), ],
             row.names = FALSE, digits = 2)
```

`r sum(!is.na(bm$sensitivity$error))` of `r nrow(bm$sensitivity)` grid fits
failed in DESeq2 and are left out of the means.

# Reading

- A coherent shift on private accessions is invisible to identity at the
  sample level: protein-level Bray-Curtis detects it at about its
  false-positive rate. At protein level, DESeq2 finds a share of the
  upregulated proteins one at a time, at a 31-fold change on a single sample
  each.
- RBF-MMD detects the shift for nearly every target, and more strongly the
  more structurally coherent the target EC3. An EC3 whose members span
  several folds gives it least to read.
- MCL bins make part of the shift testable as units. At the reference floor
  of 0.5 the target EC3 spreads over several bins; higher floors and
  inflations give purer, finer bins and recover more of the target, at the
  cost of more significant bins in group B; a floor of 0.4 collapses
  everything into a few bins.
- A bin is a community in a predicted-TM similarity graph, not a functional
  family; EC3 is the truth only because this is a simulation.

# Reproducing

Run from the `gh-pages` branch root with the current `repdist` installed:
`Rscript validation/benchmark_simulation.R` regenerates
`simulation_benchmark.rds` (about 5 minutes), and
`validation/prepare_microbial_ec_universe.R` rebuilds the universe from
UniProt. The session that produced these results:

```{r session}
cat(head(bm$session, 3), sep = "\n")
```
