---
title: "Simulation benchmark: a coherent functional shift on non-overlapping accessions"
author: "Yixuan Yang"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
```

The methodological test of `repdist`: a functional shift with a known truth,
carried by accessions no two samples share. This page summarises the
benchmark over every eligible target; the worked single-target example, with
all the code, is `vignette("simulation", package = "repdist")`.

```{r libs}
library(repdist)
library(ggplot2)
bm <- readRDS(system.file("extdata", "simulation_benchmark.rds",
                          package = "repdist"))
u <- readRDS(system.file("extdata", "microbial_ec_universe.rds",
                         package = "repdist"))
```

# Design

- **Universe.** `r nrow(u$members)` bacterial and archaeal UniProtKB enzymes
  (release `r u$provenance$uniprot_release`) whose EC number sits on a
  catalytic-activity reaction with experimental evidence (ECO:0000269), one
  EC3 each, one per UniRef50 cluster; every EC3 with at least 50 such
  proteins, 50 of each (`r length(unique(u$members$ec3))` EC3s). Embedded
  with TM-Vec.
- **Groups.** Ten samples each. Every group A sample upregulates 5 private
  proteins of the target EC3; every group B sample 5 private proteins from
  distinct other EC3s. No accession is upregulated twice, every protein is
  in every sample, and group B's k-th sample takes the baseline abundances of
  group A's k-th, so the groups move the same amount of abundance.
- **Counts.** Negative binomial: lognormal baseline abundance, lognormal size
  factors, a DESeq2-style mean–dispersion trend with scatter, 8-fold mean
  increase on upregulated proteins.
- **Methods.** Protein-level Bray-Curtis; RBF-MMD (`sample_repdist()`);
  exact Wasserstein-1 with Euclidean embedding cost
  (`sample_repdist_ot()`); Bray-Curtis on MCL bins (`bin_proteins()` at the
  defaults, `min_edge_sim = 0.5`, `inflation = 2`); DESeq2 at protein and bin
  level. Summing by true EC3 is the oracle.
- **Repeats.** Every EC3 as the target, one seed each:
  `r nrow(bm$scores)` simulations.

The clustering never sees the EC3 labels; they are used only to score.

# Sample-level separation

```{r separation}
s <- bm$separation
tab <- do.call(rbind, lapply(split(s, s$distance), function(x) {
  data.frame(distance = x$distance[1], detected = mean(x$p < 0.05),
             median_R2 = median(x$R2), median_ANOSIM_R = median(x$anosim_R))
}))
tab <- tab[order(-tab$median_ANOSIM_R), ]
knitr::kable(tab, row.names = FALSE, digits = 3)
```

`detected`: share of targets with PERMANOVA $p < 0.05$ (999 permutations).
$R^2$ is unchanged by multiplying a distance by a constant, so the rows can
be compared descriptively; the distances encode different geometries, so
read gaps between them with that in mind. ANOSIM $R$ uses ranks alone.

```{r separation-figure, fig.height = 3.4, fig.alt = "ANOSIM R between groups for every target EC3 under each distance"}
s$distance <- factor(s$distance, rev(tab$distance))
ggplot(s, aes(anosim_R, distance)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_boxplot(outlier.shape = NA, colour = "grey40") +
  geom_jitter(height = 0.15, size = 1.3, alpha = 0.7) +
  labs(x = "ANOSIM R, group A against B", y = NULL) +
  theme_bw()
```

Separation follows the target's structural coherence, the median predicted
TM-score between its own members:

```{r coherence}
Z <- u$embeddings / sqrt(rowSums(u$embeddings^2))
ec3 <- setNames(u$members$ec3, u$members$protein)
coherence <- vapply(split(names(ec3), ec3), function(ids) {
  S <- tcrossprod(Z[ids, ])
  median(S[upper.tri(S)])
}, numeric(1))
round(vapply(split(bm$separation, bm$separation$distance), function(x)
  cor(x$anosim_R, coherence[x$target], method = "spearman"), numeric(1)), 2)
```

# Differential abundance

```{r da}
sc <- bm$scores
knitr::kable(data.frame(
  measure = c("protein level: target proteins significant (total)",
              "protein level: any protein significant (mean per target)",
              "MCL bins: targets with a significant target-dominated bin",
              "MCL bins: recall of the upregulated target proteins (mean)",
              "MCL bins: significant bins higher in group B (total)",
              "EC3 oracle: targets significant"),
  value = c(sum(sc$protein_sig_target), mean(sc$protein_sig),
            sum(sc$bin_sig_up_a_target > 0), mean(sc$bin_recall),
            sum(sc$bin_sig_up_b), sum(sc$oracle_padj < 0.05))
), digits = 3)
```

Per target, sorted by MCL-bin recall. `completeness` is the share of the
target EC3 in its largest bin; `target_bins` the number of bins it spans.

```{r per-target}
sc$coherence <- round(coherence[sc$target], 3)
knitr::kable(sc[order(-sc$bin_recall),
                c("target", "coherence", "target_bins", "target_completeness",
                  "bin_recall", "bin_sig_up_a_target", "target_bin_padj",
                  "target_bin_rank")],
             row.names = FALSE, digits = 3)
```

# Sensitivity to the edge floor and inflation

Averaged over targets and scored with the labels. The defaults were fixed
before this ran; this table shows how the parameters move the result and is
not used to choose them.

```{r sensitivity}
sens <- aggregate(
  cbind(bins, in_multi_bins, ec3_purity, target_completeness, bin_recall,
        recovered = bin_sig_up_a_target > 0, bin_sig_up_b) ~
    min_edge_sim + inflation,
  data = bm$sensitivity, FUN = mean)
knitr::kable(sens[order(sens$min_edge_sim, sens$inflation), ],
             row.names = FALSE, digits = 2)
```

# Reading

- A coherent shift on private accessions is invisible to identity:
  protein-level Bray-Curtis detects it at its false-positive rate, and no
  target protein is significant at protein level.
- The representation-aware distances detect it for about half to three
  quarters of targets, more strongly the more structurally coherent the
  target EC3. An EC3 whose members span several folds gives them nothing to
  read.
- MCL bins make part of the shift testable. At the fold-scale default floor
  the graph is one giant component and communities mix EC3s that share a
  fold; the sensitivity table shows purer, finer bins at higher floors or
  inflations, and a floor of 0.4 collapsing everything into a few bins.
- A bin is a community in a predicted-TM similarity graph, not a functional
  family; EC3 is the truth only because this is a simulation.

# Reproducing

`inst/scripts/benchmark_simulation.R` in the package source regenerates the
results (about 15 minutes, mostly exact optimal transport);
`inst/scripts/prepare_simulation_data.R` rebuilds the universe from UniProt.
The session that produced the shipped results:

```{r session}
cat(head(bm$session, 3), sep = "\n")
```
