---
title: "Re-analysis of a marine plastic-substrate metaproteome"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE, autodep = TRUE,
                      fig.width = 8, fig.height = 4, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

This is a second, independent test of the same comparison `case_study.Rmd` ran on the
diet metaproteome: does a TM-Vec ground metric (`sample_repdist()`) find sample-level
structure that protein-identity distances (Bray-Curtis, Jaccard) miss or dilute, and
does collapsing proteins into structural bins (`repdist_bin()`) surface a
substrate-associated bin that a per-protein test would drown in multiple testing? The
dataset here is a marine plastic-incubation metaproteome: 24 samples as collected,
eight substrates, three replicate incubations each -- four plastics (PHBV, PLA, HDPE,
a PLA/PCL blend), two PHBH formulations ("PHBH" and "PHBH Green"), filter paper (a
fast-degrading cellulose control), and sediment (the ambient community, no added
substrate). One substrate, PLA, is dropped for depth reasons below, leaving seven.

It differs from the diet study in one way that matters throughout: replicates are
independent incubations, not repeated samples of the same subject, so there is no
blocking factor in the permutation tests below.

```{r libs, cache = FALSE}
library(phyloseq)
library(vegan)
library(ggplot2)
library(gt)
library(here)
library(MicrobiomeStat)
devtools::load_all(".", quiet = TRUE)

path  <- here("data", "case_study2")
NPERM <- 9999
```

## Data

```{r data}
ps <- readRDS(file.path(path, "ps.rds"))   # built and validated in data_curation_case_study2.Rmd
ps
gt(as.data.frame(table(substrate = data.frame(sample_data(ps))$substrate)))
```

## Rarefaction

Depth here is nothing like the diet study's: that report's shallowest samples sat
within a few percent of each other, so the rarefaction floor was just `min()`. Here
total counts span roughly 90 to 25,700 across the 24 samples, close to 300-fold.

```{r depth-check, fig.height = 3.5}
depth <- data.frame(sample = sample_names(ps), depth = sample_sums(ps),
                    substrate = data.frame(sample_data(ps))$substrate)
depth <- depth[order(depth$depth), ]
depth$sample <- factor(depth$sample, levels = depth$sample)

ggplot(depth, aes(sample, depth, fill = substrate)) +
  geom_col() +
  labs(x = "sample, sorted by depth", y = "total counts (rounded PUP)") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

`PLA`'s three replicates were each searched against a similarly large protein
catalog (`r paste(sort(colSums(!is.na(read.csv(file.path(path, "ProteomeMeanPUP.csv"), row.names = 1, check.names = FALSE)[, c("B1","B2","B3")]))), collapse = ", ")`
proteins offered, B1/B2/B3) -- so this is not a case of one replicate having less
evidence available to it. What came back differs enormously within that shared
search space: `r paste(sort(colSums(read.csv(file.path(path, "ProteomeMeanPUP.csv"), row.names = 1, check.names = FALSE)[, c("B1","B2","B3")] > 0, na.rm = TRUE)), collapse = ", ")`
proteins actually detected, an 18-fold range across three replicates of the same
substrate. The low end of that range is a failed or badly degraded run, not
ordinary replicate noise, and it is not obvious from depth alone which of the
other two are trustworthy either. Rather than keep the substrate at reduced,
uncertain replication, or let its worst replicate set the floor for all 21 other
samples, the whole substrate is dropped.

```{r drop-pla}
ps <- prune_samples(data.frame(sample_data(ps))$substrate != "PLA", ps)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
sample_data(ps)$substrate <- droplevels(sample_data(ps)$substrate)
floor_depth <- min(sample_sums(ps))    # rarefy to the shallowest remaining sample, dropping none of the rest
c(samples_remaining = nsamples(ps), floor_depth = floor_depth)
```

The floor (`r floor_depth`) now comes from `S2`, a sediment replicate -- shallow,
but not an isolated outlier the way `PLA` was: the next five samples after it sit
within a similar few-thousand-count range, so there is no second cut to make.

```{r rarefy}
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)
md  <- data.frame(sample_data(psd))
psd
```

## Embedding

TM-Vec1, same model and reasoning as `case_study.Rmd`: its head regresses the TM-score
of protein pairs, so cosine similarity between two vectors is a predicted TM-score. The
full sequence is embedded with memory-aware batching. The cache is keyed by exact
sequence and embedding-policy version, so an old truncated vector, a vector from a
re-curated sequence, or a vector from a different model cannot be silently reused.

```{r embed, cache = FALSE}
model_tag <- paste0(eval(formals(embed_proteins)$model), "-full-length-v1")
cache <- file.path(path, "embeddings_tmvec1_large.rds")
cached <- if (file.exists(cache)) readRDS(cache) else list(Z = NULL, seq = character(0), model = character(0))

# keyed off ps (unrarefied, PLA already dropped), not psd -- covers every protein
# either count matrix below could need, rarefied or not
full_seq <- stats::setNames(as.character(refseq(ps)[taxa_names(ps)]), taxa_names(ps))

reusable <- intersect(taxa_names(ps), rownames(cached$Z))
reusable <- reusable[cached$seq[reusable] == full_seq[reusable] &
                      cached$model[reusable] == model_tag]
need  <- setdiff(taxa_names(ps), reusable)
Zall  <- cached$Z

if (length(need)) {
  s <- full_seq[need]
  s <- s[order(nchar(s))]
  for (i in split(seq_along(s), ceiling(seq_along(s) / 2000))) {
    new_Z <- embed_proteins(s[i])
    Zall  <- rbind(Zall[setdiff(rownames(Zall), rownames(new_Z)), , drop = FALSE], new_Z)
    cached$Z <- Zall
    cached$seq[rownames(new_Z)]   <- full_seq[rownames(new_Z)]
    cached$model[rownames(new_Z)] <- model_tag
    saveRDS(cached, cache)
  }
}
Z      <- Zall[taxa_names(ps), ]
cnt    <- t(as(otu_table(psd), "matrix"))   # samples in rows, RAREFIED -- distances/PCoA only
cnt_da <- t(as(otu_table(ps), "matrix"))    # samples in rows, UNRAREFIED -- differential abundance only
```

```{r length-abundance, fig.height = 4}
length_abundance_profile(cnt, full_seq, max_length = 1000) +
  ggplot2::geom_vline(xintercept = 1300, linetype = "dotted", color = "grey30")
```

At the 1000aa line (dashed), `r round(100*mean(nchar(full_seq[colnames(cnt)]) > 1000), 1)`%
of the catalog by count sits beyond the head's training range, and it carries
`r round(100*sum((cnt/rowSums(cnt))[, nchar(full_seq[colnames(cnt)]) > 1000])/nrow(cnt), 1)`%
of total abundance -- unlike the diet study, abundance here does **not** run ahead of
protein count past the boundary, so there is no specific large-protein-family enrichment
to call out. Full sequences are retained, but the TM-Vec predictions remain
extrapolations beyond this line; `1300` is marked for reference and no bin-level
analysis in this report depends on it.

## Accession sharing

Bray-Curtis and Jaccard can only track substrate through accessions two samples
literally share. Unlike the diet study -- one mouse colony searched against one shared
protein database, where accession sharing was close to universal -- every sample here
comes from its own per-sample metagenome assembly, so an accession is only ever shared
when the same protein happened to assemble and get called independently in more than
one sample.

```{r accession-sharing}
pres <- cnt > 0
n_shared <- colSums(pres)
n_substrates <- vapply(seq_len(ncol(cnt)), function(j) length(unique(md$substrate[pres[, j]])), integer(1))

gt(data.frame(
  quantity = c("proteins detected in exactly 1 sample",
              "proteins detected in >1 sample",
              "proteins detected in >1 substrate"),
  n = c(sum(n_shared == 1), sum(n_shared > 1), sum(n_substrates > 1)),
  pct = round(100 * c(mean(n_shared == 1), mean(n_shared > 1), mean(n_substrates > 1)), 1)))
```

So this dataset sits between the diet study's two extremes: not the near-universal
accession sharing that let Bray-Curtis and Jaccard match TM-Vec1's PERMANOVA there, and
not the fully private-accession regime the diet study's shuffle test constructed
artificially. A meaningful minority of the catalog is shared across substrates, so
protein-identity distances have real signal available to them here, not none.

## Results

```{r distances}
# repdist does not implement UniFrac. It fits the tree; phyloseq computes the
# distance. Raw (normalized = FALSE) is the form that equals W1 under the tree's
# path metric -- the normalized form divides by a scale factor and is no longer a
# transport cost.
#
# Both trees come from analysis/build_trees.R case_study2, and both from the same
# complete-linkage fit, so the sequence-vs-structure contrast is
# about the representation, not the tree-fitting algorithm.
tr <- readRDS(file.path(path, "trees_kmer_struct.rds"))
stopifnot(setequal(tr$proteins, colnames(cnt)))
tree_fidelity <- tr$fidelity   # c(sequence = ..., structure = ...)

# "store the data as another phyloseq object" -- same counts, one tree each
psd_seq <- phyloseq(otu_table(psd), sample_data(psd), phy_tree(tr$tree_seq))
psd_str <- phyloseq(otu_table(psd), sample_data(psd), phy_tree(tr$tree_str))

G_tm <- seq_repdist(Z[colnames(cnt), , drop = FALSE])

D <- list(`Bray-Curtis`           = phyloseq::distance(psd, method = "bray"),
          `Jaccard`               = vegdist(cnt, method = "jaccard", binary = TRUE),
          `w-UniFrac (sequence)`  = UniFrac(psd_seq, weighted = TRUE, normalized = FALSE),
          `w-UniFrac (structure)` = UniFrac(psd_str, weighted = TRUE, normalized = FALSE))
D <- lapply(D, function(d) as.dist(as.matrix(d)[rownames(md), rownames(md)]))
```

No nuisance covariates and no permutation blocking: each sample is an independent
incubation replicate, not a repeated measurement of the same unit.

```{r permanova}
substrate_test <- function(dl, nperm = NPERM) do.call(rbind, lapply(names(dl), function(n) {
  set.seed(1)
  a <- adonis2(dl[[n]] ~ substrate, data = md, by = "terms", permutations = nperm)
  data.frame(distance = n, R2 = round(a["substrate", "R2"], 3),
             F = round(a["substrate", "F"], 2), p = a["substrate", "Pr(>F)"])
}))
res <- substrate_test(D)
r2  <- setNames(res$R2, res$distance)
gt(res)
```

```{r pcoa, cache = FALSE, fig.height = 4.5}
for (n in names(D)) {
  print(plot_ordination(psd, ordinate(psd, "PCoA", distance = D[[n]]), color = "substrate") +
          stat_ellipse(type = "t", geom = "polygon", alpha = 0) +
          ggtitle(n) + theme_bw())
}
```

## Diagnostics

```{r diagnostics}
gt(data.frame(
  distance     = names(D),
  betadisper_p = sapply(D, function(d) {
    bd <- betadisper(d, md$substrate, bias.adjust = TRUE)
    round(permutest(bd, permutations = NPERM)$tab$`Pr(>F)`[1], 4)
  }),
  row.names = NULL))
```

## Differential abundance

Unrarefied counts (`cnt_da`, from `ps`), not the rarefied `cnt` the sections above use.
Rarefaction earns its keep for a sample-level distance -- Bray-Curtis and `sample_repdist`
need every sample on the same footing, and subsampling is the standard way to buy that.
A per-bin model doesn't have the same need: LinDA fits its own bias-corrected
centered-log-ratio transform per sample, which already absorbs each sample's own depth,
so throwing away real counts on top of that would trade statistical power for nothing
(McMurdie & Holmes, *PLoS Comp Biol* 2014, arguing rarefaction is broadly the wrong
tool for differential-abundance testing specifically, as opposed to beta-diversity). At
this dataset's aggressive floor (`r floor_depth`, from dropping `PLA`) that power is
real: unrarefied keeps `r ncol(cnt_da)` proteins against `r ncol(cnt)` rarefied, and every
member below still gets a fair shot at clearing LinDA's own zero-handling rather than
being subsampled to zero before the model ever sees it.

PERMANOVA says substrate structures the community overall; it says nothing about which
proteins carry that signal. Testing all `r ncol(cnt_da)` proteins directly would run
that many near-independent tests on units that mostly are not independent -- proteins
with similar structure are frequently near-homologous, so a chunk of those "tests"
repeat the same signal. `repdist_bin()` collapses proteins into bins where every
member has predicted TM-score above a threshold to every other member; both the
conventional TM-score same-fold boundary (0.5, Xu & Zhang 2010, used to pick the
embedding model above) and a stricter cutoff (0.7, tighter and more homogeneous bins
at the cost of more of them) are run below, as two independent passes over the same
counts and embeddings -- multiplicity is paid once per structural unit within each
pass, not once per accession, but the two passes are not compared to each other
statistically, only side by side.

```{r rollup-both}
roll_both <- repdist_bin(cnt_da, Z, cluster_threshold = c(0.5, 0.7))$bins
```

Passing both thresholds at once builds the pairwise TM-Vec distance matrix and its
linkage tree once and cuts it at both heights, rather than paying that cost (the slow
part; the cut itself is nearly free) twice over for two separate
`repdist_bin()` calls.

```{r shared-setup}
is_annotated <- function(t) !is.na(t$MantisName) | !is.na(t$MicrobeName) |
  (!is.na(t$PlasticDB) & t$PlasticDB != "-") | !is.na(t$Depolymerase)
tax_all <- as.data.frame(as(tax_table(ps), "matrix"))
flagged <- stats::setNames(
  !is.na(tax_all$PlasticDB) & tax_all$PlasticDB != "-" | !is.na(tax_all$Depolymerase),
  rownames(tax_all))
```

### TM-score >= 0.5

```{r rollup-05}
roll_05 <- roll_both[["0.5"]]
c(proteins = ncol(cnt_da), bins = ncol(roll_05$counts))
```

LinDA (Zhou *et al.*, *Genome Biology* 2022) fits a linear model per bin on
centered-log-ratio-transformed abundance and corrects the compositional bias a plain
`lm()` on relative abundance would carry -- the same reason `sample_repdist()` works on
abundance weights rather than raw proportions directly. `feature.dat.type = "count"`
since `roll_05$counts` is still integer PUP counts, just not rarefied ones -- the same
choice `case_study.Rmd` makes on its own unrarefied PSM counts below. No random effect
in the formula, unlike the diet study's `(1 | subject)` -- these are independent
incubation replicates, not repeated samples of the same unit.

`r sum(colSums(roll_05$counts > 0) < 3)` of `r ncol(roll_05$counts)` bins here are
nonzero in fewer than 3 of `r nrow(roll_05$counts)` samples, LinDA's own stated bar for
"virtually no statistical power" -- `prev.filter` drops them before fitting rather than
after, standard practice independent of anything specific to this bin count: at
`r ncol(roll_05$counts)` bins this fit is fast and small regardless of whether they are
filtered (contrast the stricter threshold below, where the same filter is also what
keeps the fit from running out of memory).

```{r linda-05}
fit_05 <- linda(t(roll_05$counts), md, formula = "~ substrate",
               feature.dat.type = "count", prev.filter = 3 / nrow(roll_05$counts), verbose = FALSE)
```

`substrate` has `r nlevels(md$substrate)` levels (sediment as the reference), so a
single coefficient is not "the" substrate effect any more than it was for the diet
study's five-level diet term. The Wald test below is the same joint test on all
`r nlevels(md$substrate) - 1` substrate coefficients per bin that `adonis2`'s
`by = "terms"` gives at the sample level, now run per structural bin.

```{r linda-wald-05}
k <- nlevels(md$substrate) - 1
L <- matrix(0, nrow = k, ncol = length(fit_05$variables) + 1)
for (i in 1:k) L[i, i + 1] <- 1   # intercept + the k substrate dummy columns
wald_05 <- linda.wald.test(fit_05, L, model = "LM", alpha = 0.05)
rownames(wald_05) <- rownames(fit_05$output[[1]])   # linda.wald.test drops them; reused below

gt(data.frame(bins_tested = nrow(wald_05), bins_substrate_sig = sum(wald_05$reject),
              fraction_sig = round(mean(wald_05$reject), 3)))
```

`r round(100 * mean(wald_05$reject))`% of bins reject at `padj <= 0.05`. One log2 fold
change per point needs one reference level; the `r nlevels(md$substrate) - 1`
substrate-vs-Sediment contrasts already inside `fit_05` are shown separately, one panel
per non-reference substrate, red where that specific contrast rejects at
`padj <= 0.05`.

```{r volcano-05, fig.height = 3.2, fig.width = 10}
substrate_vars_05 <- grep("^substrate", fit_05$variables, value = TRUE)
volc_05 <- do.call(rbind, lapply(substrate_vars_05, function(v) {
  o <- fit_05$output[[v]]
  data.frame(bin = rownames(o), substrate = sub("^substrate", "", v),
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue, reject = o$reject)
}))

ggplot(volc_05, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ substrate, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"), guide = "none") +
  labs(x = "log2 fold change vs Sediment", y = expression(-log[10](p))) + theme_bw()
```

```{r top-bin-05}
top_05 <- volc_05[volc_05$reject, ]
top_05 <- top_05[which.max(abs(top_05$log2FoldChange)), ]

members_05 <- names(roll_05$bin)[roll_05$bin == top_05$bin]
tax_05 <- as.data.frame(as(tax_table(ps), "matrix"))[members_05, ]
n_annotated_05    <- sum(is_annotated(tax_05))
n_multi_genome_05 <- sum(!is.na(tax_05$SharedGenomes))
```

The single largest shift, in either direction, is bin `r top_05$bin` on
`r top_05$substrate` vs. sediment: `r sprintf("%+.1f", top_05$log2FoldChange)` log2
fold change at $p =$ `r formatC(top_05$pvalue, digits = 1, format = "e")`. Its
`r length(members_05)` members carry no shared lineage (this workbook has no GTDB
call, `data_curation_case_study2.Rmd`), and `r n_annotated_05` of `r length(members_05)`
carries any curated call at all (`MantisName`, `MicrobeName`, a `PlasticDB` hit, or a
depolymerase flag) -- the table below has the specifics. `repdist_bin()`
grouped them by predicted fold alone, with no access to any of that either way,
including which `GenomeID` (MAG) a member's own sample assembly placed it in -- not to
be confused with this bin itself, a structural cluster; the two "bin" concepts are
unrelated, see `data_curation_case_study2.Rmd` § One genome per protein.
`r n_multi_genome_05` of the `r length(members_05)` members were independently
assembled and binned in more than one sample's `GenomeID` (the `SharedGenomes` column
below).

```{r top-bin-table-05}
info_05 <- data.frame(
  Protein = members_05,
  `Genome ID` = tax_05$GenomeID,
  `Shared genomes` = tax_05$SharedGenomes,
  `Mantis annotation` = tax_05$MantisName,
  `Microbe annotation` = tax_05$MicrobeName,
  PlasticDB = tax_05$PlasticDB,
  `Possible depolymerase` = tax_05$Depolymerase,
  `Length (aa)` = nchar(full_seq[members_05]),
  `Mean count, Sediment` = round(colMeans(cnt_da[md$substrate == "Sediment", members_05, drop = FALSE]), 2),
  check.names = FALSE
)
info_05[[paste0("Mean count, ", top_05$substrate)]] <-
  round(colMeans(cnt_da[md$substrate == top_05$substrate, members_05, drop = FALSE]), 2)
peak_05 <- pmax(info_05[["Mean count, Sediment"]], info_05[[paste0("Mean count, ", top_05$substrate)]])
info_05 <- info_05[order(-peak_05), ]   # re-used below, so the sort has to stick
peak_05 <- sort(peak_05, decreasing = TRUE)
gt(info_05)
```

```{r top-bin-driver-05}
driver_05 <- info_05$Protein[1]
driver_annotated_05 <- is_annotated(tax_05[driver_05, ])
driver_desc_05 <- if (driver_annotated_05) {
  "does carry a curated call in this workbook -- see the table above for which one"
} else {
  "carries no `MantisName`, `MicrobeName`, `PlasticDB` or depolymerase call in this workbook"
}

top_share_05     <- peak_05[1] / sum(peak_05)
n_substantial_05 <- sum(peak_05 >= 0.1 * peak_05[1])
concentrated_05  <- top_share_05 >= 0.5

spread_desc_05 <- if (concentrated_05) {
  sprintf("It sits almost entirely in %s, detected in %d of 3 %s replicates and %d of 3 sediment replicates; the other %d members are near-zero traces in every sample, %s included, and contribute negligibly to the bin's summed abundance -- that protein %s%s.",
          driver_05,
          sum(cnt_da[md$substrate == top_05$substrate, driver_05] > 0), top_05$substrate,
          sum(cnt_da[md$substrate == "Sediment", driver_05] > 0),
          length(members_05) - 1, top_05$substrate, driver_desc_05,
          if (driver_annotated_05) "" else " -- the cleanest substrate-specific signal this differential-abundance pass found is also a member this workbook's own annotation pipeline said nothing about")
} else {
  sprintf("It is spread across the bin rather than concentrated in one protein: %s is the single largest contributor (%.0f%% of the bin's peak abundance) but %d of the %d members each reach at least a tenth of that -- a coherent multi-member signal, not a single-protein artifact. Its largest contributor %s%s.",
          driver_05, 100 * top_share_05, n_substantial_05, length(members_05), driver_desc_05,
          if (driver_annotated_05) "" else " -- unusual for a bin this well-supported; most of this workbook's dark proteins turn up alone, not as part of a mostly-annotated family")
}
```

Sorted by each member's peak abundance, on whichever side of the contrast it is
larger, the table shows the bin's fold change: `r spread_desc_05`

`repdist_bin()`'s complete-linkage guarantee is that every pair inside
a bin has predicted TM-score >= 0.5 to every *other* member, not just a nearest
neighbor -- true here by construction, but a heatmap shows what it looks like: the
full pairwise predicted TM-score matrix for bin `r top_05$bin`, same protein order as
the table above.

```{r tm-heatmap-05, fig.height = 5, fig.width = 6}
Zb_05 <- Z[info_05$Protein, ]
S_05  <- 1 - as.matrix(proxy::dist(Zb_05, method = "cosine"))   # cosine == predicted TM-score
dimnames(S_05) <- list(info_05$Protein, info_05$Protein)

df_05 <- as.data.frame(as.table(S_05))
names(df_05) <- c("i", "j", "TM_hat")
df_05$i <- factor(df_05$i, levels = info_05$Protein, labels = seq_along(info_05$Protein))
df_05$j <- factor(df_05$j, levels = info_05$Protein, labels = seq_along(info_05$Protein))

ggplot(df_05, aes(i, j, fill = TM_hat)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0.5, 1), name = "predicted\nTM-score") +
  coord_fixed() +
  labs(x = "protein, table row above", y = "protein, table row above") +
  theme_minimal()
```

The floor is `r round(min(S_05[upper.tri(S_05)]), 3)`, comfortably above the 0.5
cutoff, and the diagonal is exactly 1 by construction (`cos(z, z) = 1`).

```{r depolymerase-scope-05}
sig_members_05 <- names(roll_05$bin)[roll_05$bin %in% rownames(wald_05)[wald_05$reject]]

gt(data.frame(
  scope = c("whole protein universe", "substrate-significant bins only"),
  proteins = c(ncol(cnt_da), length(sig_members_05)),
  plasticdb_or_depolymerase_flag = c(sum(flagged[colnames(cnt_da)]), sum(flagged[sig_members_05])),
  pct = round(100 * c(mean(flagged[colnames(cnt_da)]), mean(flagged[sig_members_05])), 2)))
```

### TM-score >= 0.7

```{r rollup-07}
roll_07 <- roll_both[["0.7"]]
c(proteins = ncol(cnt_da), bins = ncol(roll_07$counts))
```

Same model as the 0.5 pass above, on a different (stricter, more numerous) set of
bins.

`r sum(colSums(roll_07$counts > 0) < 3)` of `r ncol(roll_07$counts)` bins here are
nonzero in fewer than 3 of `r nrow(roll_07$counts)` samples, LinDA's own stated bar for
"virtually no statistical power." Testing them anyway is not just wasted power: fitting
every bin as one joint model (`model = "LM"`, no random effect to loop over) makes its
covariance matrix scale with bin count squared, and at the full
`r ncol(roll_07$counts)` -- unlike the looser threshold above -- it exceeds this
machine's memory outright. `prev.filter` drops them before fitting rather than after --
the same features LinDA's own warning already flags as underpowered, just excluded up
front instead of tested and ignored.

```{r linda-07}
fit_07 <- linda(t(roll_07$counts), md, formula = "~ substrate",
               feature.dat.type = "count", prev.filter = 3 / nrow(roll_07$counts), verbose = FALSE)
```

```{r linda-wald-07}
k <- nlevels(md$substrate) - 1
L <- matrix(0, nrow = k, ncol = length(fit_07$variables) + 1)
for (i in 1:k) L[i, i + 1] <- 1
wald_07 <- linda.wald.test(fit_07, L, model = "LM", alpha = 0.05)
rownames(wald_07) <- rownames(fit_07$output[[1]])

gt(data.frame(bins_tested = nrow(wald_07), bins_substrate_sig = sum(wald_07$reject),
              fraction_sig = round(mean(wald_07$reject), 3)))
```

`r round(100 * mean(wald_07$reject))`% of bins reject at `padj <= 0.05`.

```{r volcano-07, fig.height = 3.2, fig.width = 10}
substrate_vars_07 <- grep("^substrate", fit_07$variables, value = TRUE)
volc_07 <- do.call(rbind, lapply(substrate_vars_07, function(v) {
  o <- fit_07$output[[v]]
  data.frame(bin = rownames(o), substrate = sub("^substrate", "", v),
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue, reject = o$reject)
}))

ggplot(volc_07, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ substrate, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"), guide = "none") +
  labs(x = "log2 fold change vs Sediment", y = expression(-log[10](p))) + theme_bw()
```

```{r top-bin-07}
top_07 <- volc_07[volc_07$reject, ]
top_07 <- top_07[which.max(abs(top_07$log2FoldChange)), ]

members_07 <- names(roll_07$bin)[roll_07$bin == top_07$bin]
tax_07 <- as.data.frame(as(tax_table(ps), "matrix"))[members_07, ]
n_annotated_07    <- sum(is_annotated(tax_07))
n_multi_genome_07 <- sum(!is.na(tax_07$SharedGenomes))
```

The single largest shift, in either direction, is bin `r top_07$bin` on
`r top_07$substrate` vs. sediment: `r sprintf("%+.1f", top_07$log2FoldChange)` log2
fold change at $p =$ `r formatC(top_07$pvalue, digits = 1, format = "e")`. Its
`r length(members_07)` members carry no shared lineage (this workbook has no GTDB
call, `data_curation_case_study2.Rmd`), and `r n_annotated_07` of `r length(members_07)`
carries any curated call at all (`MantisName`, `MicrobeName`, a `PlasticDB` hit, or a
depolymerase flag) -- the table below has the specifics. `r n_multi_genome_07` of the
`r length(members_07)` members were independently assembled and binned in more than
one sample's `GenomeID` (the `SharedGenomes` column below).

```{r top-bin-table-07}
info_07 <- data.frame(
  Protein = members_07,
  `Genome ID` = tax_07$GenomeID,
  `Shared genomes` = tax_07$SharedGenomes,
  `Mantis annotation` = tax_07$MantisName,
  `Microbe annotation` = tax_07$MicrobeName,
  PlasticDB = tax_07$PlasticDB,
  `Possible depolymerase` = tax_07$Depolymerase,
  `Length (aa)` = nchar(full_seq[members_07]),
  `Mean count, Sediment` = round(colMeans(cnt_da[md$substrate == "Sediment", members_07, drop = FALSE]), 2),
  check.names = FALSE
)
info_07[[paste0("Mean count, ", top_07$substrate)]] <-
  round(colMeans(cnt_da[md$substrate == top_07$substrate, members_07, drop = FALSE]), 2)
peak_07 <- pmax(info_07[["Mean count, Sediment"]], info_07[[paste0("Mean count, ", top_07$substrate)]])
info_07 <- info_07[order(-peak_07), ]
peak_07 <- sort(peak_07, decreasing = TRUE)
gt(info_07)
```

```{r top-bin-driver-07}
driver_07 <- info_07$Protein[1]
driver_annotated_07 <- is_annotated(tax_07[driver_07, ])
driver_desc_07 <- if (driver_annotated_07) {
  "does carry a curated call in this workbook -- see the table above for which one"
} else {
  "carries no `MantisName`, `MicrobeName`, `PlasticDB` or depolymerase call in this workbook"
}

top_share_07     <- peak_07[1] / sum(peak_07)
n_substantial_07 <- sum(peak_07 >= 0.1 * peak_07[1])
concentrated_07  <- top_share_07 >= 0.5

spread_desc_07 <- if (concentrated_07) {
  sprintf("It sits almost entirely in %s, detected in %d of 3 %s replicates and %d of 3 sediment replicates; the other %d members are near-zero traces in every sample, %s included, and contribute negligibly to the bin's summed abundance -- that protein %s%s.",
          driver_07,
          sum(cnt_da[md$substrate == top_07$substrate, driver_07] > 0), top_07$substrate,
          sum(cnt_da[md$substrate == "Sediment", driver_07] > 0),
          length(members_07) - 1, top_07$substrate, driver_desc_07,
          if (driver_annotated_07) "" else " -- the cleanest substrate-specific signal this differential-abundance pass found is also a member this workbook's own annotation pipeline said nothing about")
} else {
  sprintf("It is spread across the bin rather than concentrated in one protein: %s is the single largest contributor (%.0f%% of the bin's peak abundance) but %d of the %d members each reach at least a tenth of that -- a coherent multi-member signal, not a single-protein artifact. Its largest contributor %s%s.",
          driver_07, 100 * top_share_07, n_substantial_07, length(members_07), driver_desc_07,
          if (driver_annotated_07) "" else " -- unusual for a bin this well-supported; most of this workbook's dark proteins turn up alone, not as part of a mostly-annotated family")
}
```

Sorted by each member's peak abundance, on whichever side of the contrast it is
larger, the table shows the bin's fold change: `r spread_desc_07`

```{r tm-heatmap-07, fig.height = 5, fig.width = 6}
Zb_07 <- Z[info_07$Protein, ]
S_07  <- 1 - as.matrix(proxy::dist(Zb_07, method = "cosine"))
dimnames(S_07) <- list(info_07$Protein, info_07$Protein)

df_07 <- as.data.frame(as.table(S_07))
names(df_07) <- c("i", "j", "TM_hat")
df_07$i <- factor(df_07$i, levels = info_07$Protein, labels = seq_along(info_07$Protein))
df_07$j <- factor(df_07$j, levels = info_07$Protein, labels = seq_along(info_07$Protein))

ggplot(df_07, aes(i, j, fill = TM_hat)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0.7, 1), name = "predicted\nTM-score") +
  coord_fixed() +
  labs(x = "protein, table row above", y = "protein, table row above") +
  theme_minimal()
```

The floor is `r round(min(S_07[upper.tri(S_07)]), 3)`, comfortably above the 0.7
cutoff.

```{r depolymerase-scope-07}
sig_members_07 <- names(roll_07$bin)[roll_07$bin %in% rownames(wald_07)[wald_07$reject]]

gt(data.frame(
  scope = c("whole protein universe", "substrate-significant bins only"),
  proteins = c(ncol(cnt_da), length(sig_members_07)),
  plasticdb_or_depolymerase_flag = c(sum(flagged[colnames(cnt_da)]), sum(flagged[sig_members_07])),
  pct = round(100 * c(mean(flagged[colnames(cnt_da)]), mean(flagged[sig_members_07])), 2)))
```

## Conclusion

All three distances find substrate highly significant ($p = 10^{-4}$, the ceiling of
9999 permutations), with diet -- here, substrate -- $R^2$ of
`r sprintf("%.3f", r2["Bray-Curtis"])` for Bray-Curtis,
`r sprintf("%.3f", r2["Jaccard"])` for Jaccard, and
`r sprintf("%.3f", r2["w-UniFrac (structure)"])` for the representation. w-UniFrac (structure) leads both,
by a larger margin than in the diet study (`r sprintf("%+.3f", r2["w-UniFrac (structure)"] - r2["Bray-Curtis"])`
over Bray-Curtis here against `case_study.Rmd`'s `r sprintf("%+.3f", 0.325 - 0.253)`).
All three pass the dispersion
diagnostic (betadisper $p > 0.6$ for every distance), so none of this is groups
differing in spread rather than location.

The accession-sharing numbers (above) explain part of why protein identity does as
well as it does here: this dataset sits between the diet study's near-universal,
single-database accession sharing and a fully private-accession regime -- a real but
partial minority of the catalog is shared across substrates, which is exactly the
scenario where a protein-identity distance still has genuine signal to read, just less
of it.

`repdist_bin()` collapsed the `r ncol(cnt_da)` proteins into
`r ncol(roll_05$counts)` bins at TM-score >= 0.5 and `r ncol(roll_07$counts)` at
>= 0.7; LinDA's joint test rejected `r round(100 * mean(wald_05$reject), 1)`% and
`r round(100 * mean(wald_07$reject), 1)`% of them respectively at `padj <= 0.05` --
substrate reorganizes abundance broadly across the structural repertoire under either
clustering, the same qualitative picture the sample-level PERMANOVA gives, and neither
threshold's top hit is a low-power artifact of the other's bin sizes: the largest
shift at 0.5 is bin `r top_05$bin` on `r top_05$substrate`
(`r sprintf("%+.1f", top_05$log2FoldChange)` log2FC), and at 0.7 it is bin
`r top_07$bin` on `r top_07$substrate` (`r sprintf("%+.1f", top_07$log2FoldChange)`
log2FC) -- `r if (identical(top_05$substrate, top_07$substrate)) "the same substrate at both thresholds" else "a different substrate at each threshold, so this particular top hit is sensitive to where the clustering line is drawn"`.
Curated `PlasticDB`/depolymerase flags are rare workbook-wide either way
(`r sum(flagged[colnames(cnt_da)])` of `r ncol(cnt_da)` proteins,
`r round(100 * mean(flagged[colnames(cnt_da)]), 2)`%); among substrate-significant
bins they are carried by `r round(100 * mean(flagged[sig_members_05]), 2)`% at 0.5 and
`r round(100 * mean(flagged[sig_members_07]), 2)`% at 0.7 -- on the annotation this
workbook happened to curate rather than a formal enrichment test either way.

This report does not repeat `case_study.Rmd`'s null-control battery (random-embedding,
amino-acid-composition and length baselines) or its accession-masking shuffle test, so
unlike that report it does not directly address whether the sample-level `sample_repdist`
signal, or the structural bins built on it, are separable from amino-acid composition
or from how much samples happen to share literal protein calls. Both are a
straightforward rerun of `case_study.Rmd`'s corresponding chunks against this
dataset's `cnt`/`Z` if that comparison is wanted.
