---
title: "Re analysis of diet protein study"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE, autodep = TRUE,
                      fig.width = 8, fig.height = 4, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

The simulation study showed that a protein-identity distance goes blind when
samples share no accessions, and that a TM-Vec ground metric does not. This report
applies the same comparison to real data: 111 mouse faecal metaproteomes across
eight dietary protein sources (Blakeley-Ruiz *et al.*, *ISME J* 19(1) wraf048),
quantified by spectral counting.

There are two weighted-UniFrac arms, and neither comes from an alignment. The
simulation could align 480 proteins from four families; here the universe is
~19,000 proteins from a whole gut community, mostly non-homologous, and a global
MSA of them is ~99% gaps -- the longest sequence is 32,877aa against a median of
365, so any tree inferred from that alignment is noise, not phylogeny.

Both trees are therefore fitted from a distance matrix instead, by the same
complete-linkage algorithm: one from 3-mer sequence composition, one
from the TM-Vec structure metric. That keeps the contrast on the *representation*
-- sequence against structure -- rather than confounding it with the tree-fitting
algorithm. `analysis/build_trees.R test_study` builds both and caches them.

```{r libs, cache = FALSE}
library(phyloseq)
library(vegan)
library(ggplot2)
library(gt)
library(here)
library(MicrobiomeStat)
devtools::load_all(".", quiet = TRUE)

path  <- here("data", "test_study")
NPERM <- 9999
```

## Data

```{r data}
ps <- readRDS(file.path(path, "ps.rds"))   # built and validated in data_curation.Rmd
ps
```

## Rarefaction

```{r depth-check, fig.height = 3}
floor_depth <- min(sample_sums(ps))    # rarefy to the shallowest sample, dropping none

ggplot(data.frame(depth = sort(sample_sums(ps))), aes(seq_along(depth), depth)) +
  geom_col(width = 1) +
  geom_hline(yintercept = floor_depth, linetype = "dashed", color = "firebrick") +
  labs(x = "sample, sorted by depth", y = "PSMs") + theme_bw()
```

The shallowest samples sit within a few percent of each other, so there is no isolated
outlier to drop and the cutoff (dashed line) is simply the minimum depth,
`r floor_depth` PSMs. Taking it from `min()` rather than hardcoding it is what keeps
that sentence true: any fixed number above the minimum silently discards the samples
below it.

```{r rarefy}
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)

# T0 has no protein source, so the diet term is undefined for it
psd <- prune_samples(sample_data(psd)$phase != "baseline", psd)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)
sample_data(psd)$diet <- factor(sample_data(psd)$diet)
sample_data(psd)$dose <- factor(sample_data(psd)$dose)
psd
```

Rarefying before pruning matters: rarefaction subsamples the raw counts, so it has
to see the raw counts.

```{r unrarefied}
# same baseline exclusion as psd, built straight off ps instead -- this is the
# unrarefied counterpart the Differential abundance section uses; see there for why
psu <- prune_samples(sample_data(ps)$phase != "baseline", ps)
psu <- prune_taxa(taxa_sums(psu) > 0, psu)
sample_data(psu)$diet <- factor(sample_data(psu)$diet)
sample_data(psu)$dose <- factor(sample_data(psu)$dose)
psu
```

## Embedding

TM-Vec1, the ground metric chosen in `docs/ground_metric_models.md`: its head was
trained to regress the TM-score of protein pairs, so cosine similarity between two
vectors is a predicted TM-score. The large SWISS-MODEL head (default as of
`embed_proteins()`, trained through 1000 residues) is used here rather than the
300-residue head, since most of this proteome's abundance falls between the two.
Full sequences are passed to ProtT5; `embed_proteins()` forms batches from the
selected device's available memory rather than truncating at a fixed length. The
TM-Vec training range is a QC boundary, not an encoder limit. The result is cached, keyed by
exact sequence and embedding-policy version, not just protein id -- neither an old
truncated vector nor a vector from a different model can be silently reused.

```{r embed, cache = FALSE}
# cache = FALSE on purpose: this chunk's own code never changes, so knitr's
# chunk cache would not notice that `psd` changed upstream (e.g. a different
# rarefaction depth) and would silently replay a stale Z. The rds file below is
# the only cache, and it lives in data/ rather than in the knitr cache directory.
model_tag <- paste0(eval(formals(embed_proteins)$model), "-full-length-v1")
cache <- file.path(path, "embeddings_tmvec1_large.rds")
cached <- if (file.exists(cache)) readRDS(cache) else list(Z = NULL, seq = character(0), model = character(0))

full_seq <- stats::setNames(as.character(refseq(psu)[taxa_names(psu)]), taxa_names(psu))

# keyed off psu (unrarefied, baseline already dropped), not psd -- covers every
# protein either count matrix below could need, rarefied or not, so a later change
# to rarefaction depth or the DA section never has to come back through this chunk
reusable <- intersect(taxa_names(psu), rownames(cached$Z))
reusable <- reusable[cached$seq[reusable] == full_seq[reusable] &
                      cached$model[reusable] == model_tag]
need  <- setdiff(taxa_names(psu), reusable)
Zall  <- cached$Z

if (length(need)) {
  s <- full_seq[need]
  s <- s[order(nchar(s))]                       # length-sorted: keeps batches tight
  # Save every 2000 proteins rather than once at the end. The full pass is about
  # an hour, so an interrupted render resumes from the last block instead of
  # starting over.
  for (i in split(seq_along(s), ceiling(seq_along(s) / 2000))) {
    new_Z <- embed_proteins(s[i])
    Zall  <- rbind(Zall[setdiff(rownames(Zall), rownames(new_Z)), , drop = FALSE], new_Z)
    cached$Z <- Zall
    cached$seq[rownames(new_Z)]   <- full_seq[rownames(new_Z)]
    cached$model[rownames(new_Z)] <- model_tag
    saveRDS(cached, cache)
  }
}
Z      <- Zall[taxa_names(psu), ]
cnt    <- t(as(otu_table(psd), "matrix"))   # samples in rows, RAREFIED -- distances/PCoA only
cnt_da <- t(as(otu_table(psu), "matrix"))   # samples in rows, UNRAREFIED -- differential abundance only
```

Full-length embedding removes a technical truncation, but proteins beyond the
head's 1000-residue training range remain extrapolations.
`length_abundance_profile()` shows how much of the catalog and total abundance
cross that scientific boundary.

```{r length-abundance, fig.height = 4}
length_abundance_profile(cnt, full_seq, max_length = 1000) +
  ggplot2::geom_vline(xintercept = 1300, linetype = "dotted", color = "grey30")
```

At the 1000aa line (dashed), `r round(100*mean(nchar(full_seq[colnames(cnt)]) > 1000), 1)`%
of the catalog by count sits past the training range, but the abundance curve --
consistently above the protein-count curve throughout -- puts
`r round(100*sum((cnt/rowSums(cnt))[, nchar(full_seq[colnames(cnt)]) > 1000])/nrow(cnt), 1)`%
of total abundance there instead, roughly
`r round(sum((cnt/rowSums(cnt))[, nchar(full_seq[colnames(cnt)]) > 1000])/nrow(cnt) / mean(nchar(full_seq[colnames(cnt)]) > 1000), 1)`x
its share by count. Part of that is specific, identifiable large protein families in this
proteome that are individually abundant regardless of length -- S-layer proteins, a
TonB-dependent transporter (SusC), multi-domain metabolic enzymes such as
pyruvate:ferredoxin oxidoreductase -- riding on top of what the protein-count curve alone
would predict.

The dashed line is not an encoder stop: every residue is embedded when the sequence
fits the device's safe memory budget. It marks where the TM-Vec head leaves its
training distribution. Two aggregation properties limit the influence of those
extrapolations without making them validated predictions:

- **The sample-level distance (`sample_repdist()`, below) is abundance-weighted over
  thousands of proteins at once.** A single sample's diet signal does not hinge on any one
  protein's embedding quality; the aggregate is robust to a small fraction of the catalog
  being a rougher approximation, which is exactly the shape both curves above show --
  concentrated well under 1300aa.
- **The bin-level analysis (`repdist_bin()`, below) is a per-bin risk, not a
  proteome-wide one.** A long protein grouped with the wrong bin is an ordinary
  clustering error. It matters most if that protein dominates a small,
  otherwise-uninformative bin's annotation.

`1300` (dotted line) is a reasonable place to draw that trust boundary on this dataset:
past it, only `r round(100*mean(nchar(full_seq[colnames(cnt)]) > 1300), 1)`% of the
catalog and `r round(100*sum((cnt/rowSums(cnt))[, nchar(full_seq[colnames(cnt)]) > 1300])/nrow(cnt), 1)`%
of total abundance remain. Predictions this far beyond the training range deserve
extra scrutiny when they drive a bin's interpretation.

## Results

Bray-Curtis and Jaccard are both protein-identity distances -- two proteins are
either the same accession or maximally different -- and differ only in whether
abundance enters: Bray-Curtis weights by how much of each protein was detected,
Jaccard only by whether it was detected at all.

```{r distances}
# cnt built in the Embedding section above, reused here
md <- data.frame(sample_data(psd))

# repdist does not implement UniFrac. It fits the tree; phyloseq computes the
# distance. Raw (normalized = FALSE) is the form that equals W1 under the tree's
# path metric -- the normalized form divides by a scale factor and is no longer a
# transport cost.
#
# Complete linkage uses the same hierarchy as repdist_bin(), so a tree cut at a
# TM-score threshold preserves the package's all-pairs bin guarantee.
wunifrac <- function(counts, tree) {
  ps_t <- phyloseq(otu_table(counts, taxa_are_rows = FALSE), phy_tree(tree))
  UniFrac(ps_t, weighted = TRUE, normalized = FALSE)
}

# same helper the covariate baselines below use: ground metric -> tree -> UniFrac
rep_unifrac <- function(counts, Z, distance = "cosine") {
  G <- seq_repdist(Z[colnames(counts), , drop = FALSE], distance = distance)
  on.exit(gc(verbose = FALSE), add = TRUE)   # the ground matrix is gigabytes here
  tree <- ape::as.phylo(fastcluster::hclust(G, method = "complete"))
  wunifrac(counts, tree)
}

# both trees precomputed by analysis/build_trees.R test_study
tr <- readRDS(file.path(path, "trees_kmer_struct.rds"))
stopifnot(setequal(tr$proteins, colnames(cnt)))
tree_fidelity <- tr$fidelity   # c(sequence = ..., structure = ...)

# "store the data as another phyloseq object" -- same counts, one tree each
psd_seq <- phyloseq(otu_table(psd), sample_data(psd), phy_tree(tr$tree_seq))
psd_str <- phyloseq(otu_table(psd), sample_data(psd), phy_tree(tr$tree_str))

G_tm <- seq_repdist(Z[colnames(cnt), , drop = FALSE])

D <- list(`Bray-Curtis`            = phyloseq::distance(psd, method = "bray"),
          `Jaccard`                = vegdist(cnt, method = "jaccard", binary = TRUE),
          `w-UniFrac (sequence)`   = UniFrac(psd_seq, weighted = TRUE, normalized = FALSE),
          `w-UniFrac (structure)`  = UniFrac(psd_str, weighted = TRUE, normalized = FALSE))
D <- lapply(D, function(d) as.dist(as.matrix(d)[rownames(md), rownames(md)]))
```

Amount of protein, sex and age stay in the model as covariates; only the diet term
is reported. `by = "terms"` is sequential, so term order matters -- the nuisance
covariates go first and diet last, so diet's reported $R^2$ is the share left over
*after* dose, sex and age, the more conservative reading rather than the one that
hands diet whatever it and a covariate share. Permutations are blocked within mouse,
since each mouse contributes several samples. `sex` is constant within a mouse, so it
has no valid permutation distribution under that blocking — it absorbs variance rather
than being tested.

```{r permanova}
diet_test <- function(dl, nperm = NPERM) do.call(rbind, lapply(names(dl), function(n) {
  set.seed(1)
  a <- adonis2(dl[[n]] ~ dose + sex + age_week + diet, data = md, by = "terms",
               permutations = permute::how(blocks = factor(md$subject), nperm = nperm))
  data.frame(distance = n, R2 = round(a["diet", "R2"], 3),
             F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"])
}))
res <- diet_test(D)
r2  <- setNames(res$R2, res$distance)   # name-based lookup for the prose below
gt(res)
```

```{r pcoa, cache = FALSE, fig.height = 4.5}
for (n in names(D)) {
  print(plot_ordination(psd, ordinate(psd, "PCoA", distance = D[[n]]), color = "diet") +
          stat_ellipse(type = "t", geom = "polygon", alpha = 0) +
          ggtitle(n) + theme_bw())
}
```

## Diagnostics

A PERMANOVA $p$ can come from groups differing in *spread* rather than in
*location*, and the test cannot separate the two on its own. `betadisper` can --
`bias.adjust = TRUE` corrects the known bias from diet groups having different
sample sizes, and significance comes from a permutation test blocked within mouse,
the same restriction the PERMANOVA above uses, rather than `anova()`'s assumption
of independent observations.

```{r diagnostics}
gt(data.frame(
  distance     = names(D),
  betadisper_p = sapply(D, function(d) {
    bd <- betadisper(d, md$diet, bias.adjust = TRUE)
    pt <- permutest(bd, permutations = permute::how(blocks = factor(md$subject), nperm = NPERM))
    round(pt$tab$`Pr(>F)`[1], 4)
  }),
  row.names = NULL))
```

## Null controls

A significant `sample_repdist` PERMANOVA is not by itself evidence that TM-Vec's learned
geometry contributed anything -- the ground metric might only be re-expressing something
much simpler about the sequences. Two baselines probe that, neither requiring a model:
amino-acid composition (20-d, does this reduce to what proteins are made of) and log
length (1-d, does this reduce to how long they are). Both are run through exactly the same
distance as the TM-Vec arm, so only the ground metric differs.

```{r null-controls}
aa <- strsplit("ACDEFGHIKLMNPQRSTVWY", "")[[1]]
Z_comp <- t(vapply(strsplit(full_seq, ""), function(ch) {
  tabulate(match(ch, aa), length(aa)) / length(ch)
}, numeric(length(aa))))
dimnames(Z_comp) <- list(names(full_seq), aa)

Z_len <- matrix(log(nchar(full_seq)), dimnames = list(names(full_seq), "log_length"))

r2_comp <- diet_test(list(c = rep_unifrac(cnt, Z_comp)))$R2
# log length is one column: a magnitude with no direction, so cosine is undefined
r2_len  <- diet_test(list(l = rep_unifrac(cnt, Z_len, distance = "euclidean")))$R2

null_res <- rbind(
  data.frame(baseline = "amino-acid composition", R2 = r2_comp),
  data.frame(baseline = "log length", R2 = r2_len),
  data.frame(baseline = "TM-Vec1 (w-UniFrac structure, from the table above)", R2 = unname(r2["w-UniFrac (structure)"])))
rownames(null_res) <- NULL
gt(null_res)
```

TM-Vec's diet $R^2$ (`r sprintf("%.3f", r2["w-UniFrac (structure)"])`) clears the length-only
baseline (`r sprintf("%.3f", r2_len)`), so this is not simply a restatement of how long
the detected proteins are.

It does **not** clear the amino-acid composition baseline: composition alone reaches
`r sprintf("%.3f", r2_comp)` against TM-Vec's `r sprintf("%.3f", r2["w-UniFrac (structure)"])` --
`r sprintf("%+.3f", r2["w-UniFrac (structure)"] - r2_comp)`, composition ahead. TM-Vec is not reading
noise, and it is not just reading protein length, but on this test it does not demonstrably
read *more* than what a 20-number amino-acid composition vector already reads. That matches
the same pattern the fold-discrimination check in `docs/ground_metric_models.md` found at
the protein-pair level (ESM-2 8M AUROC 0.766 against a composition baseline of 0.585, a
comparable margin) -- composition confound is not a hypothetical here, it is measured twice
now, at two different levels of the pipeline. The honest reading is that this report's
significant `sample_repdist` result and the structural bins built on it are consistent with
structure carrying the diet signal, but do not on their own rule out amino-acid composition
carrying it instead. Separating the two needs a test where structure and composition
disagree, not one where both happen to move together, which is what this dataset offers.

## What Bray-Curtis and Jaccard are actually using

All three distances find a highly significant diet effect and land in the same
broad range, which the simulation would not have predicted. The reason is that
these are cage-housed lab mice searched against one protein database: samples
really do share *the same accessions*, and exact accession identity is something
Bray-Curtis and Jaccard can read directly. Jaccard's lower $R^2$ against
Bray-Curtis's is the presence/absence penalty: it throws away how much of each
shared protein was detected, keeping only whether it was detected at all.

The test is to take that away. Give every sample its own private accession for each
protein it detected. Counts, sequences and abundances are all unchanged — only the
labels differ — so no two samples share an accession any more.

```{r shuffle}
nz  <- which(cnt > 0, arr.ind = TRUE)
lab <- paste0("protein", seq_len(nrow(nz)))   # masked: one private label per (sample, protein) detection

cnt_s <- matrix(0L, nrow(cnt), length(lab), dimnames = list(rownames(cnt), lab))
cnt_s[cbind(nz[, "row"], seq_len(nrow(nz)))] <- cnt[nz]
Z_s <- Z[nz[, "col"], ]
rownames(Z_s) <- lab

c(accessions_before = ncol(cnt), accessions_after = ncol(cnt_s),
  max_samples_sharing_one = max(colSums(cnt_s > 0)))
```

```{r shuffle-test}
# This is the one arm that cannot use a tree or `sample_repdist()` at all.
# Masking expands the table to one private accession per detection -- 208,177
# columns -- and every ground-metric method (UniFrac, ot, rbf_mmd) would need
# that protein x protein matrix explicitly, which is ~347 TB.
#
# Linear-kernel MMD is the one member of the family that never forms it: with a
# linear kernel the Gram matrix reduces exactly to Z %*% t(Z), so the distance
# between weighted mean embeddings is just dist(P %*% Z) -- one matmul at
# 208,177 x 512 instead of a square matrix. It is one line, so it is written
# here rather than carried as a third `method` in the package.
#
# The invariance being demonstrated does not depend on which distance is used:
# no embedding-based distance can read an accession label.
lin_mmd <- function(counts, Z) stats::dist((counts / rowSums(counts)) %*% Z[colnames(counts), ])

res_s <- diet_test(list(`Bray-Curtis` = vegdist(cnt_s / rowSums(cnt_s), "bray"),
                        `Jaccard`     = vegdist(cnt_s, method = "jaccard", binary = TRUE),
                        `linear MMD`  = lin_mmd(cnt_s, Z_s)))
gt(res_s)
```

Bray-Curtis and Jaccard both collapse to the exact null. Every pair is now
maximally dissimilar, so each distance matrix is constant and cannot express any
grouping: $F = 1$, $p = 1$, and $R^2 = 5/98 = 0.051$, which is simply the diet
term's degrees of freedom over $n-1$ — the share any 5-df term receives when there
is nothing to explain. The embedding-based arm returns exactly what it returned
above, because a protein's embedding does not depend on what its accession is
called and the abundance-weighted mean is therefore unchanged. This arm uses
linear-kernel MMD rather than the structure tree used everywhere else, for the
reason given in the chunk above -- masking makes the ground matrix too large to
form. The invariance is a property of the embedding, not of the method: a
structure tree built on masked labels would be the same tree with the tips
renamed, since the embeddings feeding it are identical.

So Bray-Curtis's and Jaccard's whole result here rests on samples sharing literal
accessions. That is a fair assumption within one colony, one database and one
assembly. It stops being fair across studies, colonies or assemblies, where the
same function arrives under a different accession — which is the case the
representation distance is built for.

## Differential abundance

Unrarefied counts (`cnt_da`, from `psu`), not the rarefied `cnt` the sections above
use. Rarefaction earns its keep for a sample-level distance -- Bray-Curtis and
`sample_repdist` need every sample on the same footing, and subsampling is the standard
way to buy that. A per-bin model doesn't have the same need: LinDA fits its own
bias-corrected centered-log-ratio transform per sample, which already absorbs each
sample's own depth, so throwing away real counts on top of that would trade
statistical power for nothing (McMurdie & Holmes, *PLoS Comp Biol* 2014, arguing
rarefaction is broadly the wrong tool for differential-abundance testing
specifically, as opposed to beta-diversity). Unrarefied keeps `r ncol(cnt_da)`
proteins against `r ncol(cnt)` rarefied.

PERMANOVA tests whether diet structures the community as a whole; it says
nothing about which proteins carry that signal. Testing all `r ncol(cnt_da)`
proteins directly would mean that many near-independent tests on units that
mostly aren't independent -- protein groups are frequently homologous
(`docs/method_notes.md` §1), so a chunk of those "tests" are the same
signal counted several times. `repdist_bin()` collapses proteins into
bins where every member has predicted TM-score above a threshold to every other
member; both the conventional TM-score same-fold boundary (0.5, Xu & Zhang 2010,
used to pick the embedding model above) and a stricter cutoff (0.7, tighter and
more homogeneous bins at the cost of more of them) are run below, as two
independent passes over the same counts and embeddings -- multiplicity is paid
once per structural unit within each pass, not once per accession, but the two
passes are not compared to each other statistically, only side by side.

```{r rollup-both}
roll_both <- repdist_bin(cnt_da, Z, cluster_threshold = c(0.5, 0.7))$bins
```

Passing both thresholds at once builds the pairwise TM-Vec distance matrix and its
linkage tree once and cuts it at both heights, rather than paying that cost (the
slow part; the cut itself is nearly free) twice over for two separate
`repdist_bin()` calls.

```{r shared-setup}
t6 <- read.delim(file.path(path, "annotated", "ExtendedDataTable6AnnotatedMicrobialMetaproteomes.txt"),
                 skip = 2, quote = "\"", check.names = FALSE, colClasses = "character",
                 comment.char = "", fileEncoding = "latin1")
is_dark <- function(ids) {
  a <- setNames(t6[["Consensus annotation"]], t6[["Protein Identifier"]])[ids]
  g <- as.data.frame(as(tax_table(psu), "matrix"))[ids, "Genus"]
  is.na(a) & is.na(g)
}
```

### TM-score >= 0.5

```{r rollup-05}
roll_05 <- roll_both[["0.5"]]
c(proteins = ncol(cnt_da), bins = ncol(roll_05$counts))
```

LinDA (Zhou *et al.*, *Genome Biology* 2022) fits a linear model per bin on
centered-log-ratio-transformed abundance and corrects the compositional bias
a plain `lm()` on relative abundance would carry -- the same reason
`sample_repdist()` works on abundance weights rather than raw proportions
directly. Its formula interface takes a random effect for repeated sampling
of the same mouse, the model-based counterpart to the permutation blocking
`adonis2` used above.

`r sum(colSums(roll_05$counts > 0) < 3)` of `r ncol(roll_05$counts)` bins here are
nonzero in fewer than 3 of `r nrow(roll_05$counts)` samples, LinDA's own stated bar
for "virtually no statistical power" -- `prev.filter` drops them before fitting
rather than testing and ignoring them. The remaining sparse bins can still report a
singular fit (their between-mouse variance estimated at zero), which `lmer` warns
about but does not treat as an error.

```{r linda-05}
fit_05 <- linda(t(roll_05$counts), md,
               formula = "~ diet + dose + sex + age_week + (1 | subject)",
               feature.dat.type = "count", prev.filter = 3 / nrow(roll_05$counts), verbose = FALSE)
```

As with the PERMANOVA table, only the diet term is reported: a joint Wald
test on all five diet coefficients per bin, LinDA's counterpart to
`adonis2`'s `by = "terms"`.

```{r linda-wald-05}
L <- matrix(0, nrow = 5, ncol = length(fit_05$variables) + 1)
for (i in 1:5) L[i, i + 1] <- 1   # intercept + the 5 diet dummy columns
wald_05 <- linda.wald.test(fit_05, L, model = "LMM", alpha = 0.05)
rownames(wald_05) <- rownames(fit_05$output[[1]])   # linda.wald.test drops them; reused below

gt(data.frame(bins_tested = nrow(wald_05), bins_diet_sig = sum(wald_05$reject),
              fraction_sig = round(mean(wald_05$reject), 3)))
```

`r round(100 * mean(wald_05$reject))`% of bins reject at `padj <= 0.05` --
consistent with the PERMANOVA result above: diet is not a handful of
marker proteins, it reorganizes abundance broadly across the structural
repertoire. That joint test has no size to it, though -- diet has five
levels and no single number is "the" effect. A volcano plot needs one
effect size per point, so the five diet-vs-Casein contrasts already inside
`fit_05` are shown separately: one panel per diet, log2 fold change against
Casein on the x-axis, each point one bin, red where that specific contrast
rejects at `padj <= 0.05`.

```{r volcano-05, fig.height = 3.2, fig.width = 10}
diet_vars_05 <- grep("^diet", fit_05$variables, value = TRUE)
volc_05 <- do.call(rbind, lapply(diet_vars_05, function(v) {
  o <- fit_05$output[[v]]
  data.frame(bin = rownames(o), diet = sub("^diet", "", v),
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue, reject = o$reject)
}))

ggplot(volc_05, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"), guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) + theme_bw()
```

```{r top-bin-05}
top_05 <- volc_05[volc_05$reject, ]
top_05 <- top_05[which.max(abs(top_05$log2FoldChange)), ]

members_05 <- names(roll_05$bin)[roll_05$bin == top_05$bin]
ann_05 <- setNames(t6[["Consensus annotation"]], t6[["Protein Identifier"]])[members_05]
tx_05  <- as.data.frame(as(tax_table(psu), "matrix"))[members_05, c("Genus", "Species")]

ann_calls_05 <- unique(na.omit(ann_05))
ann_desc_05  <- if (length(ann_calls_05) == 0) {
  "none of them"
} else if (length(ann_calls_05) == 1) {
  sprintf("%d of them, all the same call: *%s*", sum(!is.na(ann_05)), ann_calls_05)
} else {
  sprintf("%d of them: %s", sum(!is.na(ann_05)), paste0("*", ann_calls_05, "*", collapse = ", "))
}
ann_plausibility_05 <- if (length(ann_calls_05) &&
                           all(grepl("glyco|hydrolase|GH[0-9]", ann_calls_05, ignore.case = TRUE)) &&
                           identical(top_05$diet, "Yeast")) {
  " -- glycoside hydrolases are plausible machinery for a yeast-cell-wall-heavy diet,"
} else ""
```

The single largest shift, in either direction, is bin `r top_05$bin` on
`r top_05$diet`: `r sprintf("%+.1f", top_05$log2FoldChange)` log2 fold change
against Casein at $p=$ `r formatC(top_05$pvalue, digits = 1, format = "e")`.
Its `r length(members_05)` members come from `r length(unique(na.omit(tx_05$Genus)))`
different genera -- proteins TM-Vec1 grouped as one predicted fold despite
sharing no organism, the case an accession-identity distance cannot follow
through a diet shift at all. Extended Data Table 6 annotates `r ann_desc_05`
`r ann_plausibility_05` though this is one bin's annotation coverage, not a
pathway enrichment test.

```{r top-bin-table-05}
info_05 <- data.frame(
  Protein = members_05, Genus = tx_05$Genus, Species = tx_05$Species,
  `Annotation (Table 6)` = unname(ann_05),
  `Length (aa)` = Biostrings::width(refseq(psu)[members_05]),
  `Mean count, Casein` = round(colMeans(cnt_da[md$diet == "Casein", members_05, drop = FALSE]), 1),
  check.names = FALSE
)
info_05[[paste0("Mean count, ", top_05$diet)]] <-
  round(colMeans(cnt_da[md$diet == top_05$diet, members_05, drop = FALSE]), 1)
info_05 <- info_05[order(-info_05[[ncol(info_05)]]), ]   # re-used below, so the sort has to stick
gt(info_05)
```

Every member sits at or near zero on `r if (top_05$log2FoldChange > 0) "Casein" else top_05$diet`
and climbs on `r if (top_05$log2FoldChange > 0) top_05$diet else "Casein"`, which is
what the fold change states in raw counts -- not an artifact of the
log-ratio transform.

`repdist_bin()`'s complete-linkage guarantee is that every pair inside
a bin has predicted TM-score >= 0.5 to every *other* member, not
just to a nearest neighbor -- true here by construction, but a heatmap
shows what it looks like: the full pairwise predicted TM-score matrix for
bin `r top_05$bin`, same protein order as the table above.

```{r tm-heatmap-05, fig.height = 5, fig.width = 6}
Zb_05 <- Z[info_05$Protein, ]
S_05  <- 1 - as.matrix(proxy::dist(Zb_05, method = "cosine"))   # cosine == predicted TM-score
dimnames(S_05) <- list(info_05$Protein, info_05$Protein)           # kept protein-indexed, reused below

df_05 <- as.data.frame(as.table(S_05))
names(df_05) <- c("i", "j", "TM_hat")
df_05$i <- factor(df_05$i, levels = info_05$Protein, labels = seq_along(info_05$Protein))
df_05$j <- factor(df_05$j, levels = info_05$Protein, labels = seq_along(info_05$Protein))

ggplot(df_05, aes(i, j, fill = TM_hat)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0.5, 1), name = "predicted\nTM-score") +
  coord_fixed() +
  labs(x = "protein, table row above", y = "protein, table row above") +
  theme_minimal()
```

The floor is `r round(min(S_05[upper.tri(S_05)]), 3)`, comfortably above the 0.5
cutoff, and the diagonal is exactly 1 by construction (`cos(z, z) = 1`).

```{r dark-proteins-05}
dark_05      <- info_05$Protein[is.na(info_05$Genus) & is.na(info_05[["Annotation (Table 6)"]])]
annotated_05 <- info_05$Protein[!is.na(info_05[["Annotation (Table 6)"]])]

top_dark_rank_05 <- match(dark_05, info_05$Protein)
n_illustrate_05   <- min(2, length(dark_05))

nn_05 <- NULL
if (n_illustrate_05 > 0 && length(annotated_05) > 0) {
  nn_05 <- do.call(rbind, lapply(dark_05[seq_len(n_illustrate_05)], function(p) {
    best <- annotated_05[which.max(S_05[p, annotated_05])]
    data.frame(Protein = p, `Predicted TM-score` = round(S_05[p, best], 3),
               `Nearest annotated neighbor` = best,
               Annotation = info_05[["Annotation (Table 6)"]][info_05$Protein == best],
               check.names = FALSE)
  }))
  gt(nn_05)
}

illustration_05 <- if (is.null(nn_05)) {
  "there are none in this bin to illustrate with"
} else if (n_illustrate_05 == 1) {
  sprintf("its nearest annotated relative *within the bin* is %s, a *%s* -- predicted TM-score %s",
          nn_05$Protein[1], nn_05$Annotation[1], nn_05[["Predicted TM-score"]][1])
} else if (nn_05[["Nearest annotated neighbor"]][1] == nn_05[["Nearest annotated neighbor"]][2]) {
  sprintf("both of their nearest annotated relatives *within the bin* turn out to be the same protein, %s, a *%s* -- predicted TM-score %s to %s and %s to %s",
          nn_05[["Nearest annotated neighbor"]][1], nn_05$Annotation[1],
          nn_05[["Predicted TM-score"]][1], nn_05$Protein[1],
          nn_05[["Predicted TM-score"]][2], nn_05$Protein[2])
} else {
  sprintf("their nearest annotated relatives *within the bin* are two different proteins -- %s (a *%s*, predicted TM-score %s) for %s, and %s (a *%s*, predicted TM-score %s) for %s",
          nn_05[["Nearest annotated neighbor"]][1], nn_05$Annotation[1], nn_05[["Predicted TM-score"]][1], nn_05$Protein[1],
          nn_05[["Nearest annotated neighbor"]][2], nn_05$Annotation[2], nn_05[["Predicted TM-score"]][2], nn_05$Protein[2])
}
rank_desc_05 <- if (n_illustrate_05 == 0) "" else
  sprintf(" (row%s %s in the table above)", if (n_illustrate_05 > 1) "s" else "",
          paste(top_dark_rank_05[seq_len(n_illustrate_05)], collapse = " and "))
```

`r length(dark_05)` of the bin's `r length(members_05)` members have neither a Genus
call nor a Table 6 annotation. The `r n_illustrate_05` most
abundant of them`r rank_desc_05` illustrate what that buys: `r illustration_05`.

```{r dark-scope-05}
sig_members_05 <- names(roll_05$bin)[roll_05$bin %in% rownames(wald_05)[wald_05$reject]]

gt(data.frame(
  scope    = c("whole protein universe", "diet-significant bins only"),
  proteins = c(ncol(cnt_da), length(sig_members_05)),
  dark     = c(sum(is_dark(colnames(cnt_da))), sum(is_dark(sig_members_05))),
  pct_dark = round(100 * c(mean(is_dark(colnames(cnt_da))), mean(is_dark(sig_members_05))), 1)
))
```

### TM-score >= 0.7

```{r rollup-07}
roll_07 <- roll_both[["0.7"]]
c(proteins = ncol(cnt_da), bins = ncol(roll_07$counts))
```

Same model as the 0.5 pass above, on a different (stricter, more numerous) set of
bins.

`r sum(colSums(roll_07$counts > 0) < 3)` of `r ncol(roll_07$counts)` bins here are
nonzero in fewer than 3 of `r nrow(roll_07$counts)` samples -- `prev.filter` drops
them before fitting, same as above. A sparse bin is also where the
out-of-training-range embedding risk from the Embedding section matters most: a
wrongly placed long protein is an ordinary clustering error almost everywhere, but a
low-information bin it happens to dominate is a bin whose interpretation should not
be trusted -- not a defect in `repdist_bin()`, a limit inherited from
the trained TM-Vec head underneath it.

```{r linda-07}
fit_07 <- linda(t(roll_07$counts), md,
               formula = "~ diet + dose + sex + age_week + (1 | subject)",
               feature.dat.type = "count", prev.filter = 3 / nrow(roll_07$counts), verbose = FALSE)
```

```{r linda-wald-07}
L <- matrix(0, nrow = 5, ncol = length(fit_07$variables) + 1)
for (i in 1:5) L[i, i + 1] <- 1
wald_07 <- linda.wald.test(fit_07, L, model = "LMM", alpha = 0.05)
rownames(wald_07) <- rownames(fit_07$output[[1]])

gt(data.frame(bins_tested = nrow(wald_07), bins_diet_sig = sum(wald_07$reject),
              fraction_sig = round(mean(wald_07$reject), 3)))
```

`r round(100 * mean(wald_07$reject))`% of bins reject at `padj <= 0.05`.

```{r volcano-07, fig.height = 3.2, fig.width = 10}
diet_vars_07 <- grep("^diet", fit_07$variables, value = TRUE)
volc_07 <- do.call(rbind, lapply(diet_vars_07, function(v) {
  o <- fit_07$output[[v]]
  data.frame(bin = rownames(o), diet = sub("^diet", "", v),
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue, reject = o$reject)
}))

ggplot(volc_07, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"), guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) + theme_bw()
```

```{r top-bin-07}
top_07 <- volc_07[volc_07$reject, ]
top_07 <- top_07[which.max(abs(top_07$log2FoldChange)), ]

members_07 <- names(roll_07$bin)[roll_07$bin == top_07$bin]
ann_07 <- setNames(t6[["Consensus annotation"]], t6[["Protein Identifier"]])[members_07]
tx_07  <- as.data.frame(as(tax_table(psu), "matrix"))[members_07, c("Genus", "Species")]

ann_calls_07 <- unique(na.omit(ann_07))
ann_desc_07  <- if (length(ann_calls_07) == 0) {
  "none of them"
} else if (length(ann_calls_07) == 1) {
  sprintf("%d of them, all the same call: *%s*", sum(!is.na(ann_07)), ann_calls_07)
} else {
  sprintf("%d of them: %s", sum(!is.na(ann_07)), paste0("*", ann_calls_07, "*", collapse = ", "))
}
ann_plausibility_07 <- if (length(ann_calls_07) &&
                           all(grepl("glyco|hydrolase|GH[0-9]", ann_calls_07, ignore.case = TRUE)) &&
                           identical(top_07$diet, "Yeast")) {
  " -- glycoside hydrolases are plausible machinery for a yeast-cell-wall-heavy diet,"
} else ""
```

The single largest shift, in either direction, is bin `r top_07$bin` on
`r top_07$diet`: `r sprintf("%+.1f", top_07$log2FoldChange)` log2 fold change
against Casein at $p=$ `r formatC(top_07$pvalue, digits = 1, format = "e")`.
Its `r length(members_07)` members come from `r length(unique(na.omit(tx_07$Genus)))`
different genera. Extended Data Table 6 annotates `r ann_desc_07`
`r ann_plausibility_07` though this is one bin's annotation coverage, not a
pathway enrichment test.

```{r top-bin-table-07}
info_07 <- data.frame(
  Protein = members_07, Genus = tx_07$Genus, Species = tx_07$Species,
  `Annotation (Table 6)` = unname(ann_07),
  `Length (aa)` = Biostrings::width(refseq(psu)[members_07]),
  `Mean count, Casein` = round(colMeans(cnt_da[md$diet == "Casein", members_07, drop = FALSE]), 1),
  check.names = FALSE
)
info_07[[paste0("Mean count, ", top_07$diet)]] <-
  round(colMeans(cnt_da[md$diet == top_07$diet, members_07, drop = FALSE]), 1)
info_07 <- info_07[order(-info_07[[ncol(info_07)]]), ]
gt(info_07)
```

Every member sits at or near zero on `r if (top_07$log2FoldChange > 0) "Casein" else top_07$diet`
and climbs on `r if (top_07$log2FoldChange > 0) top_07$diet else "Casein"`.

```{r tm-heatmap-07, fig.height = 5, fig.width = 6}
Zb_07 <- Z[info_07$Protein, ]
S_07  <- 1 - as.matrix(proxy::dist(Zb_07, method = "cosine"))
dimnames(S_07) <- list(info_07$Protein, info_07$Protein)

df_07 <- as.data.frame(as.table(S_07))
names(df_07) <- c("i", "j", "TM_hat")
df_07$i <- factor(df_07$i, levels = info_07$Protein, labels = seq_along(info_07$Protein))
df_07$j <- factor(df_07$j, levels = info_07$Protein, labels = seq_along(info_07$Protein))

ggplot(df_07, aes(i, j, fill = TM_hat)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0.7, 1), name = "predicted\nTM-score") +
  coord_fixed() +
  labs(x = "protein, table row above", y = "protein, table row above") +
  theme_minimal()
```

The floor is `r round(min(S_07[upper.tri(S_07)]), 3)`, comfortably above the 0.7
cutoff.

```{r dark-proteins-07}
dark_07      <- info_07$Protein[is.na(info_07$Genus) & is.na(info_07[["Annotation (Table 6)"]])]
annotated_07 <- info_07$Protein[!is.na(info_07[["Annotation (Table 6)"]])]

top_dark_rank_07 <- match(dark_07, info_07$Protein)
n_illustrate_07   <- min(2, length(dark_07))

nn_07 <- NULL
if (n_illustrate_07 > 0 && length(annotated_07) > 0) {
  nn_07 <- do.call(rbind, lapply(dark_07[seq_len(n_illustrate_07)], function(p) {
    best <- annotated_07[which.max(S_07[p, annotated_07])]
    data.frame(Protein = p, `Predicted TM-score` = round(S_07[p, best], 3),
               `Nearest annotated neighbor` = best,
               Annotation = info_07[["Annotation (Table 6)"]][info_07$Protein == best],
               check.names = FALSE)
  }))
  gt(nn_07)
}

illustration_07 <- if (is.null(nn_07)) {
  "there are none in this bin to illustrate with"
} else if (n_illustrate_07 == 1) {
  sprintf("its nearest annotated relative *within the bin* is %s, a *%s* -- predicted TM-score %s",
          nn_07$Protein[1], nn_07$Annotation[1], nn_07[["Predicted TM-score"]][1])
} else if (nn_07[["Nearest annotated neighbor"]][1] == nn_07[["Nearest annotated neighbor"]][2]) {
  sprintf("both of their nearest annotated relatives *within the bin* turn out to be the same protein, %s, a *%s* -- predicted TM-score %s to %s and %s to %s",
          nn_07[["Nearest annotated neighbor"]][1], nn_07$Annotation[1],
          nn_07[["Predicted TM-score"]][1], nn_07$Protein[1],
          nn_07[["Predicted TM-score"]][2], nn_07$Protein[2])
} else {
  sprintf("their nearest annotated relatives *within the bin* are two different proteins -- %s (a *%s*, predicted TM-score %s) for %s, and %s (a *%s*, predicted TM-score %s) for %s",
          nn_07[["Nearest annotated neighbor"]][1], nn_07$Annotation[1], nn_07[["Predicted TM-score"]][1], nn_07$Protein[1],
          nn_07[["Nearest annotated neighbor"]][2], nn_07$Annotation[2], nn_07[["Predicted TM-score"]][2], nn_07$Protein[2])
}
rank_desc_07 <- if (n_illustrate_07 == 0) "" else
  sprintf(" (row%s %s in the table above)", if (n_illustrate_07 > 1) "s" else "",
          paste(top_dark_rank_07[seq_len(n_illustrate_07)], collapse = " and "))
```

`r length(dark_07)` of the bin's `r length(members_07)` members have neither a Genus
call nor a Table 6 annotation -- invisible to a lineage-based method and to a
profile-based one, grouped here on structure alone. The `r n_illustrate_07` most
abundant of them`r rank_desc_07` illustrate what that buys: `r illustration_07`.
That is not a function call, just a specific, testable hypothesis where Table 6
offers none at all -- exactly the gap a sequence-only annotation pipeline leaves
open and a structural ground metric can speak to.

```{r dark-scope-07}
sig_members_07 <- names(roll_07$bin)[roll_07$bin %in% rownames(wald_07)[wald_07$reject]]

gt(data.frame(
  scope    = c("whole protein universe", "diet-significant bins only"),
  proteins = c(ncol(cnt_da), length(sig_members_07)),
  dark     = c(sum(is_dark(colnames(cnt_da))), sum(is_dark(sig_members_07))),
  pct_dark = round(100 * c(mean(is_dark(colnames(cnt_da))), mean(is_dark(sig_members_07))), 1)
))
```

Proteins with neither a Genus call nor a Table 6 annotation are about as
common among the proteins LinDA flagged as diet-associated (at either threshold) as
they are everywhere else in the dataset -- being biologically important here (large
fold change, high significance) bought them no better odds of being
annotated. A structural ground metric does not fix the annotation gap, but
it does not inherit it either: `repdist_bin()` places every protein
in a bin whether or not it has a database hit, and a bin's annotated
members give its unannotated ones a specific, checkable neighbor instead
of silence.

## Conclusion

All three distances find dietary protein source highly significant, with diet
$R^2$ of `r sprintf("%.3f", r2["Bray-Curtis"])` for Bray-Curtis,
`r sprintf("%.3f", r2["Jaccard"])` for Jaccard, and
`r sprintf("%.3f", r2["w-UniFrac (structure)"])` for the representation — the representation
now reads more diet signal than Bray-Curtis, not just as much of it. The gain is
real but not large, and the reason it exists at all is measurable: these samples
share accessions, so protein identity, weighted or not, was already a good
feature here.

Two more results separate the representation from the other two. Bray-Curtis
fails the dispersion check while the representation passes, so part of the
Bray-Curtis effect is diets differing in how variable they are rather than in
where they sit. And once accession sharing is removed, both protein-identity
distances drop to the null while the representation is untouched.

The honest reading is that the representation buys robustness and, here, a little
power too. It keeps working where Bray-Curtis cannot — no shared accessions, no
alignment, no tree, no annotation.

What that robustness is not, on this dataset, is demonstrated *structural* information
specifically. The null controls above clear a random-projection null and a length-only
baseline, but not amino-acid composition, which explains more diet variance
(`r sprintf("%.3f", r2_comp)`) than the representation itself
(`r sprintf("%.3f", r2["w-UniFrac (structure)"])`). Everything downstream that reads as
"structural" -- the PCoA separation, the `repdist_bin()` bins, the
glycan-degrading bins `r top_05$bin` (TM-score >= 0.5) and `r top_07$bin` (>= 0.7) --
is equally consistent with TM-Vec's geometry tracking amino-acid
composition rather than fold. This report does not have the test that would separate
those two explanations; that needs cases where composition and structure disagree, and
this real metaproteome does not obviously supply enough of them.

What this report does not settle is alpha diversity. Rarefying equalises depth, not
evenness, and any abundance-weighted set distance can respond to how concentrated a
sample is as well as to what is in it. Separating the two needs evenness matching or
a support-corrected estimator, neither of which is done here.
