---
title: "Re analysis of diet protein study"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE,
                      fig.width = 8, fig.height = 4, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

The simulation study showed that a protein-identity distance goes blind when
samples share no accessions, and that a TM-Vec ground metric does not. This report
applies the same comparison to real data: 111 mouse faecal metaproteomes across
eight dietary protein sources (Blakeley-Ruiz *et al.*, *ISME J* 19(1) wraf048),
quantified by spectral counting.

There is no weighted-UniFrac arm. The simulation could align 480 proteins from four
families and give it a ground metric; here the universe is ~19,000 proteins from a
whole gut community, mostly non-homologous, and there is no alignment to build a
tree from.

```{r libs, cache = FALSE}
Sys.setenv(KMP_DUPLICATE_LIB_OK = "TRUE")
reticulate::use_python(Sys.which("python3"), required = TRUE)

library(phyloseq)
library(vegan)
library(ggplot2)
library(gt)
library(here)
library(MicrobiomeStat)
for (f in list.files("R", full.names = TRUE)) source(f)

path  <- here("data", "test_study")
NPERM <- 9999
```

## Data

```{r data}
ps <- readRDS(file.path(path, "ps.rds"))   # built and validated in data_curation.Rmd
ps
```

## Rarefaction

```{r depth-check, fig.height = 3}
floor_depth <- min(sample_sums(ps))    # rarefy to the shallowest sample, dropping none

ggplot(data.frame(depth = sort(sample_sums(ps))), aes(seq_along(depth), depth)) +
  geom_col(width = 1) +
  geom_hline(yintercept = floor_depth, linetype = "dashed", color = "firebrick") +
  labs(x = "sample, sorted by depth", y = "PSMs") + theme_bw()
```

The shallowest samples sit within a few percent of each other, so there is no isolated
outlier to drop and the cutoff (dashed line) is simply the minimum depth,
`r floor_depth` PSMs. Taking it from `min()` rather than hardcoding it is what keeps
that sentence true: any fixed number above the minimum silently discards the samples
below it.

```{r rarefy}
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)

# T0 has no protein source, so the diet term is undefined for it
psd <- prune_samples(sample_data(psd)$phase != "baseline", psd)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)
sample_data(psd)$diet <- factor(sample_data(psd)$diet)
sample_data(psd)$dose <- factor(sample_data(psd)$dose)
psd
```

Rarefying before pruning matters: rarefaction subsamples the raw counts, so it has
to see the raw counts.

## Embedding

TM-Vec1, the ground metric chosen in `docs/ground_metric_models.md`: its head was
trained to regress the TM-score of protein pairs, so cosine similarity between two
vectors is a predicted TM-score. Sequences are truncated at 1022 residues, TM-Vec1's
own limit from its ProtT5 encoder. The result is cached.

```{r embed, cache = FALSE}
# cache = FALSE on purpose: this chunk's own code never changes, so knitr's
# chunk cache would not notice that `psd` changed upstream (e.g. a different
# rarefaction depth) and would silently replay a stale Z. The rds file below is
# the only cache, keyed by protein id, and it lives in data/ rather than in the
# knitr cache directory. Filename is model-specific (not "embeddings_tmvec.rds")
# so switching models can't silently reuse the previous model's vectors via the
# `need <- setdiff(...)` resume logic below.
cache <- file.path(path, "embeddings_tmvec1.rds")
Zall  <- if (file.exists(cache)) readRDS(cache) else NULL
need  <- setdiff(taxa_names(psd), rownames(Zall))

if (length(need)) {
  s <- substr(as.character(refseq(psd)[need]), 1, 1022)
  s <- s[order(nchar(s))]                       # length-sorted: keeps batches tight
  # Save every 2000 proteins rather than once at the end. The full pass is about
  # an hour and the cache is keyed by protein id, so an interrupted render resumes
  # from the last block instead of starting over.
  for (i in split(seq_along(s), ceiling(seq_along(s) / 2000))) {
    Zall <- rbind(Zall, embed_proteins(s[i], batch_size = 16))
    saveRDS(Zall, cache)
  }
}
Z <- Zall[taxa_names(psd), ]
```

## Results

Bray-Curtis and Jaccard are both protein-identity distances -- two proteins are
either the same accession or maximally different -- and differ only in whether
abundance enters: Bray-Curtis weights by how much of each protein was detected,
Jaccard only by whether it was detected at all.

```{r distances}
cnt <- t(as(otu_table(psd), "matrix"))     # samples in rows
md  <- data.frame(sample_data(psd))
D <- list(`Bray-Curtis` = phyloseq::distance(psd, method = "bray"),
          `Jaccard`     = vegdist(cnt, method = "jaccard", binary = TRUE),
          `weightedRep` = weightedRep(cnt, Z, method = "mmd"))
```

Amount of protein, sex and age stay in the model as covariates; only the diet term
is reported. Permutations are blocked within mouse, since each mouse contributes
several samples. `sex` is constant within a mouse, so it has no valid permutation
distribution under that blocking — it absorbs variance rather than being tested.

```{r permanova}
diet_test <- function(dl) do.call(rbind, lapply(names(dl), function(n) {
  set.seed(1)
  a <- adonis2(dl[[n]] ~ diet + dose + sex + age_week, data = md, by = "terms",
               permutations = permute::how(blocks = factor(md$subject), nperm = NPERM))
  data.frame(distance = n, R2 = round(a["diet", "R2"], 3),
             F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"])
}))
res <- diet_test(D)
r2  <- setNames(res$R2, res$distance)   # name-based lookup for the prose below
gt(res)
```

```{r pcoa, cache = FALSE, fig.height = 4.5}
for (n in names(D)) {
  print(plot_ordination(psd, ordinate(psd, "PCoA", distance = D[[n]]), color = "diet") +
          stat_ellipse(type = "t", geom = "polygon", alpha = 0) +
          ggtitle(n) + theme_bw())
}
```

## Diagnostics

A PERMANOVA $p$ can come from groups differing in *spread* rather than in
*location*, and the test cannot separate the two on its own. `betadisper` can.

```{r diagnostics}
gt(data.frame(
  distance     = names(D),
  betadisper_p = sapply(D, function(d) round(anova(betadisper(d, md$diet))$`Pr(>F)`[1], 4)),
  row.names = NULL))
```

## What Bray-Curtis and Jaccard are actually using

All three distances find a highly significant diet effect and land in the same
broad range, which the simulation would not have predicted. The reason is that
these are cage-housed lab mice searched against one protein database: samples
really do share *the same accessions*, and exact accession identity is something
Bray-Curtis and Jaccard can read directly. Jaccard's lower $R^2$ against
Bray-Curtis's is the presence/absence penalty: it throws away how much of each
shared protein was detected, keeping only whether it was detected at all.

The test is to take that away. Give every sample its own private accession for each
protein it detected. Counts, sequences and abundances are all unchanged — only the
labels differ — so no two samples share an accession any more.

```{r shuffle}
nz  <- which(cnt > 0, arr.ind = TRUE)
lab <- paste0("protein", seq_len(nrow(nz)))   # masked: one private label per (sample, protein) detection

cnt_s <- matrix(0L, nrow(cnt), length(lab), dimnames = list(rownames(cnt), lab))
cnt_s[cbind(nz[, "row"], seq_len(nrow(nz)))] <- cnt[nz]
Z_s <- Z[nz[, "col"], ]
rownames(Z_s) <- lab

c(accessions_before = ncol(cnt), accessions_after = ncol(cnt_s),
  max_samples_sharing_one = max(colSums(cnt_s > 0)))
```

```{r shuffle-test}
res_s <- diet_test(list(`Bray-Curtis` = vegdist(cnt_s / rowSums(cnt_s), "bray"),
                        `Jaccard`     = vegdist(cnt_s, method = "jaccard", binary = TRUE),
                        `weightedRep` = weightedRep(cnt_s, Z_s, method = "mmd")))
gt(res_s)
```

Bray-Curtis and Jaccard both collapse to the exact null. Every pair is now
maximally dissimilar, so each distance matrix is constant and cannot express any
grouping: $F = 1$, $p = 1$, and $R^2 = 5/98 = 0.051$, which is simply the diet
term's degrees of freedom over $n-1$ — the share any 5-df term receives when there
is nothing to explain. `weightedRep()` returns exactly what it returned above,
because a protein's embedding does not depend on what its accession is called and
the abundance-weighted mean is therefore unchanged.

So Bray-Curtis's and Jaccard's whole result here rests on samples sharing literal
accessions. That is a fair assumption within one colony, one database and one
assembly. It stops being fair across studies, colonies or assemblies, where the
same function arrives under a different accession — which is the case the
representation distance is built for.

## Differential abundance

PERMANOVA tests whether diet structures the community as a whole; it says
nothing about which proteins carry that signal. Testing all `r ncol(cnt)`
proteins directly would mean that many near-independent tests on units that
mostly aren't independent -- protein groups are frequently homologous
(`docs/method_notes.md` §1), so a chunk of those "tests" are the same
signal counted several times. `rollupStructural()` collapses proteins into
bins where every member
has predicted TM-score >= 0.5 to every other member -- the same-fold
threshold from Xu & Zhang (2010), already used above -- before any testing
happens, so multiplicity is paid once per structural unit, not once per
accession.

```{r rollup}
roll <- rollupStructural(cnt, Z, cluster_threshold = 0.5)
c(proteins = ncol(cnt), bins = ncol(roll$counts))
```

LinDA (Zhou *et al.*, *Genome Biology* 2022) fits a linear model per bin on
centered-log-ratio-transformed abundance and corrects the compositional bias
a plain `lm()` on relative abundance would carry -- the same reason
`weightedRep()` works on abundance weights rather than raw proportions
directly. Its formula interface takes a random effect for repeated sampling
of the same mouse, the model-based counterpart to the permutation blocking
`adonis2` used above.

No prevalence filter -- every one of the `r ncol(roll$counts)` bins is
tested, including the ones with only a handful of nonzero samples. LinDA
warns about those rather than dropping them; the mixed model can also
report a singular fit for a bin (its between-mouse variance estimated at
zero), which `lmer` warns about but does not treat as an error.

```{r linda}
fit <- linda(t(roll$counts), md,
            formula = "~ diet + dose + sex + age_week + (1 | subject)",
            feature.dat.type = "count", verbose = FALSE)
```

As with the PERMANOVA table, only the diet term is reported: a joint Wald
test on all five diet coefficients per bin, LinDA's counterpart to
`adonis2`'s `by = "terms"`.

```{r linda-wald}
L <- matrix(0, nrow = 5, ncol = length(fit$variables) + 1)
for (i in 1:5) L[i, i + 1] <- 1   # intercept + the 5 diet dummy columns
wald <- linda.wald.test(fit, L, model = "LMM", alpha = 0.05)
rownames(wald) <- rownames(fit$output[[1]])   # linda.wald.test drops them; reused below

gt(data.frame(bins_tested = nrow(wald), bins_diet_sig = sum(wald$reject),
              fraction_sig = round(mean(wald$reject), 3)))
```

`r round(100 * mean(wald$reject))`% of bins reject at `padj <= 0.05` --
consistent with the PERMANOVA result above: diet is not a handful of
marker proteins, it reorganizes abundance broadly across the structural
repertoire. That joint test has no size to it, though -- diet has five
levels and no single number is "the" effect. A volcano plot needs one
effect size per point, so the five diet-vs-Casein contrasts already inside
`fit` are shown separately: one panel per diet, log2 fold change against
Casein on the x-axis, each point one bin, red where that specific contrast
rejects at `padj <= 0.05`.

```{r volcano, fig.height = 3.2, fig.width = 10}
diet_vars <- grep("^diet", fit$variables, value = TRUE)
volc <- do.call(rbind, lapply(diet_vars, function(v) {
  o <- fit$output[[v]]
  data.frame(bin = rownames(o), diet = sub("^diet", "", v),
             log2FoldChange = o$log2FoldChange, pvalue = o$pvalue, reject = o$reject)
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"), guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) + theme_bw()
```

```{r top-bin}
top <- volc[volc$reject, ]
top <- top[which.max(abs(top$log2FoldChange)), ]

members <- names(roll$bin)[roll$bin == top$bin]
t6 <- read.delim(file.path(path, "annotated", "ExtendedDataTable6AnnotatedMicrobialMetaproteomes.txt"),
                 skip = 2, quote = "\"", check.names = FALSE, colClasses = "character",
                 comment.char = "", fileEncoding = "latin1")
ann <- setNames(t6[["Consensus annotation"]], t6[["Protein Identifier"]])[members]
tx  <- as.data.frame(as(tax_table(psd), "matrix"))[members, c("Genus", "Species")]
```

The single largest shift, in either direction, is bin `r top$bin` on
`r top$diet`: `r sprintf("%+.1f", top$log2FoldChange)` log2 fold change
against Casein at $p=$ `r formatC(top$pvalue, digits = 1, format = "e")`.
Its `r length(members)` members come from `r length(unique(na.omit(tx$Genus)))`
different genera -- proteins TM-Vec1 grouped as one predicted fold despite
sharing no organism, the case an accession-identity distance cannot follow
through a diet shift at all. Extended Data Table 6 annotates
`r sum(!is.na(ann))` of them, and every annotated one is a glycan-degrading
enzyme -- alpha-L-fucosidases and GH65/GH92 glycoside hydrolases, plausible
machinery for a yeast-cell-wall-heavy diet, though this is one bin's
annotation coverage, not a pathway enrichment test.

```{r top-bin-table}
info <- data.frame(
  Protein = members, Genus = tx$Genus, Species = tx$Species,
  `Annotation (Table 6)` = unname(ann),
  `Length (aa)` = Biostrings::width(refseq(psd)[members]),
  `Mean count, Casein` = round(colMeans(cnt[md$diet == "Casein", members, drop = FALSE]), 1),
  check.names = FALSE
)
info[[paste0("Mean count, ", top$diet)]] <-
  round(colMeans(cnt[md$diet == top$diet, members, drop = FALSE]), 1)
info <- info[order(-info[[ncol(info)]]), ]   # re-used below, so the sort has to stick
gt(info)
```

Every member sits at or near zero on Casein and climbs on `r top$diet`,
which is what the fold change states in raw counts -- not an artifact of
the log-ratio transform.

`rollupStructural()`'s complete-linkage guarantee is that every pair inside
a bin has predicted TM-score >= `cluster_threshold` to every *other* member, not
just to a nearest neighbor -- true here by construction, but a heatmap
shows what it looks like: the full pairwise predicted TM-score matrix for
bin9, same protein order as the table above.

```{r tm-heatmap, fig.height = 5, fig.width = 6}
Zb <- Z[info$Protein, ]
S  <- 1 - as.matrix(proxy::dist(Zb, method = "cosine"))   # cosine == predicted TM-score
dimnames(S) <- list(info$Protein, info$Protein)           # kept protein-indexed, reused below

df <- as.data.frame(as.table(S))
names(df) <- c("i", "j", "TM_hat")
df$i <- factor(df$i, levels = info$Protein, labels = seq_along(info$Protein))
df$j <- factor(df$j, levels = info$Protein, labels = seq_along(info$Protein))

ggplot(df, aes(i, j, fill = TM_hat)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0.5, 1), name = "predicted\nTM-score") +
  coord_fixed() +
  labs(x = "protein, table row above", y = "protein, table row above") +
  theme_minimal()
```

The floor is `r round(min(S[upper.tri(S)]), 3)`, comfortably above the 0.5
cutoff, and the diagonal is exactly 1 by construction (`cos(z, z) = 1`).
The brighter blocks are sub-structure *within* the bin -- proteins closer
to a TM-score of 1 with each other than with the rest of the bin -- but
nothing drops below where complete linkage promised it wouldn't.

Rows 1 and 2 stand out for a different reason: they are the two most
abundant members of the whole bin, and also two of the ones with neither a
Genus call nor a Table 6 annotation -- invisible to a lineage-based method
and to a profile-based one, grouped here on structure alone.

```{r dark-proteins}
dark      <- info$Protein[is.na(info$Genus) & is.na(info[["Annotation (Table 6)"]])]
annotated <- info$Protein[!is.na(info[["Annotation (Table 6)"]])]

nn <- do.call(rbind, lapply(dark[1:2], function(p) {
  best <- annotated[which.max(S[p, annotated])]
  data.frame(Protein = p, `Predicted TM-score` = round(S[p, best], 3),
             `Nearest annotated neighbor` = best,
             Annotation = info[["Annotation (Table 6)"]][info$Protein == best],
             check.names = FALSE)
}))
gt(nn)
```

Both of their nearest annotated relatives *within the bin* turn out to be
the same protein: `r nn[["Nearest annotated neighbor"]][1]`, a
*`r nn$Annotation[1]`* from a different genus entirely -- predicted TM-score
`r nn[["Predicted TM-score"]][1]` to `r nn$Protein[1]` and
`r nn[["Predicted TM-score"]][2]` to `r nn$Protein[2]`. That is not a
function call, just a specific, testable hypothesis where Table 6 offers
none at all -- exactly the gap a sequence-only annotation pipeline leaves
open and a structural ground metric can speak to.

bin9's two dark proteins are not a cherry-picked pair. Annotation-dependent
tools -- eggNOG-mapper, Pfam/HMMER, and whatever built Table 6's "Consensus
annotation" column here -- only speak for a protein with a detectable match
to something already characterized; the rest usually just drop out of any
function-level discussion, not because they don't matter, but because
there is nothing to put in the annotation column. That gap is not rare
here, and it is not concentrated in the boring part of the data:

```{r dark-scope}
is_dark <- function(ids) {
  a <- setNames(t6[["Consensus annotation"]], t6[["Protein Identifier"]])[ids]
  g <- as.data.frame(as(tax_table(psd), "matrix"))[ids, "Genus"]
  is.na(a) & is.na(g)
}
sig_members <- names(roll$bin)[roll$bin %in% rownames(wald)[wald$reject]]

gt(data.frame(
  scope    = c("whole protein universe", "diet-significant bins only"),
  proteins = c(ncol(cnt), length(sig_members)),
  dark     = c(sum(is_dark(colnames(cnt))), sum(is_dark(sig_members))),
  pct_dark = round(100 * c(mean(is_dark(colnames(cnt))), mean(is_dark(sig_members))), 1)
))
```

Proteins with neither a Genus call nor a Table 6 annotation are about as
common among the proteins LinDA flagged as diet-associated as they are
everywhere else in the dataset -- being biologically important here (large
fold change, high significance) bought them no better odds of being
annotated. A structural ground metric does not fix the annotation gap, but
it does not inherit it either: `rollupStructural()` places every protein
in a bin whether or not it has a database hit, and a bin's annotated
members give its unannotated ones a specific, checkable neighbor instead
of silence.

## Conclusion

All three distances find dietary protein source highly significant, with diet
$R^2$ of `r sprintf("%.3f", r2["Bray-Curtis"])` for Bray-Curtis,
`r sprintf("%.3f", r2["Jaccard"])` for Jaccard, and
`r sprintf("%.3f", r2["weightedRep"])` for the representation — the representation
now reads more diet signal than Bray-Curtis, not just as much of it. The gain is
real but not large, and the reason it exists at all is measurable: these samples
share accessions, so protein identity, weighted or not, was already a good
feature here.

Two more results separate the representation from the other two. Bray-Curtis
fails the dispersion check while the representation passes, so part of the
Bray-Curtis effect is diets differing in how variable they are rather than in
where they sit. And once accession sharing is removed, both protein-identity
distances drop to the null while the representation is untouched.

The honest reading is that the representation buys robustness and, here, a little
power too. It keeps working where Bray-Curtis cannot — no shared accessions, no
alignment, no tree, no annotation.

What this report does not settle is alpha diversity. Rarefying equalises depth, not
evenness, and any abundance-weighted set distance can respond to how concentrated a
sample is as well as to what is in it. Separating the two needs evenness matching or
a support-corrected estimator, neither of which is done here.
