---
title: "Full case study: homology-aware beta diversity with repdist"
author: "Yixuan Yang"
package: repdist
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

# Why a representation distance

Bray-Curtis, Jaccard and their relatives compare samples through protein
*identity*: two proteins are either the same accession or maximally different.
Metaproteomes break that assumption. Two communities can carry the same
enzymatic function on accessions that never overlap, and an identity-based
distance cannot tell that case apart from any other difference.

`repdist` replaces identity with distance in a protein language model's
representation space. Each sample becomes an abundance-weighted distribution
over protein embeddings, and two samples are compared by maximum mean
discrepancy (MMD) under an RBF kernel.

This report follows the package vignette's workflow on the complete retained
protein set:

1. **Beta diversity** with MMD and PERMANOVA.
2. **Structural bins** at a predicted TM-score of 0.7.
3. **Differential abundance** on the bins with DESeq2.
4. **Functional reading** of the most responsive annotated bins.

```{r libs}
library(repdist)
library(ggplot2)
library(phyloseq)
```

# The full dataset

The study is Blakeley-Ruiz *et al.*, *ISME J* **19**(1) wraf048: mouse faecal
metaproteomes across six dietary protein sources, quantified by spectral
counting.

```{r data}
path <- here::here("data", "test_study")
MAX_AA <- 3000L

ps <- readRDS(file.path(path, "ps.rds"))
floor_depth <- min(sample_sums(ps))

set.seed(1)
psd <- rarefy_even_depth(ps, sample.size = floor_depth, rngseed = 1,
                         replace = FALSE, verbose = FALSE)
psd <- prune_samples(sample_data(psd)$phase != "baseline", psd)
psd <- prune_taxa(taxa_sums(psd) > 0, psd)

n_after_rarefaction <- ntaxa(psd)
protein_length <- Biostrings::width(refseq(psd)[taxa_names(psd)])
too_long <- protein_length > MAX_AA
long_abundance <- sum(taxa_sums(psd)[too_long]) / sum(taxa_sums(psd))
psd <- prune_taxa(!too_long, psd)

keep <- taxa_names(psd)
psu <- prune_samples(sample_data(ps)$phase != "baseline", ps)
psu <- prune_taxa(taxa_names(psu) %in% keep, psu)

counts <- Matrix::Matrix(t(as(otu_table(psd), "matrix")), sparse = TRUE)
counts_raw <- t(as(otu_table(psu), "matrix"))[rownames(counts), keep, drop = FALSE]

md <- data.frame(sample_data(psd))[rownames(counts), ]
md$diet <- factor(md$diet)
md$dose <- factor(md$dose)
md$subject <- factor(md$subject)

t6 <- read.delim(file.path(path, "annotated",
                           "ExtendedDataTable6AnnotatedMicrobialMetaproteomes.txt"),
                 skip = 2, quote = "\"", check.names = FALSE,
                 colClasses = "character", comment.char = "", fileEncoding = "latin1")
tx <- as.data.frame(as(tax_table(psd), "matrix"))[keep, ]
taxa <- data.frame(
  protein = keep,
  genus = tx$Genus,
  species = tx$Species,
  annotation = unname(setNames(t6[["Consensus annotation"]],
                               t6[["Protein Identifier"]])[keep]),
  length = Biostrings::width(refseq(psd)[keep]),
  row.names = keep, stringsAsFactors = FALSE)
taxa$annotation[is.na(taxa$annotation) | !nzchar(trimws(taxa$annotation))] <- NA_character_

d <- list(counts = counts, counts_raw = counts_raw, metadata = md, taxa = taxa,
          sequences = setNames(as.character(refseq(psd)[keep]), keep))

c(samples = nrow(counts), proteins = ncol(counts),
  removed_over_3000aa = sum(too_long),
  removed_abundance_pct = round(100 * long_abundance, 3))
table(md$diet)
```

Rarefaction provides the abundance filter: proteins absent after subsampling
are removed. It retained `r n_after_rarefaction` proteins; the 3,000aa ceiling
then removed `r sum(too_long)` more, representing only
`r sprintf("%.3f", 100 * long_abundance)`% of rarefied abundance. The beta-diversity
matrix is sparse; `counts_raw` contains unrarefied counts for exactly the same
proteins and samples.

# Embedding

`embed_proteins()` embeds every sequence whole. The full TM-Vec embeddings are
cached outside the render because model inference is not part of the statistical
analysis.

```{r Z}
cached <- readRDS(file.path(path, "embeddings_tmvec1_large.rds"))$Z
stopifnot(all(keep %in% rownames(cached)))
Z <- cached[keep, , drop = FALSE]
dim(Z)
```

TM-Vec's head was trained to regress TM-score, so cosine similarity between two
embeddings is a predicted TM-score. `known_models()` lists the other registered
representations.

```{r models}
names(known_models())
```

# The protein ground distance

`repdist_matrix()` builds the condensed protein-by-protein cosine distance used
by binning and QC. This is the expensive object in the full analysis.

```{r ground}
G_cos <- repdist_matrix(Z)
```

`bin_proteins()` reads this scale as predicted TM-score. `sample_repdist()`
instead builds a Euclidean RBF internally so the kernel remains positive
semi-definite. On unit-norm TM-Vec embeddings the two ground distances are
monotone transforms of one another.

```{r scales}
c(median_predicted_TM = round(1 - median(G_cos), 3),
  pairs_above_0.7 = sum(G_cos <= 0.3))
```

# Step 1: beta diversity

Bray-Curtis is the protein-identity baseline. MMD asks how far apart different
proteins actually are. The sparse count matrix stays sparse throughout the MMD
calculation.

```{r distances}
D <- list(
  `Bray-Curtis` = vegan::vegdist(as.matrix(d$counts), "bray"),
  `MMD` = sample_repdist(d$counts, Z))
```

Both are abundance-weighted, so richness is checked before interpreting the
diet effect.

```{r qc-richness}
rich <- vegan::specnumber(as.matrix(d$counts))
round(vapply(D, function(x) cor(rich, rowMeans(as.matrix(x))), numeric(1)), 2)
```

## PERMANOVA

Amount of protein, sex and age remain as covariates and diet enters last.
Permutations are blocked within mouse.

```{r permanova}
perm <- permute::how(blocks = md$subject, nperm = 999)

res <- do.call(rbind, lapply(names(D), function(n) {
  set.seed(1)
  a <- vegan::adonis2(D[[n]] ~ dose + sex + age_week + diet, data = md,
                      by = "terms", permutations = perm)
  data.frame(distance = n, R2 = round(a["diet", "R2"], 3),
             F = round(a["diet", "F"], 2), p = a["diet", "Pr(>F)"])
}))
knitr::kable(res, row.names = FALSE)
```

## Ordination

```{r pcoa, fig.height = 3.6, fig.width = 6.5, fig.alt = "PCoA under Bray-Curtis and RBF-MMD, coloured by dietary protein source"}
ord <- do.call(rbind, lapply(names(D), function(n) {
  e <- cmdscale(D[[n]], k = 2, eig = TRUE)
  rel <- 100 * e$eig / sum(e$eig[e$eig > 0])
  data.frame(
    Axis1 = e$points[, 1], Axis2 = e$points[, 2],
    diet = md[rownames(e$points), "diet"],
    panel = sprintf("%s  (PCo1 %.1f%%, PCo2 %.1f%%)", n, rel[1], rel[2]))
}))

ggplot(ord, aes(Axis1, Axis2, colour = diet)) +
  geom_point(size = 1.6, alpha = 0.85) +
  stat_ellipse(type = "t", linewidth = 0.3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "PCo1", y = "PCo2", colour = "diet") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Step 2: structural bins at TM-score 0.7

`bin_proteins()` uses complete linkage, so every pair inside a bin meets the
threshold. A 0.7 cut is stricter than the conventional same-fold boundary of
0.5: bins are tighter, at the cost of more singletons.

```{r bins}
b <- bin_proteins(G_cos, min_sim = 0.7)
```

```{r qc-bins, fig.height = 5, fig.alt = "Nearest-neighbour similarity and structural-bin profiles"}
plot_similarity_profile(G_cos, min_sim = 0.7)
plot_bin_profile(b, min_sim = 0.7)
```

The unrarefied counts roll up to bin level with one `rowsum()`.

```{r bin-counts}
bin <- setNames(rep(names(b$clusters), lengths(b$clusters)), unlist(b$clusters))
bin_counts <- t(rowsum(t(d$counts_raw), bin[colnames(d$counts_raw)]))
c(proteins = ncol(d$counts_raw), bins = ncol(bin_counts))
```

```{r bin-sizes}
sizes <- lengths(b$clusters)
table(cut(sizes, c(0, 1, 2, 5, 10, Inf),
          labels = c("1", "2", "3-5", "6-10", ">10")))
```

`r sum(sizes == 1)` proteins remain singleton bins. The other
`r sum(sizes > 1)` bins combine structurally related accessions, reducing the
number of tested units by a factor of
`r round(ncol(d$counts_raw) / ncol(bin_counts), 1)`.

# Step 3: differential abundance with DESeq2

DESeq2 fits a negative-binomial model per bin on unrarefied counts. The paired
design blocks on subject, and diet is tested by likelihood ratio against the
model without diet.

```{r deseq}
dds <- DESeq2::DESeqDataSetFromMatrix(t(bin_counts), md,
                                      design = ~ subject + diet)
dds <- DESeq2::DESeq(dds, test = "LRT", reduced = ~ subject,
                     sfType = "poscounts", quiet = TRUE)
res_lrt <- DESeq2::results(dds)

c(bins_tested = nrow(res_lrt),
  diet_significant = sum(res_lrt$padj <= 0.05, na.rm = TRUE))
```

Each significant bin is assigned its largest shrunken diet-vs-Casein fold
change for ranking and display.

```{r effects}
lev <- setdiff(levels(md$diet), "Casein")
ddsw <- DESeq2::nbinomWaldTest(dds, quiet = TRUE)
shrunk <- lapply(lev, function(l)
  DESeq2::lfcShrink(ddsw, coef = paste0("diet_", l, "_vs_Casein"),
                    type = "normal", quiet = TRUE))
names(shrunk) <- lev
fc <- vapply(shrunk, function(o) o$log2FoldChange, numeric(nrow(res_lrt)))
rownames(fc) <- rownames(res_lrt)

sig <- rownames(res_lrt)[!is.na(res_lrt$padj) & res_lrt$padj <= 0.05]
top <- data.frame(
  bin = sig,
  members = sizes[sig],
  padj = formatC(res_lrt[sig, "padj"], digits = 2, format = "e"),
  diet = lev[apply(abs(fc[sig, , drop = FALSE]), 1, which.max)],
  log2FC = round(fc[cbind(sig, lev[apply(abs(fc[sig, , drop = FALSE]),
                                        1, which.max)])], 2))
top <- top[order(-abs(top$log2FC)), ]
knitr::kable(head(top, 8), row.names = FALSE)
```

```{r volcano, fig.height = 3.4, fig.width = 9, fig.alt = "Diet-vs-Casein volcano plots for structural bins"}
volc <- do.call(rbind, lapply(lev, function(l) {
  o <- shrunk[[l]]
  data.frame(bin = rownames(o), diet = l, log2FoldChange = o$log2FoldChange,
             pvalue = o$pvalue, reject = !is.na(o$padj) & o$padj <= 0.05)
}))

ggplot(volc, aes(log2FoldChange, -log10(pvalue), color = reject)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ diet, nrow = 1) +
  scale_color_manual(values = c(`FALSE` = "grey70", `TRUE` = "firebrick"),
                     guide = "none") +
  labs(x = "log2 fold change vs Casein", y = expression(-log[10](p))) +
  theme_bw()
```

# Step 4: responsive bins, read functionally

Select the three most responsive multi-member bins with at least two annotated
proteins.

```{r pick}
n_ann <- vapply(b$clusters[top$bin],
                function(m) sum(!is.na(d$taxa[m, "annotation"])), integer(1))
picks <- head(top[top$members > 1 & n_ann >= 2, ], 3)
knitr::kable(picks, row.names = FALSE)
```

```{r members}
info <- do.call(rbind, lapply(seq_len(nrow(picks)), function(i) {
  m <- b$clusters[[picks$bin[i]]]
  data.frame(
    bin = picks$bin[i],
    genus = d$taxa[m, "genus"],
    annotation = substr(d$taxa[m, "annotation"], 1, 52),
    mean_Casein = round(colMeans(d$counts_raw[md$diet == "Casein", m,
                                               drop = FALSE]), 1),
    mean_top = round(colMeans(d$counts_raw[md$diet == picks$diet[i], m,
                                           drop = FALSE]), 1))
}))
knitr::kable(info, row.names = FALSE)
```

The complete-linkage guarantee can be inspected directly for the leading
annotated bin.

```{r floor, fig.height = 4.5, fig.alt = "Pairwise predicted TM-score heatmap for the leading annotated bin"}
plot_bin_similarity(b, picks$bin[1], G_cos)
```

# Practical notes

**Rarefy first** for whole-community distances. The differential-abundance
model uses the corresponding unrarefied counts and handles depth through size
factors.

**Memory.** `repdist_matrix()` remains quadratic because binning needs the
condensed protein distance. `sample_repdist()` does not: it accumulates
two-dimensional 1000-by-1000 kernel tiles directly into the sample Gram matrix,
and sparse counts remain sparse.

**Bandwidth.** `sample_repdist()` returns its bandwidth as `attr(, "sigma")`;
pass it back when comparisons across studies need one fixed kernel.

```{r session}
sessionInfo()
```
