---
title: "Data curation: plastic-substrate metaproteome to a phyloseq object"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE, fig.width = 7)
knitr::opts_knit$set(root.dir = "..")   # this file lives in analysis/
```

```{r libs, cache = FALSE, message = FALSE}
library(readxl)
library(phyloseq)
library(Biostrings)
devtools::load_all(".", quiet = TRUE)
HERE <- file.path("data", "case_study2")
```

Source: a marine plastic-incubation metaproteome. Eight substrates -- four
plastics (PHBV, PLA, HDPE, a PLA/PCL blend), two formulations of PHBH ("PHBH"
and "PHBH Green"), filter paper (a readily-biodegradable cellulose control),
and sediment (the ambient community, no added substrate) -- three replicate
incubations each, 24 samples. Two abundance layers exist for this study:
the workbook's own **%NSAF** (normalized spectral abundance factor, no raw
counts behind it) and a separately supplied **PUP** table that *is* the raw
count layer %NSAF was normalized from. PUP is what this curation uses, so
that rarefaction -- not available for the %NSAF-only first pass of this
report -- is available downstream in `case_study2.Rmd`.

# Reading the tables

One label row, a header on row 2, then one row per protein. A trailing block
of ~1000 rows is a KO-term lookup legend (`MantisKO`/`MicrobeKO` reference
entries with no protein data attached, separated from the real table by a
blank row); `readxl::read_excel()` stops at that blank row on its own, so no
manual filtering is needed for it. The workbook is read only for its
annotation columns and its `Protein ID` join key now -- abundance comes from
`ProteomeMeanPUP.csv` below.

```{r read}
raw <- read_excel(file.path(HERE,
  "ProteomeMeanPUPNSAFPlastic_DepolyAB_TIGERFAM_FinalAnot_Master_MantisSorted_Current.xlsx"),
  skip = 1, .name_repair = "unique_quiet")

c(annotation_rows = nrow(raw))
```

# Protein IDs

Contig-derived protein ids (`{sample}_k141_{contig}_{gene}`) are not
guaranteed unique in the workbook: two ids each turned up twice, pointing
at proteins of different lengths (so genuinely different proteins that
happen to share an id, most likely a contig number reused across two
independent per-sample assemblies) rather than a duplicated row. There is no
way to tell from the workbook alone which of the two the sequence in the
FASTA belongs to, so both instances of each are dropped rather than
guessed at. A further single row carries free-text ("Porin Omp2b ...") in
the Protein ID column instead of an id -- a data-entry slip -- and is
dropped by the FASTA join below since it matches no sequence.

```{r dedupe}
ids <- raw[["Protein ID...25"]]
dupe_ids <- unique(ids[duplicated(ids)])
raw <- raw[!(ids %in% dupe_ids), ]

c(dropped_ambiguous_ids = length(dupe_ids), remaining_rows = nrow(raw))
```

# Abundance: raw PUP counts

`ProteomeMeanPUP.csv` carries one row per protein, one column per sample,
already keyed by the same replicate labels used throughout (`A1`-`A3` for
PHBV, `B1`-`B3` for PLA, ... `S1`-`S3` for sediment). Two things need
handling before this is a count matrix `rarefy_even_depth()` can use:

- **A blank cell is not the same claim as a zero.** A protein absent from
  the search database offered to a given sample's spectral match (most
  proteins are called from one sample's own assembly, not searched against
  every other sample) is blank; a protein that *was* searched for and
  simply not seen is `0`. This report cannot tell the two apart from the
  CSV alone, and for a per-sample abundance vector the distinction does not
  matter anyway -- both mean zero counted evidence in that sample, which is
  also how the workbook's own %NSAF sheet already treated it (that sheet
  has zero `NA` cells in its sample columns). Blanks are filled with `0`
  here on the same basis.
- **Values are not all integers** (`r sum(as.matrix(read.csv(file.path(HERE, "ProteomeMeanPUP.csv"), row.names = 1, check.names = FALSE)[-1]) %% 1 != 0, na.rm = TRUE)`
  of them, e.g. `1.5`, `38.5`) -- "Mean PUP" in the filename is exactly
  that, spectral evidence already averaged (technical replicates, or
  peptides distributed across more than one matching protein group) before
  this table was exported. `rarefy_even_depth()` needs whole counts to
  subsample without replacement, so these are rounded to the nearest
  integer. That is a real, if small, distortion of the original evidence,
  introduced here rather than left for `rarefy_even_depth()` to fail on
  silently.

```{r pup}
pup_cols <- c("A1","A2","A3","B1","B2","B3","C1","C2","C3","D1","D2","D3",
             "F2","F3","F7","G1","G2","G3","I1","I2","I3","S1","S2","S3")

pup <- read.csv(file.path(HERE, "ProteomeMeanPUP.csv"), row.names = 1, check.names = FALSE)
stopifnot(all(pup_cols %in% names(pup)), !anyDuplicated(rownames(pup)))

counts <- round(as.matrix(pup[pup_cols]))
counts[is.na(counts)] <- 0
storage.mode(counts) <- "integer"
stopifnot(all(counts >= 0))

c(pup_rows = nrow(counts), samples = ncol(counts), total_counts = sum(counts))
```

# Sequences

```{r fasta}
full_seq <- read_fasta(file.path(HERE, "ProteomOnlyFasta.fasta"))

keep <- Reduce(intersect, list(raw[["Protein ID...25"]], rownames(counts), names(full_seq)))
c(annotation_rows = nrow(raw), pup_rows = nrow(counts),
 fasta_sequences = length(full_seq), matched = length(keep))

counts   <- counts[keep, , drop = FALSE]
full_seq <- full_seq[keep]
raw      <- raw[match(keep, raw[["Protein ID...25"]]), ]
counts   <- counts[, colSums(counts) > 0, drop = FALSE]  # samples never drop; keep guard for reruns
counts   <- counts[rowSums(counts) > 0, , drop = FALSE]
full_seq <- full_seq[rownames(counts)]
```

The malformed "Porin Omp2b..." row, any FASTA-only sequences (proteins
called but never detected in any sample), and any PUP-only rows (no
workbook annotation row at all) fall out here as the non-overlapping part
of the three-way join.

# Design, from the sample columns

The column names themselves encode the design (substrate + replicate);
there is no sample name string to parse.

```{r design}
md <- data.frame(
  sample_id = pup_cols,
  substrate = rep(c("PHBV", "PLA", "HDPE", "PLA/PCL", "Filter paper",
                    "PHBH", "PHBH Green", "Sediment"), each = 3),
  replicate = pup_cols,
  row.names = pup_cols)
md$substrate <- factor(md$substrate,
  levels = c("Sediment", "Filter paper", "PHBV", "PHBH", "PHBH Green", "PLA", "PLA/PCL", "HDPE"))

table(md$substrate)
```

Levels are ordered sediment (no substrate) → filter paper (cellulose, fast) →
the three PHA-family bioplastics → the two PLA-family plastics → HDPE
(petroleum-derived, slowest), a rough biodegradability gradient used only
for consistent plotting order below, not analysis.

# Annotation

Curated function calls (`MantisName`, `MicrobeName`, `PlasticDB`, `Tigerfam`,
the free-text "Potential plastic depolymerase" flag) and the two clustering
ids the sheet provides (`Bin ID`, a per-assembly MAG; `Species Group`, a
coarser cross-sample protein cluster) are carried into `tax_table` -- no
lineage is available for this dataset (no GTDB call, unlike the diet study),
so there is no rank-based taxonomy to build.

`Bin ID` is renamed to `GenomeID` in `tax_table` below, because `case_study2.Rmd`
already has a "bin" -- the `repdist_bin()` structural cluster a differential-
abundance test runs on -- and that bin has nothing to do with this one. A `GenomeID`
is a MAG (metagenome-assembled genome): a group of contigs the original assembly
pipeline believed came from the same organism. A `repdist_bin()` bin groups
proteins by predicted fold, blind to organism entirely, so a single structural bin
routinely contains members from a dozen different `GenomeID`s -- that is not a
coincidence to explain away, it is the whole reason a structural ground metric can
find a signal an organism- or accession-bound method cannot.

`MantisName` and `MicrobeName` mark "no call" with the literal text `"NA"` rather
than a blank cell (`r sum(raw[["MantisName"]] == "NA", na.rm = TRUE)` and
`r sum(raw[["MicrobeName"]] == "NA", na.rm = TRUE)` rows respectively) -- every
other annotation column in this sheet uses a real blank or `"-"` for the same
thing. Left alone, that string survives every `is.na()` check downstream as a
false non-missing value, so it is normalized to a real `NA` here, at the one
place this whole report reads these two columns from.

## One genome per protein, and who else has it

`Bin ID` sometimes lists more than one genome for a single protein row
(`r sum(lengths(strsplit(raw[["Bin ID"]], ",")) > 1, na.rm = TRUE)` of
`r sum(!is.na(raw[["Bin ID"]]))` rows) -- e.g. `8-2DC2.bin.66.fa,9-2DC3.bin.85.fa,7-2DC1.bin.12.fa`.
That is not one protein assigned to three genomes at once: this table's rows are
themselves cross-sample clusters (per-sample assemblies were searched for
near-identical protein sequences and merged into one row before this workbook was
built), and binning into a MAG happens *per assembly*, *before* that merge -- so a
cluster spanning several samples' assemblies carries one genome call from each of
them, and a list of more than one is direct evidence that near-identical copies of
this protein were independently assembled and binned in more than one sample.

Two things are kept from that list, not one. `GenomeID` is the call from the
protein's *own* originating sample (its id prefix, e.g. `7-2DC1` in
`7-2DC1_k141_774037_95`) -- checked here to actually be present in every row's list
before relying on it (it is, with zero exceptions). The rare row where that sample
appears twice (`.bin.unbinned.fa` alongside a real genome number, or two different
real genome numbers) keeps a real genome over "unbinned", and otherwise just the
first -- an edge case affecting
`r sum(vapply(strsplit(raw[["Bin ID"]], ","), function(b) sum(sub("\\.bin\\..*$", "", b) == sub("\\.bin\\..*$", "", b)[1]) > 1, logical(1)), na.rm = TRUE)`
of `r sum(!is.na(raw[["Bin ID"]]))` rows. `SharedGenomes` is everything *else* on the
list -- every genome call from a *different* sample, i.e. every other organism this
exact protein cluster was independently found and binned in. Genome calls from the
protein's own sample are excluded from that list even in the rare multi-genome-per-
sample rows above: a second bin number from the same assembly is not a second
organism.

```{r genome-id}
resolve_genomes <- function(bin_list, own_sample) {
  b  <- strsplit(bin_list, ",")
  bs <- lapply(b, function(x) sub("\\.bin\\..*$", "", x))
  out <- mapply(function(bi, bsi, os) {
    if (anyNA(bi)) return(c(genome = NA_character_, shared = NA_character_))
    own_cand <- bi[bsi == os]
    stopifnot(length(own_cand) >= 1)        # own sample must appear -- checked above
    real <- own_cand[!grepl("\\.bin\\.unbinned\\.fa$", own_cand)]
    genome <- if (length(real)) real[1] else own_cand[1]
    other <- bi[bsi != os]                  # other SAMPLES, not just other list entries
    c(genome = genome, shared = if (length(other)) paste(other, collapse = ", ") else NA_character_)
  }, b, bs, own_sample)
  list(genome = unname(out["genome", ]), shared = unname(out["shared", ]))
}
genomes <- resolve_genomes(raw[["Bin ID"]], sub("_k141_.*$", "", raw[["Protein ID...25"]]))
c(proteins_with_shared_genomes = sum(!is.na(genomes$shared)))
```

`r sum(!is.na(genomes$shared))` of `r sum(!is.na(raw[["Bin ID"]]))` proteins have a
non-empty `SharedGenomes` list under that rule -- fewer than the
`r sum(lengths(strsplit(raw[["Bin ID"]], ",")) > 1, na.rm = TRUE)` rows with more
than one list entry above, because excluding same-sample duplicates removes rows
that only ever pointed at one organism to begin with.

```{r tax}
na_marker <- function(x) ifelse(x == "NA", NA_character_, x)

tax <- as.matrix(data.frame(
  GenomeID      = genomes$genome,
  SharedGenomes = genomes$shared,
  SpeciesGroup  = raw[["Species Group"]],
  MantisName    = na_marker(raw[["MantisName"]]),
  MicrobeName   = na_marker(raw[["MicrobeName"]]),
  PlasticDB     = raw[["PlasticDB"]],
  Tigerfam      = raw[["Tigerfam"]],
  Depolymerase  = raw[["Potential plastic depolymerase"]],
  Description   = raw[["Description"]],
  row.names = rownames(counts), check.names = FALSE))
```

## No phylogenetic tree

Same situation as the diet study (`docs/...`, `analysis/data_curation.Rmd`
§ No phylogenetic tree): these proteins are not homologous to each other as
a whole catalog, so there is no alignment to build a tree from, and no
curated lineage exists here to fall back to a taxonomy cladogram the way the
diet study's Table 6 allowed. No weighted-UniFrac arm in the analysis that
follows, for the same reason.

# The phyloseq object

`otu_table` holds the rounded PUP counts, unrarefied and with every sample this
workbook has -- including the ones a later depth check may drop. Rarefaction is a
sample-level decision made from the full depth distribution (which samples are
outliers depends on seeing all of them), so it happens in `case_study2.Rmd`, not
here, the same separation `data_curation.Rmd` used for the diet study.

```{r build}
ps <- phyloseq(otu_table(counts, taxa_are_rows = TRUE),
               sample_data(md),
               tax_table(tax),
               refseq(AAStringSet(full_seq)))
ps
```

```{r save}
saveRDS(ps, file.path(HERE, "ps.rds"))
```
