---
title: "Simulation study: when protein identity fails and function does not"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE, autodep = TRUE,
                      fig.width = 8, fig.height = 3.4, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

Bray-Curtis treats two proteins as either the same accession or maximally
different. Metaproteomes break that: samples can differ in function while sharing
no accession, and Bray-Curtis cannot tell that apart from any other difference.

Here two groups of samples differ **only** in protein function. Depth, richness
and abundance profile are identical by construction, and no accession appears
in more than one sample. Three distances are compared: Bray-Curtis, weighted
UniFrac on the inferred protein phylogeny, and weighted UniFrac on a tree fitted
to a TM-Vec structural ground metric (`repdist_bin()$tree`).

Both UniFrac arms are the same estimator in the same (unnormalized) form, so they
differ in exactly one thing -- the tree -- which is what makes this a ground-metric
ablation rather than a contest between estimators. This dataset is also the one
place the phylogenetic arm is honest: 480 proteins from four families genuinely
align, where the case studies' ~19,000-protein communities do not.

```{r libs, cache = FALSE}
library(phyloseq)
library(Biostrings)
library(vegan)
library(ggplot2)
library(gt)
library(here)
devtools::load_all(here(), quiet = TRUE)
set.seed(1)
```

## Data

Two functional groups, each built from **two** Pfam families rather than one,
from UniProt via `analysis/fetch_pfam.R`:

| Pfam | family | role | median aa | class | reaction |
|---|---|---|---|---|---|
| PF00162 | PGK | A | 395 | transferase, EC 2.7.2.3 | ATP-dependent phosphoryl transfer |
| PF01546 | Peptidase_M20 | B | 379 | hydrolase, EC 3.4.-.- | zinc-dependent amide hydrolysis |
| PF00406 | Adenylate_kinase | A | 214 | transferase, EC 2.7.4.3 | ATP-dependent phosphoryl transfer |
| PF00293 | Nudix_hydrolase | B | 175 | hydrolase, EC 3.6.1.- | nucleotide pyrophosphohydrolysis |

A single family per group would make family identity and group the same
variable, so any family-specific quirk -- length included -- is perfectly
collinear with group and no covariate can separate them. Two families per
group breaks that: role A spans both a ~395 aa and a ~214 aa family, role B a
~379 aa and a ~175 aa one, so the only thing consistent within a role is the
functional class.

For each role, candidates from its two families (one protein per organism per
family, reviewed bacterial UniProt entries) are pooled and 600 are drawn at
random (`set.seed(1)`, once, at the top of `fetch_pfam.R`) -- no matching of
organisms between roles. Length is a consequence of that draw, not a
constraint on it, so it is checked below rather than engineered:

```{r universe}
universe <- readRDS(here("data", "simulated", "pfam_universe.rds"))
tapply(nchar(universe$seq), universe$role, summary)
```

## Simulation

30 samples per group. Group A expresses PGK + Adenylate_kinase, group B
Peptidase_M20 + Nudix_hydrolase.

- **Proteins.** Each role's 600 candidates are shuffled once and the first 240
  (30 samples x 8 proteins) are dealt out in disjoint slices of 8, so every
  sample gets a random mix of both families and no accession lands in two
  samples. The remaining 360 candidates per role stay in `universe` unused --
  slack for the length check above, not part of the simulation.
- **Abundance.** One weight profile is reused by every sample, then counts are
  multinomial at a fixed depth.

```{r simulate}
M <- 8L; DEPTH <- 2000L; N <- 30L

# two columns on purpose: plot_ordination() drops a one-column sample_data and
# then cannot find the variable to colour by
meta <- data.frame(sample = sprintf("s%02d", seq_len(2 * N)),
                   group = rep(c("A", "B"), each = N),
                   row.names = sprintf("s%02d", seq_len(2 * N)))
pool <- lapply(split(universe$protein, universe$role), sample)
prof <- sort(rlnorm(M, 0, 0.6), decreasing = TRUE)

counts <- t(sapply(seq_len(2 * N), function(s) {
  slice <- pool[[meta$group[s]]][((s - 1) %% N) * M + seq_len(M)]
  p <- setNames(numeric(nrow(universe)), universe$protein)
  p[slice] <- prof
  rmultinom(1, DEPTH, p)[, 1]
}))
dimnames(counts) <- list(rownames(meta), universe$protein)

stopifnot(all(rowSums(counts) == DEPTH),             # depth matched
          length(unique(rowSums(counts > 0))) == 1L, # richness matched
          max(colSums(counts > 0)) == 1L)            # no accession in two samples
```

A tree over the proteins actually used (not the full 1200-candidate universe),
so weighted UniFrac has a ground metric to use:

```{r tree}
used <- universe[match(colnames(counts), universe$protein), ]
seqs <- AAStringSet(setNames(used$seq, used$protein))
aln  <- DECIPHER::AlignSeqs(seqs, verbose = FALSE)
dm   <- DECIPHER::DistanceMatrix(aln, verbose = FALSE, includeTerminalGaps = FALSE)
dm[is.na(dm)] <- max(dm, na.rm = TRUE)   # unalignable pairs => maximally distant
tree <- ape::as.phylo(as.hclust(DECIPHER::Treeline(myDistMatrix = dm, method = "NJ",
                                                   verbose = FALSE)))
```

Counts, tree, sequences and metadata go into one `phyloseq` object, which is the
input to everything below.

```{r phyloseq}
ps <- phyloseq(otu_table(counts, taxa_are_rows = FALSE), sample_data(meta),
               phy_tree(tree), refseq(seqs))
saveRDS(ps, here("data", "simulated", "sim_ps.rds"))
ps
```

## Embedding

TM-Vec1: its head was trained to regress the TM-score of protein pairs, so cosine
similarity between two vectors is a predicted TM-score.

```{r embed}
Z <- embed_proteins(as.character(refseq(ps)))[taxa_names(ps), ]
```

## Results

```{r distances}
cnt <- as(otu_table(ps), "matrix")

# The ground metric is built once and reused by everything downstream. The
# structure tree is the complete-linkage hierarchy used for the structural bins.
G    <- seq_repdist(Z)                                     # 1 - predicted TM-score
tree <- repdist_bin(cnt, G)$tree
tree_fidelity <- cor(as.vector(G),
                     as.vector(as.dist(ape::cophenetic.phylo(tree)[labels(G), labels(G)])))

# repdist fits the tree; phyloseq computes the UniFrac. Raw (normalized = FALSE)
# is the form that equals W1 under the tree's path metric. Unlike the case
# studies, this dataset has a real phylogeny -- 480 proteins from four families
# genuinely align -- so the phylogenetic arm here is an inferred tree, not the
# alignment-free stand-in those reports have to use.
ps_str <- phyloseq(otu_table(ps), sample_data(ps), phy_tree(tree))

# Both UniFrac arms are normalized = FALSE. phyloseq::distance(ps, "wunifrac")
# would give the *normalized* form, which divides by a scale factor and is no
# longer a transport cost -- pairing it with an unnormalized structure arm would
# make the two arms differ in two things instead of one.
D <- list(
  `Bray-Curtis`              = phyloseq::distance(ps, method = "bray"),
  `w-UniFrac (phylogenetic)` = UniFrac(ps,     weighted = TRUE, normalized = FALSE),
  `w-UniFrac (structure)`    = UniFrac(ps_str, weighted = TRUE, normalized = FALSE)
)
```

The two roles are not length-matched by construction (see the length check
above), so `group` is tested with each sample's abundance-weighted mean
protein length entered first -- the same covariate-adjustment pattern
`case_study.Rmd` already uses for dose/sex/age, applied here to rule out
length as the source of the separation rather than genuine functional signal.
Reporting the raw (unadjusted) $R^2$ alongside it shows how much work the
adjustment is actually doing: with a single family per group that gap was
large, because family and group were the same variable; with two length
scales per group it should be small.

```{r permanova}
md <- data.frame(sample_data(ps))
len <- setNames(nchar(as.character(refseq(ps))), taxa_names(ps))
md$mean_log_len <- as.numeric((cnt %*% log(len[colnames(cnt)])) / rowSums(cnt))

gt(do.call(rbind, lapply(names(D), function(n) {
  raw <- adonis2(D[[n]] ~ group, data = md, by = "terms", permutations = 9999)
  adj <- adonis2(D[[n]] ~ mean_log_len + group, data = md, by = "terms", permutations = 9999)
  data.frame(distance = n,
             R2_raw = round(raw["group", "R2"], 3),
             R2_adjusted = round(adj["group", "R2"], 3),
             F = round(adj["group", "F"], 2), p = adj["group", "Pr(>F)"])
})))
```

```{r pcoa, warning = FALSE}
plot_ordination(ps, ordinate(ps, "PCoA", distance = D[["Bray-Curtis"]]), color = "group") + theme_bw()
plot_ordination(ps, ordinate(ps, "PCoA", distance = D[["w-UniFrac (phylogenetic)"]]), color = "group") + theme_bw()
plot_ordination(ps, ordinate(ps, "PCoA", distance = D[["w-UniFrac (structure)"]]), color = "group") + theme_bw()
```

## Conclusion

Bray-Curtis returns 1.000 for every pair. A constant distance matrix cannot
express any grouping, so PERMANOVA lands on the exact null $1/(n-1) = 0.017$ with
$p = 1$. It is not weak here, it is blind: no accession is shared, and accessions
are all it reads.

Weighted UniFrac on the TM-Vec1 structure tree recovers the difference from sequence
alone, without an alignment or annotation. `R2_raw` and `R2_adjusted`
are close, which is the point of building two length scales into each group:
with a single family per group, family and group were the same variable and
length adjustment was doing real work to separate them; here it barely moves
the number, because no single length characterizes either group any more (per
the length check above, both roles span everything from the ~175-220 aa family
to the ~380-400 aa one) — the only thing PGK and Adenylate_kinase share, or
Peptidase_M20 and Nudix_hydrolase share, is the functional class, so that is
what a real effect has to generalize over rather than any one family's length
or sequence idiosyncrasies.

Weighted UniFrac on a protein tree also finds it, at a lower $R^2$. That arm is
not a fair fight and should not be quoted as one: none of the four families are
homologous to each other, so the alignment behind the tree is largely gap and
between-family branch lengths are gap-penalty artifacts rather than measured
divergence. A better tree would likely close some of the gap. The claim worth
making is applicability — an embedding stays defined where an alignment across
non-homologous proteins does not — rather than power.

Two limits. This is the zero-overlap extreme; real metaproteomes share some
accessions, and Bray-Curtis regains ability as that overlap rises. And the
functional contrast here is total, every protein in one group a transferase and
in the other a hydrolase, so the effect size says nothing about realistic ones.
